---
title: "Usage of beezdemand"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage of beezdemand}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA)
```

## Rationale Behind beezdemand

Behavioral economic demand is gaining in popularity. The motivation behind beezdemand was to 
create an alternative tool to conduct these
analyses. This package is not necessarily meant to be a replacement
for other softwares; rather, it is meant to serve as an
additional tool in the behavioral economist's toolbox. It is meant for
researchers to conduct behavioral economic (be) demand the easy (ez) way.

[`R`](https://www.r-project.org/) is an open-source statistical programming language. It is powerful and allows for nearly endless
customizability.

## Note About Use

This package is a work in progress. I welcome suggestions, feedback,
questions, and comments regarding what other researchers might be
interested in seeing. If you encounter bugs, errors, or other
discrepancies please either open an issue or
[contact me](mailto:bkaplan4@ku.edu?Subject=Question/comment%20Regarding%20beezdemand%20Package "Email bkaplan4@ku.edu") and I will do my best to fix
the problem.

## Installation

Right now the package can be obtained from my GitHub page:
[github.com/brentkaplan](https://github.com/brentkaplan). There are plans to make it available
properly on CRAN. In any case, to install the package first install
Hadley Wickham's `devtools` package:

```install.packages("devtools")```

Then simply install the package using the following command:

```devtools::install_github("brentkaplan/beezdemand", build_vignettes
= TRUE)```

This will compile this vignette during the installation.

```{r include = FALSE}
library(beezdemand)
```

## Using the Package

### Example Dataset

I include an example dataset to demonstrate how data should be
entered and how to use the functions. This example dataset consists
of participant responses on an alcohol purchase task. Participants (id)
reported the number of alcoholic drinks (y) they would be willing to
purchase and consume at various prices (x; USD). Note the long format:

```{r example-data-set, echo=FALSE, results='asis'}
knitr::kable(apt[c(1:8, 17:24), ])
```

### Obtain Descriptive Data

Descriptive values of responses at each price. Includes mean, standard
deviation, proportion of zeros, and numer of NAs. If `bwplot = TRUE`, a box-and-whisker
plot (in PDF format) is also created and saved into the relative path: `../plots/bwplot-MMDD/`

```{r descriptive, eval=FALSE}
GetDescriptives(apt, bwplot = TRUE)
```

```{r descriptive-output, echo=FALSE, results='asis'}
descr <- GetDescriptives(apt)
knitr::kable(descr)
```

<center><img src="bwplot.pdf" alt="Box-and-whisker plot"  width="504"
height="432"/></center>

### Identify Unsystematic Responses

Examine consistency of demand data using Stein et al.'s (2015)
alogrithm for identifying unsystematic responses. Default values
shown, but they can be customized.

```{r unsystematic, eval=FALSE}
CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.1, reversals = 0, ncons0 = 2)
```

```{r unsystematic-output, echo=FALSE, results='asis'}
knitr::kable(head(CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.1, reversals = 0, ncons0 = 2), 5))
```

### Analyze Demand Data

Results of the analysis return both empirical and derived measures for
use in additional analyses and model specification. Equations include
the linear model, exponential model, and exponentiated model. Soon, I
will be including the nonlinear mixed effects model, mixed effects
versions of the exponential and exponentiated model, and the
experimental "Q0+1" model.

```{r hs, eval=FALSE}
FitCurves(apt, "hs")
```

```{r hs-output, echo=FALSE, results='asis', eval = FALSE}
knitr::kable(head(FitCurves(apt, "hs"), 5)[ , 1:6], caption = "Empirical Measures")
knitr::kable(head(FitCurves(apt, "hs"), 5)[ , 7:11], caption = "Fitted Measures")
knitr::kable(head(FitCurves(apt, "hs"), 5)[ , 12:20], caption = "Uncertainty and Model Information")
knitr::kable(head(FitCurves(apt, "hs"), 5)[ , 21:24], caption = "Derived Measures")
```


```{r koff, eval=FALSE}
FitCurves(apt, "koff")
```

```{r koff-output, echo=FALSE, results='asis', eval = FALSE}
knitr::kable(head(FitCurves(apt, "koff"), 5)[ , 1:6], caption = "Empirical Measures")
knitr::kable(head(FitCurves(apt, "koff"), 5)[ , 7:11], caption = "Fitted Measures")
knitr::kable(head(FitCurves(apt, "koff"), 5)[ , 12:20], caption = "Uncertainty and Model Information")
knitr::kable(head(FitCurves(apt, "koff"), 5)[ , 21:24], caption = "Derived Measures")
```

Graphs can automatically be created at both an aggregate
(fitting on the means or on the pooled data
[NOTE: This disregards clustering within individual]) and individual
level.


### Share k Globally; Fit Other Parameters Locally

Provides the ability to share k globally (across all participants)
while estimating Q0 and alpha locally. While this works, it may take
some time with larger sample sizes.

```{r share, eval=FALSE}
FitCurves(apt, "hs", k = "share")
```

```{r, include=FALSE, eval = FALSE}
df <- FitCurves(apt, "hs", k = "share")
```

```{r share-output, echo=FALSE, results='asis', eval = FALSE}
knitr::kable(head(df, 5)[ , 1:6], caption = "Empirical Measures")
knitr::kable(head(df, 5)[ , 7:11], caption = "Fitted Measures")
knitr::kable(head(df, 5)[ , 12:20], caption = "Uncertainty and Model Information")
knitr::kable(head(df, 5)[ , 21:24], caption = "Derived Measures")
```

### Learn More About Functions

To learn more about a function and what arguments it takes, type "?" in front of the function name.

```{r learn, eval=FALSE}
?CheckUnsystematic
```

```{r learn-output, eval=FALSE}
CheckUnsystematic          package:beezdemand          R Documentation

Systematic Purchase Task Data Checker

Description:

     Applies Stein, Koffarnus, Snider, Quisenberry, & Bickels (2015)
     criteria for identification of nonsystematic purchase task data.

Usage:

     CheckUnsystematic(dat, deltaq = 0.025, bounce = 0.1, reversals = 0,
       ncons0 = 2)

Arguments:

     dat: Dataframe in long form. Colums are id, x, y.

  deltaq: Numeric vector of length equal to one. The criterion by which
          the relative change in quantity purchased will be compared.
          Relative changes in quantity purchased below this criterion
          will be flagged. Default value is 0.025.

  bounce: Numeric vector of length equal to one. The criterion by which
          the number of price-to-price increases in consumption that
          exceed 25% of initial consumption at the lowest price,
          expressed relative to the total number of price increments,
          will be compared. The relative number of price-to-price
          increases above this criterion will be flagged. Default value
          is 0.10.

reversals: Numeric vector of length equal to one. The criterion by
          which the number of reversals from number of consecutive (see
          ncons0) 0s will be compared. Number of reversals above this
          criterion will be flagged. Default value is 0.

  ncons0: Numer of consecutive 0s prior to a positive value is used to
          flag for a reversal. Value can be either 1 (relatively more
          conservative) or 2 (default; as recommended by Stein et al.,
          (2015).

Details:

     This function applies the 3 criteria proposed by Stein et al.,
     (2015) for identification of nonsystematic purchase task data. The
     three criteria include trend (deltaq), bounce, and reversals from
     0. Also reports number of positive consumption values.

Value:

     Dataframe

Author(s):

     Brent Kaplan <bkaplan4@ku.edu>

Examples:

     ## Using all default values
     CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.10, reversals = 0, ncons0 = 2)
     ## Specifying just 1 zero to flag as reversal
     CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.10, reversals = 0, ncons0 = 1)
```

## Acknowledgments

- Derek D. Reed, Applied Behavioral Economics Laboratory
[www.behavioraleconlab.com](www.behavioraleconlab.com)

- Paul E. Johnson, Center for Research Methods and Data Analysis, University of Kansas
[www.crmda.ku.edu](www.crmda.ku.edu)

- Michael Amlung, Cognitive Neuroscience of Addictions Laboratory
[www.cnalab.weebly.com](www.cnalab.weebly.com)

- Peter G. Roma, Institutes for Behavior Resources, Inc.
[www.ibrinc.org](www.ibrinc.org)

- Steven R. Hursh, Institutes for Behavior Resources, Inc.
[www.ibrinc.org](www.ibrinc.org)

- Shawn P. Gilroy, [GitHub](https://github.com/miyamot0)

## Recommended Readings

- Reed, D. D., Niileksela, C. R., & Kaplan, B. A. (2013). Behavioral economics: A tutorial for behavior analysts in practice. *Behavior Analysis in Practice, 6* (1), 34–54.

- Reed, D. D., Kaplan, B. A., & Becirevic, A. (2015). Basic research on the behavioral economics of reinforcer value. In *Autism Service Delivery* (pp. 279-306). Springer New York.

- Hursh, S. R., & Silberberg, A. (2008). Economic demand and essential value. *Psychological Review, 115* (1), 186-198.

- Koffarnus, M. N., Franck, C. T., Stein, J. S., & Bickel, W. K. (2015). A modified exponential behavioral economic demand model to better describe consumption data. *Experimental and Clinical Psychopharmacology, 23* (6), 504-512.

- Stein, J. S., Koffarnus, M. N., Snider, S. E., Quisenberry, A. J., & Bickel, W. K. (2015). Identification and management of nonsystematic purchase task data: Toward best practice. *Experimental and Clinical Psychopharmacology*

