---
title: "How to Use Cross-Price Demand Model Functions"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Use Cross-Price Demand Model Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    include = TRUE,
    collapse = TRUE,
    comment = NA,
    dev = "png",
    dpi = 144,
    fig.width = 6,
    fig.height = 4
)

library(beezdemand)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)

# Load package datasets
data("etm", package = "beezdemand")
data("lowNicClean", package = "beezdemand")
data("cannabisCigarettes", package = "beezdemand")
data("ongoingETM", package = "beezdemand")

lnic <- lowNicClean |>
    mutate(
        target = if_else(commodity == "cigarettesAdj", "adjusting", "fixed"),
        group = "cigarettes"
    ) |>
    select(
        -commodity
    ) |>
    relocate(condition, .after = group)

can_cig <- cannabisCigarettes |>
    select(
        id,
        x,
        y,
        target = commodity
    )

ongoing_etm <- ongoingETM |>
    pivot_longer(
        -c(id, x, flavor),
        names_to = "target",
        values_to = "y",
        values_drop_na = TRUE
    ) |>
    select(id, x, y, target)
```

## Data Structure

```{r}
glimpse(etm)
```

Typical columns:
- `id`: participant identifier

- `x`: alternative product price

- `y`: consumption level

- `target`: condition type (e.g., "alt")

- `group`: product category

## Complete Cross-Price Analysis Workflow

This section demonstrates a complete analysis workflow using data that contains multiple
experimental conditions: target consumption when the alternative is absent (`alone`),
target consumption when the alternative is present (`own`), and alternative consumption
as a function of target price (`alt`).

### Loading the cp Dataset

```{r load-cp}
# Load the cross-price example dataset
data("cp", package = "beezdemand")

# Examine structure
glimpse(cp)

# View conditions
table(cp$target)
```

The `cp` dataset contains:

- **alone**: Target commodity (cigarettes) consumption when alternative is *not* available
- **own**: Target commodity consumption when alternative *is* available
- **alt**: Alternative commodity (e-cigarettes) consumption as a function of target price

Note that `x` represents the **target commodity price** throughout all conditions.

### Step 1: Fit Target Demand (Alone Condition)

First, we fit a standard demand curve to the target commodity when the alternative is absent:

```{r fit-alone}
# Filter to alone condition
alone_data <- cp |>
    dplyr::filter(target == "alone")

# Fit demand curve using FitCurves
fit_alone <- FitCurves(
    dat = alone_data,
    equation = "koff",
    k = 2,
    detailed = TRUE
)

# View results
fit_alone
```

### Step 2: Fit Target Demand (Own Condition)

Next, fit the same demand model to target consumption when the alternative is present:

```{r fit-own}
# Filter to own condition
own_data <- cp |>
    dplyr::filter(target == "own")

# Fit demand curve
fit_own <- FitCurves(
    dat = own_data,
    equation = "koff",
    k = 2,
    detailed = TRUE
)

# View results
fit_own
```

### Step 3: Fit Cross-Price Model (Alt Condition)

Finally, fit the cross-price model to alternative consumption as a function of target price:

```{r fit-alt-cp}
# Filter to alt condition
alt_data <- cp |>
    dplyr::filter(target == "alt")

# Fit cross-price model
fit_alt <- fit_cp_nls(
    dat = alt_data,
    equation = "exponentiated",
    return_all = TRUE
)

# View results
summary(fit_alt)
```

### Comparing Results Across Conditions

```{r compare-conditions}
# Extract key parameters from dfres
comparison <- data.frame(
    Condition = c("Alone (Target)", "Own (Target)", "Alt (Cross-Price)"),
    Q0_or_Qalone = c(
        fit_alone$dfres$Q0d,
        fit_own$dfres$Q0d,
        coef(fit_alt)["qalone"]
    ),
    Alpha_or_I = c(
        fit_alone$dfres$Alpha,
        fit_own$dfres$Alpha,
        coef(fit_alt)["I"]
    )
)

comparison
```

**Interpretation:**

- Comparing `Q0` between alone and own conditions shows how the presence of an alternative
  affects baseline consumption of the target commodity
- The `I` parameter from the cross-price model indicates whether the products are
  substitutes (I < 0) or complements (I > 0)

### Combined Visualization

```{r combined-plot, fig.width=8, fig.height=4}
# Create prediction data
x_seq <- seq(0.01, max(cp$x), length.out = 100)

# Extract parameters from dfres
Q0_alone <- fit_alone$dfres$Q0d
K_alone <- fit_alone$dfres$K
Alpha_alone <- fit_alone$dfres$Alpha

Q0_own <- fit_own$dfres$Q0d
K_own <- fit_own$dfres$K
Alpha_own <- fit_own$dfres$Alpha

# Get predictions for each condition using the exponential demand equation
pred_alone <- Q0_alone * 10^(K_alone * (exp(-Alpha_alone * Q0_alone * x_seq) - 1))
pred_own <- Q0_own * 10^(K_own * (exp(-Alpha_own * Q0_own * x_seq) - 1))
pred_alt <- predict(fit_alt, newdata = data.frame(x = x_seq))$y_pred

# Combine into plot data
plot_data <- data.frame(
    x = rep(x_seq, 3),
    y = c(pred_alone, pred_own, pred_alt),
    Condition = rep(c("Target (Alone)", "Target (Own)", "Alternative (Cross-Price)"), each = length(x_seq))
)

# Plot
ggplot() +
    geom_point(data = cp, aes(x = x, y = y, color = target), alpha = 0.6) +
    geom_line(data = plot_data, aes(x = x, y = y, color = Condition), linewidth = 1) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
        x = "Target Price",
        y = "Consumption",
        title = "Cross-Price Analysis: All Conditions",
        color = "Condition"
    ) +
    theme_bw()
```

## Checking Unsystematic Data

```{r}
etm |>
    dplyr::filter(group %in% "E-Cigarettes" & id %in% 1)


unsys_one <- etm |>
    filter(group %in% "E-Cigarettes" & id %in% 1) |>
    check_unsystematic_cp()

unsys_one

unsys_one_lnic <- lnic |>
    filter(
        target == "adjusting",
        id == "R_00Q12ahGPKuESBT"
    ) |>
    check_unsystematic_cp(detailed = TRUE)

unsys_one_lnic
```

```{r}
unsys_all <- etm |>
    group_by(id, group) |>
    nest() |>
    mutate(unsys = map(data, check_unsystematic_cp)) |>
    unnest(unsys)

unsys_all

summary(unsys_all)
```

### Demonstration from Rzeszutek et al. (under review)

#### Low Nicotine Study (Kaplan et al., 2018)

```{r}
unsys_all_lnic <- lnic |>
    filter(target == "fixed") |>
    group_by(id, condition) |>
    nest() |>
    mutate(
        unsys = map(
            data,
            check_unsystematic_cp,
            detailed = TRUE
        )
    ) |>
    unnest(unsys) |>
    arrange(id)

unsys_all_lnic

cat("--- 100% ---\n")
unsys_all_lnic |>
    dplyr::filter(condition == "100%") |>
    summary()
cat("\n\n")

cat("--- 2% ---\n")
unsys_all_lnic |>
    dplyr::filter(condition == "2%") |>
    summary()
cat("\n\n")
cat("--- 2% NegFrame ---\n")
unsys_all_lnic |>
    dplyr::filter(condition == "2% NegFrame") |>
    summary()

summary(unsys_all_lnic)
```


#### Unpublished Cannabis and Cigarette Data

```{r}
unsys_all_can_cig <- can_cig |>
    filter(target %in% c("cannabisFix", "cigarettesFix")) |>
    group_by(id, target) |>
    nest() |>
    mutate(
        unsys = map(
            data,
            check_unsystematic_cp,
            detailed = TRUE
        )
    ) |>
    unnest(unsys) |>
    arrange(id)

cat("--- Cigarettes ---\n")
unsys_all_can_cig |>
    dplyr::filter(target == "cigarettesFix") |>
    summary()
cat("\n\n")


cat("--- Cannabis ---\n")
unsys_all_can_cig |>
    dplyr::filter(target == "cannabisFix") |>
    summary()
cat("\n\n")
```

#### In Progress Experimental Tobacco Marketplace Data

```{r}
unsys_all_ongoing_etm <- ongoing_etm |>
    # one person is doubled up
    distinct() |>
    filter(target %in% c("FixCig", "ECig")) |>
    group_by(id, target) |>
    nest() |>
    mutate(
        unsys = map(
            data,
            check_unsystematic_cp,
            detailed = TRUE
        )
    ) |>
    unnest(unsys)


cat("--- Fixed Cigarettes ---\n")
unsys_all_ongoing_etm |>
    dplyr::filter(target == "FixCig") |>
    summary()
cat("\n\n")

cat("--- E-Cigarettes ---\n")
unsys_all_ongoing_etm |>
    dplyr::filter(target == "ECig") |>
    summary()
cat("\n\n")
```

## Nonlinear Model Fitting

### Two Stage

```{r}
fit_one <- etm |>
    dplyr::filter(group %in% "E-Cigarettes" & id %in% 1) |>
    fit_cp_nls(
        equation = "exponentiated",
        return_all = TRUE
    )

summary(fit_one)

plot(fit_one, x_trans = "log10")
```

```{r}
fit_all <- etm |>
    group_by(id, group) |>
    nest() |>
    mutate(
        unsys = map(data, check_unsystematic_cp),
        fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
        summary = map(fit, summary),
        plot = map(fit, plot, x_trans = "log10"),
        glance = map(fit, glance),
        tidy = map(fit, tidy)
    )

fit_all

fit_all$summary[[2]]

rand_id <- sample(unique(etm$id), 1)
print(rand_id)
fit_all$summary[[rand_id]]
fit_all$plot[[rand_id]]
```

### Fit to Group (pooled by group)

```{r}
fit_pooled <- etm |>
    group_by(group) |>
    nest() |>
    mutate(
        unsys = map(data, check_unsystematic_cp),
        fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
        summary = map(fit, summary),
        plot = map(fit, plot, x_trans = "log10"),
        glance = map(fit, glance),
        tidy = map(fit, tidy)
    )

fit_pooled

fit_pooled |>
    dplyr::filter(group %in% "E-Cigarettes") |>
    pull(summary)

fit_pooled |>
    dplyr::filter(group %in% "E-Cigarettes") |>
    pull(plot)
```

### Fit to Group (mean)

```{r}
fit_mean <- etm |>
    group_by(group, x) |>
    summarise(
        y = mean(y)
    ) |>
    ungroup() |>
    group_by(group) |>
    nest() |>
    mutate(
        unsys = map(data, check_unsystematic_cp),
        fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
        summary = map(fit, summary),
        plot = map(fit, plot, x_trans = "log10"),
        glance = map(fit, glance),
        tidy = map(fit, tidy)
    )

fit_mean

fit_mean |>
    unnest(cols = c(glance, tidy)) |>
    select(
        group,
        term,
        estimate
    ) |>
    ggplot(aes(x = group, y = estimate, group = term)) +
    geom_bar(stat = "identity") +
    geom_point() +
    facet_wrap(~term, ncol = 1, scales = "free_y")

fit_mean |>
    dplyr::filter(group %in% "E-Cigarettes") |>
    pull(summary)

fit_mean |>
    dplyr::filter(group %in% "E-Cigarettes") |>
    pull(plot)
```


## Linear Model Fitting

```{r}
fit_one_linear <- etm |>
    dplyr::filter(group %in% "E-Cigarettes" & id %in% 1) |>
    fit_cp_linear(
        type = "fixed",
        log10x = TRUE,
        return_all = TRUE
    )

summary(fit_one_linear)
plot(fit_one_linear, x_trans = "log10")
```

## Linear Mixed-Effects Model

```{r}
fit_mixed <- fit_cp_linear(
    etm,
    type = "mixed",
    log10x = TRUE,
    group_effects = "interaction",
    random_slope = FALSE,
    return_all = TRUE
)

summary(fit_mixed)

# plot fixed effects only
plot(fit_mixed, x_trans = "log10", pred_type = "fixed")

# plot random effects only
plot(fit_mixed, x_trans = "log10", pred_type = "random")

# plot both fixed and random effects
plot(fit_mixed, x_trans = "log10", pred_type = "all")
```

## Extracting Model Coefficients

```{r}
glance(fit_one)
tidy(fit_one)
```

```{r}
extract_coefficients(fit_mixed)
```

## Post-hoc Estimated Marginal Means and Comparisons

```{r}
cp_posthoc_slopes(fit_mixed)
cp_posthoc_intercepts(fit_mixed)
```

## Two-Part Hurdle Cross-Price Model

The hurdle cross-price model extends the standard cross-price demand framework to handle
datasets with many zero consumption values. It uses a two-part structure:

**Part I (Zero Model):** Logistic regression for probability of zero consumption
$$\text{logit}(P(\text{zero})) = \beta_0 + \beta_1 \ln(x + \epsilon) + a_i$$

**Part II (Intensity Model):** Exponential cross-price model for positive consumption
$$\ln(Q) = \ln(Q_{alone}) + I \cdot \exp(-\beta \cdot x) + b_i + e$$

Where:
- $Q_{alone}$ is baseline consumption when alternative is unavailable
- $I$ is the interaction parameter (negative = substitutes, positive = complements)
- $\beta$ is the rate of decay
- $a_i, b_i$ are subject-specific random effects

### Fitting a 2-RE Hurdle Model

```{r hurdle-2re, warning=FALSE}
# Filter to E-Cigarettes alternative condition
ecig_data <- etm |>
    dplyr::filter(group == "E-Cigarettes", target == "alt") |>
    as.data.frame()

# Check zero proportion
cat("Zero consumption:", round(100 * mean(ecig_data$y == 0), 1), "%\n")

# Fit 2-RE hurdle model (random effects on zeros and Qalone)
fit_hurdle_2re <- fit_cp_hurdle(
    data = ecig_data,
    y_var = "y",
    x_var = "x",
    id_var = "id",
    random_effects = c("zeros", "qalone"),
    verbose = 0
)

summary(fit_hurdle_2re)
```

### Fitting a 3-RE Hurdle Model

The 3-RE model adds a random effect on the interaction parameter $I$:

```{r hurdle-3re, warning=FALSE}
# Fit 3-RE hurdle model
fit_hurdle_3re <- fit_cp_hurdle(
    data = ecig_data,
    y_var = "y",
    x_var = "x",
    id_var = "id",
    random_effects = c("zeros", "qalone", "I"),
    verbose = 0
)

summary(fit_hurdle_3re)
```

### Comparing Models

Use a likelihood ratio test to compare nested models:

```{r hurdle-compare}
compare_cp_hurdle_models(fit_hurdle_3re, fit_hurdle_2re)
```

### Subject-Level Parameters

```{r hurdle-params}
# Get parameter summary
get_cp_hurdle_param_summary(fit_hurdle_2re)

# View individual subject parameters
head(predict(fit_hurdle_2re, type = "parameters"))
```

### Predictions

```{r hurdle-predict}
# Population-level predictions
new_prices <- data.frame(x = c(2, 8, 16, 32, 64))
cbind(
    new_prices,
    demand = predict(fit_hurdle_2re, new_prices, type = "response", level = "population")$.fitted,
    prob_zero = predict(fit_hurdle_2re, new_prices, type = "probability", level = "population")$.fitted
)
```

### Plotting

```{r hurdle-plot, fig.width=7, fig.height=5}
# Population demand curve
plot(fit_hurdle_2re, type = "demand")

# Probability of zero consumption
plot(fit_hurdle_2re, type = "probability")

# Individual curves
plot(fit_hurdle_2re, type = "individual")
```

## Joint Hurdle Cross-Price Model

The joint hurdle model extends the separate demand and cross-price models by estimating
all parameters within a **single likelihood**. This approach:

- Shares information across conditions via correlated random effects
- Provides a unified framework for analyzing complete cross-price experiments
- Enables direct comparison of target demand with and without an alternative

### Model Structure

The joint model estimates three "streams" simultaneously:

1. **alone.target**: Target commodity demand when alternative is absent
2. **own.target**: Target commodity demand when alternative is present
3. **own.alt**: Alternative commodity consumption as a function of target price

**Key design features:**

- **Shared hurdle**: A single subject-level random effect ($a_i$) captures propensity
  to purchase anything, with stream-specific intercepts
- **Global k**: The scaling parameter $k$ is shared across target demand streams
- **price_T is the predictor for all streams**: Alternative price is assumed fixed
  and is not modeled as a predictor

### Preparing Data for Joint Model

The `prepare_joint_data()` function transforms data into the required format:

```{r joint-prep}
# Load the cross-price example dataset
data("cp", package = "beezdemand")

# Examine the data structure
cat("Data dimensions:", nrow(cp), "rows x", ncol(cp), "cols\n")
cat("Subjects:", length(unique(cp$id)), "\n")
cat("Conditions:", paste(unique(cp$target), collapse = ", "), "\n")

# View sample data
head(cp)

# Prepare for joint model
prep <- prepare_joint_data(
    data = cp,
    y_var = "y",
    price_var = "x",
    id_var = "id",
    target_var = "target",
    alone_value = "alone",
    own_value = "own",
    alt_value = "alt"
)

# Check stream counts
prep$stream_counts
```

### Comparing Separate Fits vs Joint Fit

#### Step 1: Separate Fits (Traditional Workflow)

First, let's fit each condition separately using existing functions:

```{r joint-separate-fits}
# Filter data by condition
alone_data <- cp |> dplyr::filter(target == "alone")
own_data <- cp |> dplyr::filter(target == "own")
alt_data <- cp |> dplyr::filter(target == "alt")

# Fit target demand (alone condition)
fit_alone <- FitCurves(
    dat = alone_data,
    equation = "koff",
    k = 2,
    detailed = TRUE
)

cat("=== Alone Condition (Target Demand) ===\n")
cat("Q0:", round(fit_alone$dfres$Q0d, 2), "\n")
cat("Alpha:", round(fit_alone$dfres$Alpha, 6), "\n")

# Fit target demand (own condition)
fit_own <- FitCurves(
    dat = own_data,
    equation = "koff",
    k = 2,
    detailed = TRUE
)

cat("\n=== Own Condition (Target Demand) ===\n")
cat("Q0:", round(fit_own$dfres$Q0d, 2), "\n")
cat("Alpha:", round(fit_own$dfres$Alpha, 6), "\n")

# Fit cross-price model (alt condition)
fit_alt <- fit_cp_nls(
    dat = alt_data,
    equation = "exponentiated",
    return_all = TRUE
)

cat("\n=== Alt Condition (Cross-Price) ===\n")
cat("Qalone:", round(coef(fit_alt)["qalone"], 2), "\n")
cat("I:", round(coef(fit_alt)["I"], 4), "\n")
cat("beta:", round(coef(fit_alt)["beta"], 4), "\n")
```

#### Step 2: Joint Model Fit

Now fit all streams simultaneously with the joint hurdle model:

```{r joint-fit, warning=FALSE}
# Fit the joint hurdle model
fit_joint <- fit_joint_hurdle(
    data = cp,
    y_var = "y",
    price_var = "x",
    id_var = "id",
    target_var = "target",
    alone_value = "alone",
    own_value = "own",
    alt_value = "alt",
    k = 2, # Fixed k (matches existing workflow)
    epsilon = 0.001,
    verbose = 1
)
```

### Examining Joint Model Results

```{r joint-summary}
# Full summary
summary(fit_joint)
```

### Extracting Coefficients

```{r joint-coef}
# All coefficients
coef(fit_joint, type = "all")

# Hurdle parameters only
coef(fit_joint, type = "hurdle")

# Demand parameters (target streams)
coef(fit_joint, type = "demand")

# Cross-price parameters
coef(fit_joint, type = "crossprice")
```

### Comparing Parameter Estimates

```{r joint-compare-params}
# Create comparison table
comparison <- data.frame(
    Stream = c("alone.target", "own.target", "own.alt"),
    Separate_Q0 = c(
        fit_alone$dfres$Q0d,
        fit_own$dfres$Q0d,
        coef(fit_alt)["qalone"]
    ),
    Joint_Q0 = c(
        exp(coef(fit_joint)["logQ0_AT"]),
        exp(coef(fit_joint)["logQ0_OT"]),
        exp(coef(fit_joint)["logQalone_OA"])
    ),
    Separate_Sens = c(
        fit_alone$dfres$Alpha,
        fit_own$dfres$Alpha,
        coef(fit_alt)["I"]
    ),
    Joint_Sens = c(
        coef(fit_joint)["alpha_AT"],
        coef(fit_joint)["alpha_OT"],
        coef(fit_joint)["I"]
    )
)

comparison
```

### Predictions by Stream

```{r joint-predict}
# Population-level predictions for all streams
prices <- data.frame(price_T = c(0.5, 1, 2, 4, 8, 16, 32))

# Response predictions (expected consumption)
pred_response <- predict(fit_joint, newdata = prices, type = "response", stream = "all")
head(pred_response, 12)

# Probability of zero consumption
pred_prob <- predict(fit_joint, newdata = prices, type = "probability", stream = "all")
head(pred_prob, 12)
```

### Subject-Level Parameters

```{r joint-subject-pars}
# Extract subject-specific parameters
subject_pars <- predict(fit_joint, type = "parameters")
subject_pars
```

### Visualizing Joint Model Results

```{r joint-plot-demand, fig.width=8, fig.height=5}
# Demand curves for all streams
plot(fit_joint, type = "demand")
```

```{r joint-plot-prob, fig.width=8, fig.height=4}
# Zero probability curves
plot(fit_joint, type = "probability")
```

### Using Tidy Methods

```{r joint-tidy}
# Tidy coefficient table
tidy(fit_joint)

# Model summary statistics
glance(fit_joint)
```

### Key Takeaways

**When to use the joint model:**

- You have data across multiple conditions (alone, own, alt)
- You want correlated random effects across conditions
- You want a single unified analysis rather than separate fits

**When separate fits may be preferred:**

- You only have one condition
- You want maximum flexibility in model specification per condition
- Computational resources are limited

**Important notes:**

- `price_T` (target price) is the predictor for **all** streams
- Alternative price is assumed **fixed** and does not enter the model
- The joint model produces consistent parameter orderings with the separate fits

---

## Variant 2: Latent-Trait Joint Model

### Why a Latent-Trait Model?

The saturated joint model (Variant 1) uses **stream-specific random effects** — each
subject has separate deviations for alone-target, own-target, and own-alt streams.
This is flexible but can be unstable with sparse data.

The **latent-trait model** (Variant 2) takes a different approach:

- Instead of stream-specific random effects, subjects have **4 latent behavioral traits**
- These traits **load into** the model parameters via fixed mappings
- Result: fewer parameters, more interpretable structure, better stability

| Latent Trait | Behavioral Interpretation |
|--------------|---------------------------|
| `u_buy`      | Propensity to purchase (drives hurdle) |
| `u_val`      | Baseline consumption intensity |
| `u_sens`     | Price sensitivity (elasticity) |
| `u_sub`      | Substitution propensity |

**Key insight:** Variant 2 trades flexibility for structure, not correctness. Both
models use the exact same mean functions — only the random effect structure differs.

### Fitting the Latent-Trait Model

```{r fit-latent-model}
# Fit the latent-trait joint model
fit_latent <- fit_joint_hurdle(
    data = cp,
    joint_type = "latent",
    k = 2,
    verbose = 0
)

print(fit_latent)
```

### Comparing Variant 1 and Variant 2

Let's compare the two variants side-by-side:
```{r compare-variants}
# Hurdle parameters (should be similar)
cat("=== Hurdle Parameters ===\n")
cat("\nSaturated (Variant 1):\n")
print(coef(fit_joint, type = "hurdle"))
cat("\nLatent (Variant 2):\n")
print(coef(fit_latent, type = "hurdle"))

# Cross-price parameters
cat("\n=== Cross-Price Parameters ===\n")
cat("\nSaturated (Variant 1):\n")
print(coef(fit_joint, type = "crossprice"))
cat("\nLatent (Variant 2):\n")
print(coef(fit_latent, type = "crossprice"))
```

```{r compare-fit-stats}
# Model fit comparison
comparison <- data.frame(
    Variant = c("Saturated", "Latent"),
    LogLik = c(fit_joint$logLik, fit_latent$logLik),
    AIC = c(fit_joint$AIC, fit_latent$AIC),
    BIC = c(fit_joint$BIC, fit_latent$BIC),
    Converged = c(fit_joint$convergence == 0, fit_latent$convergence == 0)
)
comparison
```

### Latent Trait Loadings

The latent model includes **loading parameters** that control how the substitution
trait (`u_sub`) differentially affects each stream:

```{r latent-loadings}
# Extract latent loadings
coef(fit_latent, type = "latent")
```

- `lambda_sub_q0`: How substitution affects own-target baseline demand
- `lambda_sub_alpha`: How substitution affects own-target price sensitivity
- `lambda_sub_alt`: How substitution affects alternative consumption

### Predictions with the Latent Model

Predictions work identically to the saturated model:

```{r latent-predictions}
# Predicted demand across streams
prices <- data.frame(price_T = c(0.5, 1, 2, 4, 8, 16))

pred_response <- predict(fit_latent, newdata = prices, type = "response", stream = "all")
pred_response

# Zero probability predictions
pred_prob <- predict(fit_latent, newdata = prices, type = "probability", stream = "all")
pred_prob
```

### Interpreting Latent Traits

The latent traits provide a **behavioral decomposition** of individual differences:

| Trait | High Value Means | Low Value Means |
|-------|------------------|-----------------|
| `u_buy` | More likely to purchase at any price | More likely to abstain |
| `u_val` | Higher baseline consumption | Lower baseline consumption |
| `u_sens` | More price-sensitive (steeper curves) | Less price-sensitive |
| `u_sub` | Stronger substitution behavior | Weaker substitution |

**Important:** These are latent constructs, not directly observed. Their values
are estimated from the data pattern and should be interpreted with appropriate
caution.

### When to Use Each Variant

| Criterion | Saturated (Variant 1) | Latent (Variant 2) |
|-----------|----------------------|-------------------|
| **Data richness** | Many obs per subject | Sparse data |
| **Primary interest** | Stream-specific effects | Behavioral traits |
| **Stability** | May require more data | More stable |
| **Interpretability** | Random effects per stream | Explicit trait meanings |
| **Flexibility** | Maximum flexibility | Constrained structure |

**Recommendation:**

- Start with **Variant 1 (saturated)** if you have rich data and want maximum flexibility
- Use **Variant 2 (latent)** if:
 - Data is sparse
 - Model convergence is problematic
 - You want interpretable behavioral dimensions
 - Stability is more important than stream-specific flexibility

Both variants produce valid inference — the choice depends on your research goals
and data characteristics.
