---
title: "How to Use Cross-Price Demand Model Functions"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Use Cross-Price Demand Model Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  include = TRUE,
  collapse = TRUE,
  comment = NA,
  fig.width = 6,
  fig.height = 4
)

# library(beezdemand)
devtools::load_all()
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(here)

conflicted::conflict_prefer("filter", "dplyr")

here::i_am("vignettes/cross-price-models.Rmd")

load(here::here("data", "etm.RData"))

lnic <- readr::read_csv(here::here("data", "lowNicClean.csv")) |>
  mutate(
    target = if_else(commodity == "cigarettesAdj", "adjusting", "fixed"),
    group = "cigarettes"
  ) |>
  select(
    -commodity
  ) |>
  relocate(condition, .after = group)

can_cig <- readr::read_csv(here::here("data", "cannabisCigarettes.csv")) |>
  select(
    id,
    x,
    y,
    target = commodity
  )

ongoing_etm <- readr::read_csv(here::here("data", "ongoingETM.csv")) |>
  pivot_longer(
    -c(id, x, flavor),
    names_to = "target",
    values_to = "y",
    values_drop_na = TRUE
  ) |>
  select(id, x, y, target)

# load("../data/etm.RData")
```

## Data Structure

```{r}
glimpse(etm)
```

Typical columns:
- `id`: participant identifier

- `x`: alternative product price

- `y`: consumption level

- `target`: condition type (e.g., "alt")

- `group`: product category

## Checking Unsystematic Data

```{r}
etm |>
  dplyr::filter(group %in% "E-Cigarettes" & id %in% 1)


unsys_one <- etm |>
  filter(group %in% "E-Cigarettes" & id %in% 1) |>
  check_unsystematic_cp()

unsys_one

unsys_one_lnic <- lnic |>
  filter(
    target == "adjusting",
    id == "R_00Q12ahGPKuESBT"
  ) |>
  check_unsystematic_cp(detailed = TRUE)

unsys_one_lnic

```

```{r}
unsys_all <- etm |>
  group_by(id, group) |>
  nest() |>
  mutate(unsys = map(data, check_unsystematic_cp)) |>
  unnest(unsys)

unsys_all

summary(unsys_all)
```

### Demonstration from Rzeszutek et al. (under review)

#### Low Nicotine Study (Kaplan et al., 2018)

```{r}
unsys_all_lnic <- lnic |>
  filter(target == "fixed") |>
  group_by(id, condition) |>
  nest() |>
  mutate(
    unsys = map(
      data,
      check_unsystematic_cp,
      detailed = TRUE
    )
  ) |>
  unnest(unsys) |>
  arrange(id)

unsys_all_lnic

cat("--- 100% ---\n")
unsys_all_lnic |>
  dplyr::filter(condition == "100%") |>
  summary()
cat("\n\n")

cat("--- 2% ---\n")
unsys_all_lnic |>
  dplyr::filter(condition == "2%") |>
  summary()
cat("\n\n")
cat("--- 2% NegFrame ---\n")
unsys_all_lnic |>
  dplyr::filter(condition == "2% NegFrame") |>
  summary()

summary(unsys_all_lnic)
```


#### Unpublished Cannabis and Cigarette Data

```{r}

unsys_all_can_cig <- can_cig |>
  filter(target %in% c("cannabisFix", "cigarettesFix")) |>
  group_by(id, target) |>
  nest() |>
  mutate(
    unsys = map(
      data,
      check_unsystematic_cp,
      detailed = TRUE
    )
  ) |>
  unnest(unsys) |>
  arrange(id)

cat("--- Cigarettes ---\n")
unsys_all_can_cig |>
  dplyr::filter(target == "cigarettesFix") |>
  summary()
cat("\n\n")


cat("--- Cannabis ---\n")
unsys_all_can_cig |>
  dplyr::filter(target == "cannabisFix") |>
  summary()
cat("\n\n")


```

#### In Progress Experimental Tobacco Marketplace Data

```{r}
unsys_all_ongoing_etm <- ongoing_etm |>
  # one person is doubled up
  distinct() |>
  filter(target %in% c("FixCig", "ECig")) |>
  group_by(id, target) |>
  nest() |>
  mutate(
    unsys = map(
      data,
      check_unsystematic_cp,
      detailed = TRUE
    )
  ) |>
  unnest(unsys)


cat("--- Fixed Cigarettes ---\n")
unsys_all_ongoing_etm |>
  dplyr::filter(target == "FixCig") |>
  summary()
cat("\n\n")

cat("--- E-Cigarettes ---\n")
unsys_all_ongoing_etm |>
  dplyr::filter(target == "ECig") |>
  summary()
cat("\n\n")


```

## Nonlinear Model Fitting

### Two Stage

```{r}
fit_one <- etm |>
  dplyr::filter(group %in% "E-Cigarettes" & id %in% 1) |>
  fit_cp_nls(
    equation = "exponentiated",
    return_all = TRUE
  )

summary(fit_one)

plot(fit_one, x_trans = "log10")
```

```{r}
fit_all <- etm |>
  group_by(id, group) |>
  nest() |>
  mutate(
    unsys = map(data, check_unsystematic_cp),
    fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
    summary = map(fit, summary),
    plot = map(fit, plot, x_trans = "log10"),
    glance = map(fit, glance),
    tidy = map(fit, tidy)
  )

fit_all

fit_all$summary[[2]]

rand_id <- sample(unique(etm$id), 1)
print(rand_id)
fit_all$summary[[rand_id]]
fit_all$plot[[rand_id]]
```

### Fit to Group (pooled by group)

```{r}
fit_pooled <- etm |>
  group_by(group) |>
  nest() |>
  mutate(
    unsys = map(data, check_unsystematic_cp),
    fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
    summary = map(fit, summary),
    plot = map(fit, plot, x_trans = "log10"),
    glance = map(fit, glance),
    tidy = map(fit, tidy)
  )

fit_pooled

fit_pooled |>
  dplyr::filter(group %in% "E-Cigarettes") |>
  pull(summary)

fit_pooled |>
  dplyr::filter(group %in% "E-Cigarettes") |>
  pull(plot)


```

### Fit to Group (mean)

```{r}
fit_mean <- etm |>
  group_by(group, x) |>
  summarise(
    y = mean(y)
  ) |>
  ungroup() |>
  group_by(group) |>
  nest() |>
  mutate(
    unsys = map(data, check_unsystematic_cp),
    fit = map(data, fit_cp_nls, equation = "exponentiated", return_all = TRUE),
    summary = map(fit, summary),
    plot = map(fit, plot, x_trans = "log10"),
    glance = map(fit, glance),
    tidy = map(fit, tidy)
  )

fit_mean

fit_mean |>
  unnest(cols = c(glance, tidy)) |>
  select(
    group,
    term,
    estimate
  ) |>
  ggplot(aes(x = group, y = estimate, group = term)) +
  geom_bar(stat = "identity") +
  geom_point() +
  facet_wrap(~term, ncol = 1, scales = "free_y")

fit_mean |>
  dplyr::filter(group %in% "E-Cigarettes") |>
  pull(summary)

fit_mean |>
  dplyr::filter(group %in% "E-Cigarettes") |>
  pull(plot)


```


## Linear Model Fitting

```{r}
fit_one_linear <- etm |>
  dplyr::filter(group %in% "E-Cigarettes" & id %in% 1) |>
  fit_cp_linear(
    type = "fixed",
    log10x = TRUE,
    return_all = TRUE
  )

summary(fit_one_linear)
plot(fit_one_linear, x_trans = "log10")
```

## Linear Mixed-Effects Model

```{r}
fit_mixed <- fit_cp_linear(
  etm,
  type = "mixed",
  log10x = TRUE,
  group_effects = "interaction",
  random_slope = FALSE,
  return_all = TRUE
)

summary(fit_mixed)

# plot fixed effects only
plot(fit_mixed, x_trans = "log10", pred_type = "fixed")

# plot random effects only
plot(fit_mixed, x_trans = "log10", pred_type = "random")

# plot both fixed and random effects
plot(fit_mixed, x_trans = "log10", pred_type = "all")
```

## Extracting Model Coefficients

```{r}
glance(fit_one)
tidy(fit_one)
```

```{r}
extract_coefficients(fit_mixed)
```

## Post-hoc Estimated Marginal Means and Comparisons

```{r}
cp_posthoc_slopes(fit_mixed)
cp_posthoc_intercepts(fit_mixed)
```
