---
title: "Fixed-Effect Demand Modeling with `beezdemand`"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixed-Effect Demand Modeling with `beezdemand`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "fixed-demand_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)

data("apt", package = "beezdemand", envir = environment())

cache_key_object <- function(x) {
  tmp <- tempfile(fileext = ".rds")
  saveRDS(x, tmp)
  on.exit(unlink(tmp), add = TRUE)
  unname(tools::md5sum(tmp))
}

knitr::opts_chunk$set(
  cache.extra = list(
    beezdemand_version = as.character(utils::packageVersion("beezdemand")),
    apt = cache_key_object(apt)
  )
)
```

# Introduction

This vignette demonstrates how to fit individual (fixed-effect) demand curves
using `fit_demand_fixed()`. This function fits separate nonlinear least squares
(NLS) models for each subject, producing per-subject estimates of demand
parameters like $Q_{0}$ (intensity) and $\alpha$ (elasticity).

**When to use fixed-effect models:** Fixed-effect models are appropriate when
you want independent curve fits for each participant. They make no assumptions
about the distribution of parameters across subjects and are the simplest
approach to demand curve analysis. For hierarchical models that share
information across subjects, see `vignette("mixed-demand")`. For guidance on
choosing between approaches, see `vignette("model-selection")`.

**Data format:** All modeling functions expect long-format data with columns
`id` (subject identifier), `x` (price), and `y` (consumption). See
`vignette("beezdemand")` for details on data preparation and conversion from
wide format.

# Fitting with Different Equations

`fit_demand_fixed()` supports several equation forms. We demonstrate the three
most common below using the `apt` (Alcohol Purchase Task) dataset.

## Hursh & Silberberg ("hs")
The exponential model of demand (Hursh & Silberberg, 2008):

$$\log_{10}(Q) = \log_{10}(Q_0) + k \cdot (e^{-\alpha \cdot Q_0 \cdot x} - 1)$$

```{r fit_hs}
fit_hs <- fit_demand_fixed(apt, equation = "hs", k = 2)
fit_hs
```

## Koffarnus ("koff")

The exponentiated model (Koffarnus et al., 2015):

$$Q = Q_0 \cdot 10^{k \cdot (e^{-\alpha \cdot Q_0 \cdot x} - 1)}$$

```{r fit_koff}
fit_koff <- fit_demand_fixed(apt, equation = "koff", k = 2)
fit_koff
```

## Simplified ("simplified")

The simplified exponential model (Rzeszutek et al., 2025) does not require a
scaling constant $k$:

$$Q = Q_0 \cdot e^{-\alpha \cdot Q_0 \cdot x}$$

```{r fit_simplified}
fit_simplified <- fit_demand_fixed(apt, equation = "simplified")
fit_simplified
```

# The k Parameter

The scaling constant $k$ controls the range of the demand function. For the
`"hs"` and `"koff"` equations, `k` can be specified in several ways:

| `k` value | Behavior |
|-----------|----------|
| Numeric (e.g., `2`) | Fixed constant for all subjects (default) |
| `"ind"` | Individual $k$ per subject, computed from each subject's data range |
| `"share"` | Single shared $k$ estimated across all subjects via global regression |
| `"fit"` | $k$ is a free parameter estimated jointly with $Q_0$ and $\alpha$ |

```{r fit_k_options, eval = FALSE}
## Fixed k (default)
fit_demand_fixed(apt, equation = "hs", k = 2)

## Individual k per subject
fit_demand_fixed(apt, equation = "hs", k = "ind")

## Shared k across subjects
fit_demand_fixed(apt, equation = "hs", k = "share")

## Fitted k as free parameter
fit_demand_fixed(apt, equation = "hs", k = "fit")
```

The `param_space` argument controls whether optimization is performed on the
natural scale (`"natural"`, the default) or log10 scale (`"log10"`). The log10
scale can improve convergence for some datasets:

```{r fit_log10, eval = FALSE}
fit_demand_fixed(apt, equation = "hs", k = 2, param_space = "log10")
```

# Inspecting Fits

All `beezdemand_fixed` objects support the standard `tidy()`, `glance()`,
`augment()`, and `confint()` methods for programmatic access to results.

## tidy(): Per-Subject Parameter Estimates

```{r tidy}
tidy(fit_hs)
```

## glance(): Model-Level Summary

```{r glance}
glance(fit_hs)
```

## augment(): Fitted Values and Residuals

```{r augment}
augment(fit_hs)
```

## confint(): Confidence Intervals

```{r confint}
confint(fit_hs)
```

## summary(): Formatted Summary

The `summary()` method provides a formatted overview including parameter
distributions across subjects:

```{r summary}
summary(fit_hs)
```

# Normalized Alpha ($\alpha^*$)

When $k$ varies across participants or studies (e.g., `k = "ind"` or
`k = "fit"`), raw $\alpha$ values are not directly comparable. The `alpha_star`
column in `tidy()` output provides a normalized version (Strategy B; Rzeszutek
et al., 2025) that adjusts for the scaling constant:

$$\alpha^* = \frac{-\alpha}{\ln\!\left(1 - \frac{1}{k \cdot \ln(b)}\right)}$$

where $b$ is the logarithmic base (10 for HS/Koff equations). Standard errors
are computed via the delta method. `alpha_star` requires $k \cdot \ln(b) > 1$;
otherwise `NA` is returned.

```{r alpha_star}
tidy(fit_hs) |>
  filter(term == "alpha_star") |>
  select(id, term, estimate, std.error)
```

# Plotting

## Basic Demand Curves

The `plot()` method displays fitted demand curves with observed data points.
The x-axis defaults to a log10 scale:

```{r plot_basic, cache = FALSE}
plot(fit_hs)
```

## Faceted by Subject

Use `facet = TRUE` to show each subject in a separate panel:

```{r plot_facet, cache = FALSE, fig.height = 8}
plot(fit_hs, facet = TRUE)
```

## Axis Transformations

Control the x- and y-axis transformations with `x_trans` and `y_trans`:

```{r plot_transforms, cache = FALSE}
plot(fit_hs, x_trans = "pseudo_log", y_trans = "pseudo_log")
```

# Diagnostics

## Model Checks

`check_demand_model()` summarizes convergence, residual properties, and
potential issues:

```{r diagnostics}
check_demand_model(fit_hs)
```

## Residual Plots

`plot_residuals()` produces diagnostic plots. Use `$fitted` for a
residuals-vs-fitted plot:

```{r residuals, cache = FALSE}
plot_residuals(fit_hs)$fitted
```

# Predictions

## Default Predictions

`predict()` with no arguments returns fitted values at the observed prices:

```{r predict_default}
predict(fit_hs)
```

## Custom Price Grid

Supply `newdata` to predict at specific prices:

```{r predict_custom}
new_prices <- data.frame(x = c(0, 0.5, 1, 2, 5, 10, 20))
predict(fit_hs, newdata = new_prices)
```

# Aggregated Models

Instead of fitting each subject individually, you can fit a single curve to
aggregated data.

## Mean Aggregation

`agg = "Mean"` computes the mean consumption at each price across subjects,
then fits a single curve to those means:

```{r agg_mean}
fit_mean <- fit_demand_fixed(apt, equation = "hs", k = 2, agg = "Mean")
fit_mean
```

```{r agg_mean_plot, cache = FALSE}
plot(fit_mean)
```

## Pooled Aggregation

`agg = "Pooled"` fits a single curve to all data points pooled together,
retaining error around each observation but ignoring within-subject clustering:

```{r agg_pooled}
fit_pooled <- fit_demand_fixed(apt, equation = "hs", k = 2, agg = "Pooled")
fit_pooled
```

```{r agg_pooled_plot, cache = FALSE}
plot(fit_pooled)
```

# Grouped Analysis with `by`

When your dataset includes grouping variables (e.g., gender, condition, drug),
you can run fits separately per group using the `by` parameter. This replaces
manual `split()` / `lapply()` / `bind_rows()` workflows.

We use the `apt_full` dataset, which includes demographic columns like `gender`:

```{r grouped-setup}
data("apt_full", package = "beezdemand", envir = environment())
```

## Fitting by Group

```{r grouped-fit}
fit_by_gender <- fit_demand_fixed(
  apt_full, equation = "hs", k = 2, by = "gender"
)
fit_by_gender
```

The result is a `beezdemand_fixed_grouped` object containing one child
`beezdemand_fixed` fit per group. All standard accessors prepend the grouping
column(s) automatically:

```{r grouped-tidy}
tidy(fit_by_gender) |> head(10)
```

```{r grouped-glance}
glance(fit_by_gender)
```

## Grouped Plotting

`plot()` combines per-group panels:

```{r grouped-plot, cache = FALSE, fig.height = 8}
plot(fit_by_gender)
```

## Grouped Systematicity and Descriptive Checks

The `by` parameter works the same way in `check_systematic_demand()` and
`get_descriptive_summary()`:

```{r grouped-systematic}
sys <- check_systematic_demand(apt_full, by = "gender")
sys
```

```{r grouped-descriptive, cache = FALSE, fig.height = 6}
desc <- get_descriptive_summary(apt_full, by = "gender")
plot(desc)
```

# Conclusion

The `fit_demand_fixed()` interface provides a modern, consistent API for
individual demand curve fitting with full support for the `tidy()` / `glance()`
/ `augment()` workflow. For datasets where borrowing strength across subjects
is desirable, consider the mixed-effects or hurdle model approaches.

# See Also

- `vignette("beezdemand")` -- Getting started with beezdemand
- `vignette("model-selection")` -- Choosing the right model class
- `vignette("group-comparisons")` -- Extra sum-of-squares F-test for group comparisons
- `vignette("mixed-demand")` -- Mixed-effects nonlinear demand models (NLME)
- `vignette("mixed-demand-advanced")` -- Advanced mixed-effects topics
- `vignette("hurdle-demand-models")` -- Two-part hurdle demand models via TMB
- `vignette("cross-price-models")` -- Cross-price demand analysis
- `vignette("migration-guide")` -- Migrating from `FitCurves()` to `fit_demand_fixed()`
