---
title: "Migration Guide: FitCurves to fit_demand_fixed"
author: "beezdemand Developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migration Guide: FitCurves to fit_demand_fixed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  dev = "png",
  dpi = 144
)
library(beezdemand)
library(dplyr)
```

# Overview

Starting with beezdemand version 0.2.0, `FitCurves()` has been superseded by
`fit_demand_fixed()`. This guide helps you migrate existing code to use the
modern API.
## Why Migrate?

The new `fit_demand_fixed()` function provides:

- **Structured S3 objects** with consistent methods (`summary()`, `tidy()`,
  `glance()`, `predict()`, `plot()`)
- **Tidyverse integration** via tibble outputs from `tidy()` and `glance()`
- **Standardized API** consistent with other beezdemand model classes
- **Better reproducibility** with stored call and parameter information

`FitCurves()` will continue to work but is no longer actively developed.
New features will only be added to `fit_demand_fixed()`.

# Quick Migration Reference

| FitCurves() | fit_demand_fixed() | Notes |
|-------------|---------------------|-------|
| `dat` | `data` | Renamed for consistency |
| `xcol` | `x_var` | Renamed for consistency |
| `ycol` | `y_var` | Renamed for consistency |
| `idcol` | `id_var` | Renamed for consistency |
| `detailed = TRUE` | Always detailed | fit_demand_fixed always returns full results |
| `groupcol` | Not supported | Use factor models instead |
| Returns data.frame | Returns S3 object | Use `tidy()` for data frame output |

# Basic Migration

## Before (FitCurves)

```{r fitcurves-example, eval = TRUE}
# Old approach
results <- FitCurves(
  dat = apt,
  equation = "hs",
  k = 2,
  xcol = "x",
  ycol = "y",
  idcol = "id"
)

# results is a data.frame with all parameters
head(results)
```

## After (fit_demand_fixed)

```{r fit-demand-fixed-example}
# New approach
fit <- fit_demand_fixed(
  data = apt,
  equation = "hs",
  k = 2,
  x_var = "x",
  y_var = "y",
  id_var = "id"
)

# fit is a structured S3 object
print(fit)
```

# Extracting Results

## Getting Parameter Estimates

### Before (FitCurves)

```{r extract-old, eval = TRUE}
# FitCurves returns a data frame directly
results <- FitCurves(apt, "hs", k = 2)
q0_values <- results$Q0d
alpha_values <- results$Alpha
```

### After (fit_demand_fixed)

```{r extract-new}
# Use tidy() for a tibble of coefficients
fit <- fit_demand_fixed(apt, equation = "hs", k = 2)
coefs <- tidy(fit)
head(coefs)

# Or access the results data frame directly
head(fit$results)
```

## Getting Model-Level Statistics

### Before (FitCurves)

```{r stats-old, eval = TRUE}
# Manual calculation required
results <- FitCurves(apt, "hs", k = 2)
n_converged <- sum(!is.na(results$Alpha))
mean_r2 <- mean(results$R2, na.rm = TRUE)
```

### After (fit_demand_fixed)

```{r stats-new}
# Use glance() for model-level statistics
fit <- fit_demand_fixed(apt, equation = "hs", k = 2)
glance(fit)

# Or use summary() for comprehensive output
summary(fit)
```

# Working with Predictions

## Before (FitCurves)

```{r pred-old, eval = TRUE}
# Required detailed = TRUE and manual extraction
results <- FitCurves(apt, "hs", k = 2, detailed = TRUE)
# Predictions stored in list element
predictions <- results$newdats  # List of data frames
```

## After (fit_demand_fixed)

```{r pred-new}
# Use predict() method
fit <- fit_demand_fixed(apt, equation = "hs", k = 2)

# Predict at new prices
new_prices <- data.frame(x = c(0, 0.5, 1, 2, 5, 10))
preds <- predict(fit, newdata = new_prices)
head(preds)
```

# Plotting

## Before (FitCurves)

```{r plot-old, eval = TRUE}
# Used separate PlotCurves() function
results <- FitCurves(apt, "hs", k = 2, detailed = TRUE)
PlotCurves(results)
```

## After (fit_demand_fixed)

```{r plot-new, fig.width = 6, fig.height = 4}
# Use plot() method directly on the fit object
fit <- fit_demand_fixed(apt, equation = "hs", k = 2)

# Plot first 5 subjects
plot(fit, ids = unique(apt$id)[1:5], facet = TRUE)
```

# Advanced Features

## Aggregated Data (Mean/Pooled)

Both functions support aggregation, with identical syntax:

```{r agg-example}
# Fit to mean data
fit_mean <- fit_demand_fixed(apt, equation = "hs", k = 2, agg = "Mean")
tidy(fit_mean)
```

## Parameter Space

Both functions support log10 parameterization:

```{r param-space}
# Fit in log10 space
fit_log <- fit_demand_fixed(
  apt,
  equation = "hs",
  k = 2,
  param_space = "log10"
)

# tidy() can report in either scale
tidy(fit_log, report_space = "log10") |> head()
tidy(fit_log, report_space = "natural") |> head()
```

# Feature Comparison

| Feature | FitCurves() | fit_demand_fixed() |
|---------|-------------|---------------------|
| Equations | hs, koff, linear | hs, koff, linear |
| k options | numeric, "ind", "fit", "range", "share" | numeric, "ind", "fit", "range" |
| Aggregation | "Mean", "Pooled" | "Mean", "Pooled" |
| Parameter space | natural, log10 | natural, log10 |
| `summary()` method | No | Yes |
| `tidy()` method | No | Yes |
| `glance()` method | No | Yes |
| `predict()` method | No | Yes |
| `plot()` method | No | Yes |
| `coef()` method | No | Yes |

# Suppressing Deprecation Warnings

If you need to continue using `FitCurves()` without warnings (e.g., in legacy
code or during a gradual migration), you can suppress the deprecation message:

```{r suppress-warning, eval = TRUE}
# Suppress deprecation warning temporarily
rlang::with_options(
  lifecycle_verbosity = "quiet",
  FitCurves(apt, "hs", k = 2)
)
```

However, we strongly recommend migrating to `fit_demand_fixed()` for new code.

# Need Help?

If you encounter issues during migration:

1. Check the function documentation: `?fit_demand_fixed`
2. Review the [beezdemand GitHub repository](https://github.com/brentkaplan/beezdemand)
3. Open an issue if you find bugs or have feature requests

# Session Information

```{r session-info}
sessionInfo()
```
