---
title: "Mixed-Effects Demand Modeling with `beezdemand`"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Effects Demand Modeling with `beezdemand`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, # Keep code chunks separate
  comment = "#>",
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE, # Suppress warnings for cleaner output in vignette
  message = FALSE, # Suppress messages for cleaner output in vignette
  cache = TRUE,
  cache.path = "mixed-demand_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)

# Set to FALSE when you explicitly want to run the slower / less stable examples.
render_fast <- TRUE

cache_key_object <- function(x) {
  tmp <- tempfile(fileext = ".rds")
  saveRDS(x, tmp)
  on.exit(unlink(tmp), add = TRUE)
  unname(tools::md5sum(tmp))
}

# Invalidate cache when the package data changes.
knitr::opts_chunk$set(
  cache.extra = list(
    beezdemand_version = as.character(utils::packageVersion("beezdemand")),
    apt = cache_key_object(apt),
    ko = cache_key_object(ko)
  )
)

# Optional fits used later (kept NULL in fast renders).
fit_two_factors_add <- NULL
fit_two_factors_int <- NULL
fit_simplified_example <- NULL
fit_collapsed_same <- NULL
fit_collapsed_asymmetric <- NULL
fit_drug_slope_random <- NULL
fit_age_add <- NULL
fit_age_rhs <- NULL
```

# Introduction

This vignette demonstrates how to perform mixed-effects nonlinear modeling
of behavioral economic demand data using the beezdemand package. We will focus
on the fit_demand_mixed() function and its associated helper functions for
extracting results, making predictions, and performing comparisons. These models
allow for individual differences (random effects) and the examination of how
various factors (fixed effects) influence demand parameters like $Q_{0}$ (maximum
consumption at zero price) and $\alpha$ (sensitivity of demand to price). The
parameters $Q_{0}$ and $\alpha$ are estimated on a log10 scale for numerical stability,
but reporting functions will provide them on their natural, interpretable scale.

## Data Preparation

For these examples, we will use the apt and ko datasets, which is
assumed to be available and pre-processed.

The apt dataset should contain:

- id: A unique identifier for each subject.

- x: The price of the drug.

- y: The consumption of the drug.

The ko dataset should contain:

- monkey: A subject or group identifier for random effects.

- x: The price of the commodity (in this case the fixed-ratio requirement).

- y: Raw consumption values. This is typically used with the simplified exponentiated equation.

- y_ll4: Consumption, ll4 transformed. This is typically used with the zben equation.

- Factor columns like drug and dose.


```{r}
quick_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100,
  niterEM = 20,
  maxIter = 100, # Low iterations for speed
  pnlsTol = 0.1,
  tolerance = 1e-4, # Looser tolerance
  opt = "nlminb",
  msVerbose = FALSE
)
```

## Fitting Demand Models with fit_demand_mixed()

The core function for fitting nonlinear mixed-effects demand models is `fit_demand_mixed()`.

### APT Fit and Plot

#### LL4 transformation with ZBEn

```{r apt_zben_fit}
apt_ll4 <- apt |>
  mutate(y_ll4 = ll4(y))

fit_apt_zben <- fit_demand_mixed(
  data = apt_ll4,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "id",
  equation_form = "zben",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_zben)
```

```{r apt_zben_plot, cache = FALSE}
plot(
  fit_apt_zben,
  inv_fun = ll4_inv,
  y_trans = "pseudo_log",
  x_trans = "pseudo_log",
  show_pred_lines = c("population", "individual")
) +
  facet_wrap(
    ~id
  )
```

#### Simplified Exponential

```{r apt_simplified_fit}
fit_apt_simplified <- fit_demand_mixed(
  data = apt_ll4,
  y_var = "y",
  x_var = "x",
  id_var = "id",
  equation_form = "simplified",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_simplified)
```

```{r apt_simplified_plot, cache = FALSE}
plot(
  fit_apt_simplified,
  x_trans = "pseudo_log",
  show_pred_lines = c("population", "individual")
) +
  facet_wrap(
    ~id
  )
```

#### Koffarnus (Exponentiated) Equation Form

`fit_demand_mixed()` also supports the Koffarnus et al. (2015) equation via
`equation_form = "koff"`. By default, the scaling constant `k` will be computed
from the data range (you can also specify it directly).

```{r apt_koff_fit}
fit_apt_koff <- fit_demand_mixed(
  data = apt,
  y_var = "y",
  x_var = "x",
  id_var = "id",
  equation_form = "koff",
  k = NULL,
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_koff)
```

## Inspecting Fits (tidy / glance / augment)

All modern model classes support `tidy()`, `glance()`, and `augment()` to
standardize programmatic access to estimates, model summaries, and residuals.

```{r inspect_fit}
glance(fit_apt_zben)
tidy(fit_apt_zben) |> head()
augment(fit_apt_zben) |> head()
```

## Diagnostics

Use `check_demand_model()` and the residual plotting helpers as standard
post-fit checks.

```{r diagnostics}
check_demand_model(fit_apt_zben)
plot_residuals(fit_apt_zben)$fitted
```

### Basic Model (No Factors)

This model estimates global $Q_{0}$ and $\alpha$ parameters with random effects for subjects.

```{r}
# Make sure a similar 'fit_no_factors' was created successfully in your environment
# For the vignette, let's create one that is more likely to converge quickly
# by using only Alfentanil data, which is less complex than the full dataset.
ko_alf <- ko[ko$drug == "Alfentanil", ]

fit_no_factors_vignette <- fit_demand_mixed(
  data = ko_alf,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "monkey",
  equation_form = "zben",
  nlme_control = quick_nlme_control, # Use quicker control for vignette
  start_value_method = "heuristic" # Heuristic is faster for simple model
)
print(fit_no_factors_vignette)
```

The output shows the model call, selected equation form, and if the model converged, it prints the nlme model summary.

### Model with One Factor

Let's model $Q_{0}$ and $\alpha$ as varying by dose for Alfentanil.

```{r}
fit_one_factor_dose <- fit_demand_mixed(
  data = ko_alf,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "monkey",
  factors = "dose",
  equation_form = "zben",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_one_factor_dose)
```

### Model with Two Factors (Additive)

Here, $Q_{0}$ and $\alpha$ vary by drug and dose additively.
For this, we need more data than just ko_alf, so we use `ko`.
Note: With complex models and small sample sizes, convergence can be challenging.
The start_value_method = "pooled_nls" is often more robust for complex models.

```{r two_factor_add_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r two_factor_add, eval = !render_fast, include = !render_fast}
# This model might be slow or fail to converge with the small 'ko' and quick_nlme_control
# In a real vignette, use a dataset known to converge or increase iterations.
# For demonstration, we'll proceed.
fit_two_factors_add <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug", "dose"),
    factor_interaction = FALSE, # Additive effects
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_two_factors_add, "try-error")) {
  print(fit_two_factors_add)
} else {
  cat(
    "Two-factor additive model failed to converge with current settings for the vignette.\n"
  )
  fit_two_factors_add <- NULL # Ensure it's NULL for later checks
}
```

### Model with Two Factors and Interaction

This allows the effect of dose to be different for each drug (and vice-versa).

```{r two_factor_int_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r two_factor_int, eval = !render_fast, include = !render_fast}
fit_two_factors_int <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug", "dose"),
    factor_interaction = TRUE, # Interaction effect
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_two_factors_int, "try-error")) {
  print(fit_two_factors_int)
} else {
  cat(
    "Two-factor interaction model failed to converge with current settings for the vignette.\n"
  )
  fit_two_factors_int <- NULL
}
```

### Using Different equation_form and y_var

The simplified equation form expects y_var to be raw consumption.

```{r simplified_equation_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r simplified_equation, eval = !render_fast, include = !render_fast}
better_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100, # Max iterations for nlmeStruct optimization
  niterEM = 50, # Max EM iterations
  maxIter = 100, # Max iterations for the optim call for fixed effects
  pnlsTol = 1e-4, # Tolerance for PNLS convergence
  tolerance = 1e-7, # Tolerance for outer optimization loop
  opt = "nlminb",
  msVerbose = TRUE # Set to TRUE to see optimization steps
)

fit_simplified_example <- try(
  fit_demand_mixed(
    data = ko_alf, # Using simpler Alfentanil data
    y_var = "y", # Raw consumption
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    equation_form = "simplified",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_simplified_example, "try-error")) {
  print(fit_simplified_example)
} else {
  cat(
    "Simplified equation model failed to converge with current settings for the vignette.\n"
  )
  fit_simplified_example <- NULL
}
```

### Collapsing Factor Levels

Sometimes you might want to group levels of a factor, and you may want different
groupings for $Q_{0}$ versus $\alpha$. The `collapse_levels` argument allows you
to specify separate collapsing schemes for each parameter.

**Structure:**
```r
collapse_levels = list(
  Q0 = list(factor_name = list(new_level = c("old_level1", "old_level2"), ...)),
  alpha = list(factor_name = list(new_level = c("old_level1", ...), ...))
)
```

Either `Q0` or `alpha` (or both) can be omitted to keep original levels for that parameter.

#### Example: Same collapsing for both parameters

```{r}
# Ensure levels to collapse are present in ko_alf$dose
# levels(ko_alf$dose) are "0.001", "0.003", "3e-04"
fit_collapsed_same <- try(
  fit_demand_mixed(
    data = ko_alf,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    collapse_levels = list(
      Q0 = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      ),
      alpha = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      )
    ),
    equation_form = "zben",
    nlme_control = quick_nlme_control
  ),
  silent = TRUE
)

if (
  !is.null(fit_collapsed_same) &&
    !inherits(fit_collapsed_same, "try-error") &&
    !is.null(fit_collapsed_same$model)
) {
  print(fit_collapsed_same)
  cat("\nQ0 params:", fit_collapsed_same$param_info$num_params_Q0, "\n")
  cat("alpha params:", fit_collapsed_same$param_info$num_params_alpha, "\n")
} else {
  cat("Collapsed levels model failed to converge.\n")
}
```

#### Example: Different collapsing for Q0 and alpha

This is particularly useful when you want fine-grained distinctions for one
parameter but not the other. For instance, you might hypothesize that maximum
consumption ($Q_{0}$) varies by dose, but sensitivity ($\alpha$) does not.

```{r}
# Q0: keep 2 collapsed levels (low vs high)
# alpha: collapse all to 1 level (intercept only)
fit_collapsed_asymmetric <- try(
  fit_demand_mixed(
    data = ko_alf,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    collapse_levels = list(
      Q0 = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      ),
      alpha = list(dose = list(all_doses = c("3e-04", "0.001", "0.003")))
    ),
    equation_form = "zben",
    nlme_control = quick_nlme_control
  ),
  silent = TRUE
)

if (
  !is.null(fit_collapsed_asymmetric) &&
    !inherits(fit_collapsed_asymmetric, "try-error") &&
    !is.null(fit_collapsed_asymmetric$model)
) {
  print(fit_collapsed_asymmetric)
  cat("\nQ0 params:", fit_collapsed_asymmetric$param_info$num_params_Q0, "\n")
  cat(
    "alpha params:",
    fit_collapsed_asymmetric$param_info$num_params_alpha,
    "\n"
  )
  cat(
    "\nQ0 formula:",
    fit_collapsed_asymmetric$formula_details$fixed_effects_formula_str_Q0,
    "\n"
  )
  cat(
    "alpha formula:",
    fit_collapsed_asymmetric$formula_details$fixed_effects_formula_str_alpha,
    "\n"
  )
} else {
  cat("Asymmetric collapsed levels model failed to converge.\n")
}
```

**Note:** When all levels of a factor are collapsed to a single level for a
parameter, that factor is automatically removed from the formula for that
parameter (it contributes only to the intercept). A message will inform you
when this occurs.

#### EMMs and Comparisons with Collapsed Factors

When using `collapse_levels`, the `get_demand_param_emms()` and
`get_demand_comparisons()` functions automatically handle the asymmetric
factor structures. EMMs will be computed using the collapsed levels, and
comparisons will only be performed when there are multiple levels to compare.

**Differential Collapsing Output:** When Q0 and alpha have different numbers
of factor levels (e.g., Q0 keeps original `dose` levels while alpha collapses
to fewer groups), the EMM output will include separate factor columns:

- Original factor column (e.g., `dose`) for the uncollapsed parameter (Q0)
- Suffixed column (e.g., `dose_alpha`) for the collapsed parameter (alpha)

This results in a cross-joined table showing all combinations of Q0 levels
and alpha levels. For example, if dose has 5 original levels and alpha
collapses to 2 groups, the EMM table will have 10 rows (5 × 2).

```{r}
# Using the asymmetric collapsed model from above (if it converged)
if (
  !is.null(fit_collapsed_asymmetric) &&
    !inherits(fit_collapsed_asymmetric, "try-error") &&
    !is.null(fit_collapsed_asymmetric$model)
) {
  cat("--- EMMs with collapsed factors ---\n")
  # EMMs will show collapsed Q0 levels but single alpha value
  collapsed_emms <- get_demand_param_emms(
    fit_obj = fit_collapsed_asymmetric,
    factors_in_emm = "dose", # Use original factor name
    include_ev = TRUE
  )
  print(collapsed_emms)

  cat("\n--- Comparisons with collapsed factors ---\n")
  # Q0 will have comparisons (2 levels), alpha will have none (1 level)
  collapsed_comparisons <- get_demand_comparisons(
    fit_obj = fit_collapsed_asymmetric,
    compare_specs = ~dose, # Use original factor name
    params_to_compare = c("Q0", "alpha")
  )

  cat("\nQ0 contrasts (collapsed levels):\n")
  print(collapsed_comparisons$Q0$contrasts_log10)

  cat("\nalpha contrasts (intercept-only, no comparisons):\n")
  print(collapsed_comparisons$alpha$contrasts_log10)
} else {
  cat("Asymmetric collapsed model not available for EMM demonstration.\n")
}
```

### Inspecting Model Fits

Once a model is fit, you can inspect it using several S3 methods. We'll use `fit_one_factor_dose` if it converged.

```{r}
# Use a fit that is likely to have converged, e.g., fit_one_factor_dose
active_fit <- if (!is.null(fit_one_factor_dose$model)) {
  fit_one_factor_dose
} else {
  fit_no_factors_vignette
}

if (!is.null(active_fit$model)) {
  cat("--- Using S3 print() method: ---\n")
  print(active_fit)

  cat("\n--- Using S3 summary() method: ---\n")
  # The summary.beezdemand_nlme method calls summary.nlme and formats it
  model_summary <- summary(active_fit)
  print(model_summary)

  cat("\n--- Extracting Coefficients using S3 coef(): ---\n")
  cat("Fixed effects:\n")
  print(coef(active_fit, type = "fixed"))

  cat("\nRandom effects (deviations):\n")
  # Printing ranef can be long, so let's show head
  print(head(coef(active_fit, type = "random")))

  cat("\nSubject-specific coefficients (fixed + random):\n")
  print(head(coef(active_fit, type = "combined"))) # Default type

  cat("\nUsing S3 fixef() and ranef():\n")
  print(nlme::fixef(active_fit))
  print(utils::head(nlme::ranef(active_fit)))

  cat("\nStart values that were used for the NLME fit:\n")
  print(active_fit$start_values_used)
} else {
  cat("No suitable model converged for detailed inspection in the vignette.\n")
}
```

### Making Predictions (predict())

The S3 `predict()` method can generate population-level or group-level predictions.

```{r}
if (!is.null(active_fit$model)) {
  cat("--- Population-level predictions (log10 scale for 'zben'): ---\n")
  preds_pop_log <- predict(active_fit, level = 0)
  print(head(preds_pop_log))

  cat(
    "\n--- Population-level predictions (natural scale, back-transformed): ---\n"
  )
  # Assuming y_var was ll4, inv_fun is ll4_inv
  preds_pop_natural <- predict(
    active_fit,
    level = 0,
    inv_fun = ll4_inv
  )
  print(head(preds_pop_natural))

  cat(
    "\n--- Group-level predictions for first few data points (log10 scale): ---\n"
  )
  # Need to ensure newdata has factors with correct levels if different from original
  # For simplicity, predict on a subset of the original data
  sample_newdata <- active_fit$data[1:5, ]
  preds_group_log <- predict(active_fit, newdata = sample_newdata, level = 1)
  print(preds_group_log)
} else {
  cat("No suitable model converged for predictions in the vignette.\n")
}
```

## Visualizing Model Fits with `plot()`

The `beezdemand` package provides an S3 `plot()` method for `beezdemand_nlme`
objects, built using `ggplot2`, to help visualize the fitted demand curves against observed data.

**Key Features:**

*   **Observed Data:** Can display the original data points.

*   **Prediction Lines:** Plots model-predicted demand curves at the
population level (fixed effects, `pred_level = 0`) or group/subject level
(fixed + random effects, `pred_level = 1`).

*   **Inverse Transformation (`inv_fun`):** Allows back-transformation of
the y-axis and predictions to the natural scale (e.g., from log10 consumption
back to raw consumption units).

*   **Aesthetic Mapping:** Factors can be mapped to `color`, `linetype`
(for lines), and `shape` (for points).

*   **Faceting:** Supports `ggplot2` faceting via `facet_formula`.

*   **Axis Transformations:** Allows `x_trans` and `y_trans` (e.g., "log10", "pseudo_log").

### Example 1: Plotting a Single-Factor Model

Let's use `fit_one_factor_dose` (modeling demand for Alfentanil
by `dose`, with `y_ll4` as the dependent variable).

```{r plot_one_factor, cache = FALSE, fig.cap="Demand curves for Alfentanil by dose (population level). y-axis on natural scale."}
# Assuming fit_one_factor_dose was successfully created earlier in the vignette
if (!is.null(fit_one_factor_dose$model)) {
  # Define the inverse transformation function (if y_ll4 is log10(y))
  #   ll4_inv <- function(y_log10) 10^y_log10

  plot(
    fit_one_factor_dose,
    inv_fun = ll4_inv,
    color_by = "dose", # Color lines and points by dose
    shape_by = "dose", # Different shapes for points by dose
    observed_point_alpha = 0.7,
    title = "Alfentanil Demand by Dose (Population Fit)"
  )
} else {
  cat("fit_one_factor_dose model object not available for plotting.\n")
}
```

This plot shows the population-level demand curves for each dose of Alfentanil.
The y-axis has been back-transformed to the natural consumption scale using inv_fun.

### Example 2: Plotting Subject-Specific Lines

We can also visualize the individual subject fits by setting `show_pred_lines = "individual"`.

```{r plot_one_factor_individual, cache = FALSE}
if (!is.null(fit_one_factor_dose$model)) {
  plot(
    fit_one_factor_dose,
    show_pred_lines = "individual",
    inv_fun = ll4_inv,
    color_by = "dose", # Color lines by dose
    observed_point_alpha = 0.4,
    y_trans = "pseudo_log",
    ind_line_alpha = .5, # Make subject lines semi-transparent
    title = "Alfentanil Demand by Dose (Subject-Specific Fits)"
  ) +
    ggplot2::guides(color = guide_legend(override.aes = list(alpha = 1))) + # Make legend opaque
    facet_grid(~monkey)
} else {
  cat("fit_one_factor_dose model object not available for plotting.\n")
}
```

Here, each thin line represents the fitted curve for an individual subject (monkey), colored by dose.

### Example 3: Plotting a Two-Factor Model with Faceting

Let's use fit_two_factors_add (if it converged: y_ll4 ~ drug + dose). We can
facet by one factor and color by another.

```{r plot_two_factor_faceting, cache = FALSE}
# Assuming fit_two_factors_add was successfully created earlier
# If it failed, this chunk won't produce a plot.
active_two_factor_fit <- if (
  !is.null(fit_two_factors_add) && !is.null(fit_two_factors_add$model)
) {
  fit_two_factors_add
} else {
  # Fallback to a simpler model if the two-factor one failed in the vignette
  if (!is.null(fit_one_factor_dose$model)) fit_one_factor_dose else NULL
}

if (
  !is.null(active_two_factor_fit$model) &&
    !is.null(active_two_factor_fit$param_info$factors) &&
    length(active_two_factor_fit$param_info$factors) >= 1
) {
  # Determine factors for aesthetics based on what's in active_two_factor_fit
  color_factor <- if ("dose" %in% active_two_factor_fit$param_info$factors) {
    "dose"
  } else {
    NULL
  }
  facet_factor_name <- if (
    "drug" %in% active_two_factor_fit$param_info$factors
  ) {
    "drug"
  } else {
    # If only 'dose' is available from fit_one_factor_dose, we can't facet by 'drug'
    # So, maybe facet by dose instead, or don't facet.
    if (
      "dose" %in%
        active_two_factor_fit$param_info$factors &&
        is.null(color_factor)
    ) {
      color_factor <- "dose" # color by dose if not faceting by it
      NULL # No faceting
    } else {
      NULL
    }
  }

  facet_formula_plot <- if (!is.null(facet_factor_name)) {
    stats::as.formula(paste("~", facet_factor_name))
  } else {
    NULL
  }

  plot(
    active_two_factor_fit,
    inv_fun = ll4_inv,
    color_by = color_factor,
    # linetype_by = if("dose" %in% active_two_factor_fit$param_info$factors) "dose" else NULL, # Example
    facet_formula = facet_formula_plot,
    title = "Demand Curves (Population Fit)",
    observed_point_alpha = 0.5,
    ind_line_alpha = .5
  )
} else {
  cat(
    "A suitable two-factor or one-factor model object not available for this plotting example.\n"
  )
}
```

This example attempts to use fit_two_factors_add. If that model didn't converge
(common with minimal iterations for vignette speed), it falls back to
fit_one_factor_dose and adjusts aesthetics. The plot will show population lines,
colored by one factor and faceted by another (if two factors are available).

### Example 4: Customizing Axes (e.g., Log Scale)

You can use `x_trans` and `y_trans` for axis transformations.

```{r plot_one_factor_axes, cache = FALSE}
if (!is.null(fit_one_factor_dose$model)) {
  plot(
    fit_one_factor_dose,
    inv_fun = ll4_inv,
    color_by = "dose",
    x_trans = "pseudo_log", # Log10 transform the x-axis
    y_trans = "pseudo_log", # Pseudo-log transform the y-axis
    title = "Alfentanil Demand (Log10 Price Scale)"
  )
} else {
  cat("fit_one_factor_dose model object not available for plotting.\n")
}
```

### Example 5: Plotting a Model Fit with equation_form = "simplified"

If you fit a model using equation_form = "simplified" (which models raw y), the
inv_fun is typically identity because predictions are already on the natural scale.

```{r plot_simplified_model, cache = FALSE}
# Assuming fit_simplified_example converged earlier
if (
  !is.null(fit_simplified_example) && !is.null(fit_simplified_example$model)
) {
  plot(
    fit_simplified_example,
    inv_fun = identity, # Predictions are already on raw y scale
    color_by = "dose",
    shape_by = "dose",
    title = "Demand Model ('simplified' equation, Raw Y)"
  )
} else {
  cat("fit_simplified_example model object not available for plotting.\n")
}
```

These examples demonstrate the main capabilities of the
plot.beezdemand_nlme method. Users can further customize the returned ggplot
object by adding more layers or theme adjustments. For instance, to add custom
axis limits or breaks:

```r
plot_object +
    ggplot2::scale_x_continuous(
        limits = c(0, 1000),
        breaks = c(0, 100, 500, 1000)
    )
```

## Analyzing Estimated Marginal Means (`get_demand_param_emms`)

This function helps interpret how factors affect $Q_{0}$ and $\alpha$, providing estimates on both log10 and natural scales, and optionally Essential Value (EV).

**Note on `collapse_levels`:** When a model is fit using `collapse_levels` with
asymmetric specifications (different collapsing for Q0 and alpha), the EMM functions
automatically use the appropriate collapsed factor names for each parameter. You
still specify the original factor name in `factors_in_emm`, and the output will
show the collapsed levels. If a parameter has only one level (intercept-only),
that parameter's values will be the same across all rows.

```{r}
# We'll use a model with factors.
# If fit_two_factors_add converged, use it. Otherwise, use fit_one_factor_dose.
# For the vignette, let's ensure we use one that is likely available.
# If fit_two_factors_add is NULL (failed to converge in example), this will use fit_one_factor_dose
emm_model_to_use <- if (
  !is.null(fit_two_factors_add) && !is.null(fit_two_factors_add$model)
) {
  fit_two_factors_add
} else if (!is.null(fit_one_factor_dose$model)) {
  fit_one_factor_dose
} else {
  NULL
}

if (!is.null(emm_model_to_use)) {
  cat(
    "--- EMMs for model with factors:",
    paste(emm_model_to_use$param_info$factors, collapse = ", "),
    "---\n"
  )

  factors_for_emms <- emm_model_to_use$param_info$factors

  demand_emms_output <- get_demand_param_emms(
    fit_obj = emm_model_to_use,
    factors_in_emm = factors_for_emms, # Use factors from the model
    include_ev = TRUE
  )
  print(demand_emms_output)

  cat("\n--- EMMs for observed factor combinations only: ---\n")
  # This is useful if the EMM grid includes combinations not in data
  observed_demand_emms <- get_observed_demand_param_emms(
    fit_obj = emm_model_to_use,
    factors_in_emm = factors_for_emms,
    include_ev = TRUE
  )
  print(observed_demand_emms)
} else {
  cat(
    "No suitable model with factors converged for EMM analysis in the vignette.\n"
  )
}
```

## Performing Pairwise Comparisons (`get_demand_comparisons`)

Compare levels of factors for $Q_{0}$ and $\alpha$.

**Note on `collapse_levels`:** When using models fit with asymmetric
`collapse_levels`, comparisons are only performed for parameters that have
multiple levels. If a parameter was collapsed to a single level (intercept-only),
the comparisons for that parameter will be empty.

```{r}
# Using the same emm_model_to_use
if (
  !is.null(emm_model_to_use) && length(emm_model_to_use$param_info$factors) > 0
) {
  factors_present <- emm_model_to_use$param_info$factors

  if ("dose" %in% factors_present) {
    cat(
      "--- Pairwise comparisons for 'dose' (averaging over other factors if any): ---\n"
    )
    comparisons_dose <- get_demand_comparisons(
      fit_obj = emm_model_to_use,
      compare_specs = ~dose,
      contrast_type = "pairwise",
      adjust = "fdr",
      report_ratios = TRUE
    )
    print(comparisons_dose)
  }

  if (all(c("drug", "dose") %in% factors_present)) {
    cat(
      "\n--- Pairwise comparisons for 'drug' within each level of 'dose': ---\n"
    )
    # EMMs calculated over drug*dose, then contrast drug within each dose
    comparisons_drug_by_dose <- get_demand_comparisons(
      fit_obj = emm_model_to_use,
      compare_specs = ~ drug * dose,
      contrast_type = "pairwise",
      contrast_by = "dose", # Compare 'drug' levels, holding 'dose' constant
      adjust = "fdr",
      report_ratios = TRUE
    )
    print(comparisons_drug_by_dose)
  }
} else {
  cat(
    "No suitable model with factors converged for comparisons in the vignette.\n"
  )
}
```

## Advanced Topics

```{r advanced_topics_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "The sections below are omitted from the rendered vignette by default ",
    "(they can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run them.\n"
  )
}
```

### More Complex Random Effects Structures

This example shows how to specify a more complex random-effects structure (e.g.,
random slopes) for $Q_{0}$ and $\alpha$.

```{r complex_random_effects_fit, eval = !render_fast, include = !render_fast}
robust_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100,
  niterEM = 50,
  maxIter = 100,
  pnlsTol = 1e-4,
  tolerance = 1e-7,
  opt = "nlminb",
  msVerbose = TRUE # Set to FALSE for less output once working
)

fit_drug_slope_random <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey", # This is your 'monkey' identifier
    factors = c("drug", "dose"),
    factor_interaction = FALSE, # Additive fixed effects as requested
    equation_form = "zben",
    random_effects = Q0 + alpha ~ drug, # Random slopes by drug
    covariance_structure = "pdDiag",
    nlme_control = robust_nlme_control,
    start_value_method = "pooled_nls" # Good for more complex models
  ),
  silent = TRUE
)

if (
  !inherits(fit_drug_slope_random, "try-error") &&
    !is.null(fit_drug_slope_random$model)
) {
  print(fit_drug_slope_random)

  cat("\nFixed Effects:\n")
  print(nlme::fixef(fit_drug_slope_random))

  cat("\nRandom Effects Structure (Standard Deviations and Correlations):\n")
  print(nlme::VarCorr(fit_drug_slope_random$model))

  cat(
    "\nRandom Effect Estimates for each Monkey (deviations from fixed effects):\n"
  )
  print(utils::head(nlme::ranef(fit_drug_slope_random)))

  cat("\nSubject-Specific Coefficients (Fixed + Random, log10 scale):\n")
  print(head(coef(fit_drug_slope_random))) # type="combined" is default

  all_params_emms <- get_demand_param_emms(
    fit_obj = fit_drug_slope_random,
    factors_in_emm = c("drug", "dose"), # Get all cells
    at = list(monkey = "A"), # Condition on monkey A
    include_ev = TRUE # Or FALSE if not needed
  )

  observed_params_emms <- get_observed_demand_param_emms(
    fit_obj = fit_drug_slope_random,
    factors_in_emm = c("drug", "dose"), # Get all cells
    at = list(monkey = "A"), # Condition on monkey A
    include_ev = TRUE # Or FALSE if not needed
  )
} else {
  cat("Random-slope model failed to converge with the current settings.\n")
  fit_drug_slope_random <- NULL
}
```

Now let's demonstrate how to extract individual-level predicted coefficients using the
`get_individual_coefficients()` function. This combines fixed and random effects to
calculate individual-level parameter estimates for each subject.

```{r complex_random_effects_coefs, eval = !render_fast, include = !render_fast}
if (!is.null(fit_drug_slope_random) && !is.null(fit_drug_slope_random$model)) {
  # Get individual coefficients in wide format (default)
  individual_coefs_wide <- get_individual_coefficients(
    fit_drug_slope_random,
    params = c("Q0", "alpha"),
    format = "wide"
  )

  cat("Individual coefficients (wide format, log10 scale):\n")
  print(individual_coefs_wide)

  # Get individual coefficients in long format
  individual_coefs_long <- get_individual_coefficients(
    fit_drug_slope_random,
    params = c("Q0", "alpha"),
    format = "long"
  )

  cat("\nIndividual coefficients (long format, log10 scale):\n")
  print(head(individual_coefs_long, 10))

  # Convert to natural scale for interpretation
  individual_coefs_natural <- individual_coefs_wide

  # Convert Q0 coefficients to natural scale
  q0_cols <- grep(
    "^estimated_Q0_",
    names(individual_coefs_natural),
    value = TRUE
  )
  for (col in q0_cols) {
    natural_col <- gsub("estimated_Q0_", "Q0_natural_", col)
    individual_coefs_natural[[natural_col]] <- 10^individual_coefs_natural[[
      col
    ]]
  }

  # Convert alpha coefficients to natural scale
  alpha_cols <- grep(
    "^estimated_alpha_",
    names(individual_coefs_natural),
    value = TRUE
  )
  for (col in alpha_cols) {
    natural_col <- gsub("estimated_alpha_", "alpha_natural_", col)
    individual_coefs_natural[[natural_col]] <- 10^individual_coefs_natural[[
      col
    ]]
  }

  cat("\nExample individual coefficients on natural scale:\n")
  print(individual_coefs_natural[
    1:3,
    c(
      "id",
      grep("Q0_natural_", names(individual_coefs_natural), value = TRUE)[1:2],
      grep("alpha_natural_", names(individual_coefs_natural), value = TRUE)[1:2]
    )
  ])

  # You can also get coefficients for specific parameters only
  q0_only <- get_individual_coefficients(
    fit_drug_slope_random,
    params = "Q0",
    format = "wide"
  )

  cat("\nQ0 coefficients only:\n")
  print(q0_only)
} else {
  cat("Random-slope model not available for coefficient extraction.\n")
}
```

### Continuous Covariates and `fixed_rhs`

In addition to factor predictors, you can include continuous covariates in the fixed-effects
linear models for `Q0` and `alpha`.

There are two convenient ways to do this:

1. Add continuous covariates additively using the `continuous_covariates` argument (no formula writing required).
2. Specify a full right-hand-side using `fixed_rhs` (formula), which gives you complete control,
   including interactions (e.g., factor-by-continuous).

Below we demonstrate both, using a made-up continuous covariate `age` assigned per subject (`monkey`).

```{r continuous_setup, eval = !render_fast, include = !render_fast}
set.seed(123)

# Create a per-monkey continuous covariate: age
monkey_age <- unique(ko[, c("monkey")]) |>
  dplyr::mutate(age = round(runif(dplyr::n(), min = 18, max = 25), 1))

ko_age <- dplyr::left_join(ko, monkey_age, by = "monkey")
# Ensure a numeric version of dose is available for continuous modeling
ko_age <- ko_age |>
  dplyr::mutate(dose_num = suppressWarnings(as.numeric(as.character(dose))))

# Robust median for dose_num (fallbacks if all values are NA)
median_dose_num <- suppressWarnings(stats::median(
  ko_age$dose_num,
  na.rm = TRUE
))
if (is.na(median_dose_num) || !is.finite(median_dose_num)) {
  finite_vals <- ko_age$dose_num[is.finite(ko_age$dose_num)]
  if (length(finite_vals) > 0) {
    median_dose_num <- finite_vals[1]
  } else {
    median_dose_num <- 0
  }
}

# Sanity check
head(ko_age[, c("monkey", "age")])
```

#### A) Additive continuous covariate via `continuous_covariates`

Here we keep a single factor (`dose`) and add `age` as an additive continuous covariate. The RHS becomes
`~ 1 + dose + age` for both parameters.

```{r continuous_age_add, eval = !render_fast, include = !render_fast}
fit_age_add <- try(
  fit_demand_mixed(
    data = ko_age,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    continuous_covariates = c("age"),
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_age_add, "try-error") && !is.null(fit_age_add$model)) {
  print(fit_age_add)

  # EMMs over dose at age = 21
  emms_age_add <- get_demand_param_emms(
    fit_obj = fit_age_add,
    factors_in_emm = c("dose"),
    at = list(age = 21),
    include_ev = TRUE
  )
  cat("\nEMMs (age = 21):\n")
  print(emms_age_add)

  # Pairwise comparisons for dose at age = 21 (log10 scale; ratios reported too)
  comps_age_add <- get_demand_comparisons(
    fit_obj = fit_age_add,
    compare_specs = ~dose,
    contrast_type = "pairwise",
    at = list(age = 21),
    adjust = "fdr",
    report_ratios = TRUE
  )
  cat("\nPairwise comparisons for dose (age = 21):\n")
  print(comps_age_add)

  # Plot population lines at age = 21 (natural scale on y)
  plot(
    fit_age_add,
    at = list(age = 21),
    inv_fun = ll4_inv,
    color_by = "dose",
    x_trans = "pseudo_log",
    y_trans = "pseudo_log",
    title = "Demand by Dose with Age as Covariate (Population Fit at age = 21)"
  )
} else {
  cat(
    "\nAdditive continuous covariate model failed to converge for the vignette settings.\n"
  )
}
```

#### B) fixed_rhs with a factor (drug) and a continuous covariate (dose)

Here we specify `drug` as a factor but treat `dose` as a continuous covariate
(`dose_num`, a numeric version of `dose`),
and include `age` as an additional continuous covariate. This uses a shared RHS
for `Q0` and `alpha`:
`~ 1 + drug + dose_num + age`.

```{r continuous_age_rhs, eval = !render_fast, include = !render_fast}
fit_age_rhs <- try(
  fit_demand_mixed(
    data = ko_age,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug"), # only 'drug' is treated as a factor
    fixed_rhs = "~ 1 + drug + dose_num + age",
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_age_rhs, "try-error") && !is.null(fit_age_rhs$model)) {
  print(fit_age_rhs)

  # EMMs over drug (factor) at age = 21 and dose set to its median
  emms_age_rhs <- get_demand_param_emms(
    fit_obj = fit_age_rhs,
    factors_in_emm = c("drug"),
    at = list(age = 21, dose_num = median_dose_num),
    include_ev = TRUE
  )
  cat("\nEMMs (drug; age = 21, dose_num = median):\n")
  print(emms_age_rhs)

  # Pairwise comparisons: drug at age = 21 and dose set to its median
  comps_drug_by_dose_age <- get_demand_comparisons(
    fit_obj = fit_age_rhs,
    compare_specs = ~drug,
    contrast_type = "pairwise",
    at = list(age = 21, dose_num = median_dose_num),
    adjust = "fdr",
    report_ratios = TRUE
  )
  cat("\nPairwise comparisons of drug (age = 21, dose_num = median):\n")
  print(comps_drug_by_dose_age)

  # Plot population lines at age = 21 and dose_num = median, color by drug and facet by drug
  plot(
    fit_age_rhs,
    at = list(age = 21, dose_num = median_dose_num),
    inv_fun = ll4_inv,
    color_by = "drug",
    facet_formula = ~drug,
    x_trans = "pseudo_log",
    y_trans = "pseudo_log",
    title = "Demand by Drug with Dose Continuous (Population Fit at age = 21; dose_num = median)"
  )
} else {
  cat(
    "\nfixed_rhs model (drug factor, dose continuous) failed to converge for the vignette settings.\n"
  )
}
```

### Trends with `emtrends`

We can examine how the parameters change with continuous covariates using
`get_demand_param_trends()`, which wraps `emmeans::emtrends()` and returns
tidy results for Q0 and alpha trends on the log10 scale.

Below we compute trends with respect to `age` and `dose_num`, first overall and
then by `drug` (if the model includes `drug` as a factor).

```{r trends, eval = !render_fast, include = !render_fast}
# Choose a fitted model for trends (prefer fixed_rhs fit)
fit_for_trends <- NULL
if (
  exists("fit_age_rhs") && !is.null(fit_age_rhs) && !is.null(fit_age_rhs$model)
) {
  fit_for_trends <- fit_age_rhs
} else if (
  exists("fit_age_add") && !is.null(fit_age_add) && !is.null(fit_age_add$model)
) {
  fit_for_trends <- fit_age_add
}

if (!is.null(fit_for_trends)) {
  # Determine which of the candidate covariates are actually present in the model RHS
  # Use Q0 formula (or alpha if Q0 is NULL) - they may differ if collapse_levels was used
  rhs_formula_str <- fit_for_trends$formula_details$fixed_effects_formula_str_Q0
  if (
    is.null(rhs_formula_str) ||
      is.na(rhs_formula_str) ||
      !nzchar(rhs_formula_str)
  ) {
    rhs_formula_str <- fit_for_trends$formula_details$fixed_effects_formula_str_alpha
  }
  rhs_vars_tr <- tryCatch(
    all.vars(as.formula(rhs_formula_str)),
    error = function(e) character(0)
  )
  covariates_candidate <- c("age", "dose_num")
  covariates_in_model <- intersect(rhs_vars_tr, covariates_candidate)

  if (length(covariates_in_model) > 0) {
    cat("\n--- Trends (overall) at age = 21 and dose_num = median ---\n")
    trends_overall <- get_demand_param_trends(
      fit_obj = fit_for_trends,
      params = c("Q0", "alpha"),
      covariates = covariates_in_model,
      specs = ~1,
      at = list(age = 21, dose_num = median_dose_num)
    )
    print(trends_overall)

    # If 'drug' is a model factor, show trends by drug as well
    model_factors_tr <- fit_for_trends$param_info$factors
    if (!is.null(model_factors_tr) && ("drug" %in% model_factors_tr)) {
      cat("\n--- Trends by drug at age = 21 and dose_num = median ---\n")
      trends_by_drug <- get_demand_param_trends(
        fit_obj = fit_for_trends,
        params = c("Q0", "alpha"),
        covariates = covariates_in_model,
        specs = ~drug,
        at = list(age = 21, dose_num = median_dose_num)
      )
      print(trends_by_drug)
    } else {
      cat(
        "\nNote: The selected model does not include 'drug' as a factor; skipping by-drug trends.\n"
      )
    }
  } else {
    cat(
      "\nNo candidate continuous covariates ('age', 'dose_num') are present in the model RHS; skipping trends.\n"
    )
  }
} else {
  cat(
    "\nNo suitable fitted model available for computing trends in the vignette.\n"
  )
}
```

# Conclusion

The beezdemand package provides a suite of tools for robustly fitting
nonlinear mixed-effects demand models and interpreting their parameters. By
parameterizing $Q_{0}$ and $\alpha$ on the log10 scale, numerical stability is enhanced,
while helper functions allow for easy back-transformation and interpretation on
the natural scale.

Further exploration might involve trying different random effects structures,
more complex fixed effects, or comparing model fits using AIC/BIC.
