---
title: "Mixed-Effects Demand Modeling with `beezdemand`"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Effects Demand Modeling with `beezdemand`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",   # R console style comments
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,  # Suppress warnings for cleaner output
  message = FALSE,  # Suppress messages for cleaner output
  cache = TRUE,     # Cache for faster rebuilds
  cache.path = "mixed-demand_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)

data("apt", package = "beezdemand", envir = environment())
data("ko", package = "beezdemand", envir = environment())

cache_key_object <- function(x) {
  tmp <- tempfile(fileext = ".rds")
  saveRDS(x, tmp)
  on.exit(unlink(tmp), add = TRUE)
  unname(tools::md5sum(tmp))
}

# Invalidate cache when the package data changes.
knitr::opts_chunk$set(
  cache.extra = list(
    beezdemand_version = as.character(utils::packageVersion("beezdemand")),
    apt = cache_key_object(apt),
    ko = cache_key_object(ko)
  )
)
```

# Introduction

This vignette demonstrates how to perform mixed-effects nonlinear modeling
of behavioral economic demand data using the beezdemand package. We will focus
on the `fit_demand_mixed()` function and its associated helper functions for
extracting results, making predictions, and visualizing fits. These models
allow for individual differences (random effects) and the examination of how
various factors (fixed effects) influence demand parameters like $Q_{0}$ (maximum
consumption at zero price) and $\alpha$ (sensitivity of demand to price). The
parameters $Q_{0}$ and $\alpha$ are estimated on a log10 scale for numerical stability,
but reporting functions will provide them on their natural, interpretable scale.

For advanced topics including multi-factor models, collapsing factor levels,
estimated marginal means, pairwise comparisons, continuous covariates, and
complex random effects structures, see `vignette("mixed-demand-advanced")`.

## Data Preparation

For these examples, we will use the apt and ko datasets, which is
assumed to be available and pre-processed.

The apt dataset should contain:

- id: A unique identifier for each subject.

- x: The price of the drug.

- y: The consumption of the drug.

The ko dataset should contain:

- monkey: A subject or group identifier for random effects.

- x: The price of the commodity (in this case the fixed-ratio requirement).

- y: Raw consumption values. This is typically used with the simplified exponentiated equation.

- y_ll4: Consumption, ll4 transformed. This is typically used with the zben equation.

- Factor columns like drug and dose.


```{r}
quick_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100,
  niterEM = 20,
  maxIter = 100, # Low iterations for speed
  pnlsTol = 0.1,
  tolerance = 1e-4, # Looser tolerance
  opt = "nlminb",
  msVerbose = FALSE
)
```

## Fitting Demand Models with fit_demand_mixed()

The core function for fitting nonlinear mixed-effects demand models is `fit_demand_mixed()`.

### APT Fit and Plot

#### LL4 transformation with ZBEn

```{r apt_zben_fit}
apt_ll4 <- apt |>
  mutate(y_ll4 = ll4(y))

fit_apt_zben <- fit_demand_mixed(
  data = apt_ll4,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "id",
  equation_form = "zben",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_zben)
```

```{r apt_zben_plot, cache = FALSE}
plot(
  fit_apt_zben,
  inv_fun = ll4_inv,
  y_trans = "pseudo_log",
  x_trans = "pseudo_log",
  show_pred_lines = c("population", "individual")
) +
  facet_wrap(
    ~id
  )
```

#### Simplified Exponential

```{r apt_simplified_fit}
fit_apt_simplified <- fit_demand_mixed(
  data = apt_ll4,
  y_var = "y",
  x_var = "x",
  id_var = "id",
  equation_form = "simplified",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_simplified)
```

```{r apt_simplified_plot, cache = FALSE}
plot(
  fit_apt_simplified,
  x_trans = "pseudo_log",
  show_pred_lines = c("population", "individual")
) +
  facet_wrap(
    ~id
  )
```

#### Koffarnus (Exponentiated) Equation Form

`fit_demand_mixed()` also supports the Koffarnus et al. (2015) equation via
`equation_form = "exponentiated"`. By default, the scaling constant `k` will be computed
from the data range (you can also specify it directly).

```{r apt_exponentiated_fit}
fit_apt_exponentiated <- fit_demand_mixed(
  data = apt,
  y_var = "y",
  x_var = "x",
  id_var = "id",
  equation_form = "exponentiated",
  k = NULL,
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_apt_exponentiated)
```

## Inspecting Fits (tidy / glance / augment)

All modern model classes support `tidy()`, `glance()`, and `augment()` to
standardize programmatic access to estimates, model summaries, and residuals.

```{r inspect_fit}
glance(fit_apt_zben)
tidy(fit_apt_zben) |> head()
augment(fit_apt_zben) |> head()
```

## Diagnostics

Use `check_demand_model()` and the residual plotting helpers as standard
post-fit checks.

```{r diagnostics}
check_demand_model(fit_apt_zben)
plot_residuals(fit_apt_zben)$fitted
```

### Basic Model (No Factors)

This model estimates global $Q_{0}$ and $\alpha$ parameters with random effects for subjects.

```{r}
# Make sure a similar 'fit_no_factors' was created successfully in your environment
# For the vignette, let's create one that is more likely to converge quickly
# by using only Alfentanil data, which is less complex than the full dataset.
ko_alf <- ko[ko$drug == "Alfentanil", ]

fit_no_factors_vignette <- fit_demand_mixed(
  data = ko_alf,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "monkey",
  equation_form = "zben",
  nlme_control = quick_nlme_control, # Use quicker control for vignette
  start_value_method = "heuristic" # Heuristic is faster for simple model
)
print(fit_no_factors_vignette)
```

The output shows the model call, selected equation form, and if the model converged, it prints the nlme model summary.

### Model with One Factor

Let's model $Q_{0}$ and $\alpha$ as varying by dose for Alfentanil.

```{r}
fit_one_factor_dose <- fit_demand_mixed(
  data = ko_alf,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "monkey",
  factors = "dose",
  equation_form = "zben",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
print(fit_one_factor_dose)
```

### Inspecting Model Fits

Once a model is fit, you can inspect it using several S3 methods.

```{r}
# Summary
summary(fit_one_factor_dose)

# Fixed effects
coef(fit_one_factor_dose, type = "fixed")

# Random effects (deviations from fixed)
head(coef(fit_one_factor_dose, type = "random"))

# Subject-specific coefficients (fixed + random)
head(coef(fit_one_factor_dose, type = "combined"))

# Access nlme fixef/ranef directly
nlme::fixef(fit_one_factor_dose)
utils::head(nlme::ranef(fit_one_factor_dose))

# Start values that were used for the NLME fit
fit_one_factor_dose$start_values_used
```

### Making Predictions (predict())

The S3 `predict()` method can generate population-level or group-level predictions.

```{r}
# Population-level predictions (log10 scale for 'zben')
preds_pop_log <- predict(fit_one_factor_dose, level = 0)
head(preds_pop_log)

# Population-level predictions (natural scale, back-transformed)
preds_pop_natural <- predict(
  fit_one_factor_dose,
  level = 0,
  inv_fun = ll4_inv
)
head(preds_pop_natural)

# Group-level predictions for first few data points
sample_newdata <- fit_one_factor_dose$data[1:5, ]
preds_group_log <- predict(fit_one_factor_dose, newdata = sample_newdata, level = 1)
preds_group_log
```

## Visualizing Model Fits with `plot()`

The `beezdemand` package provides an S3 `plot()` method for `beezdemand_nlme`
objects, built using `ggplot2`, to help visualize the fitted demand curves against observed data.

**Key Features:**

*   **Observed Data:** Can display the original data points.

*   **Prediction Lines:** Plots model-predicted demand curves at the
population level (fixed effects, `pred_level = 0`) or group/subject level
(fixed + random effects, `pred_level = 1`).

*   **Inverse Transformation (`inv_fun`):** Allows back-transformation of
the y-axis and predictions to the natural scale (e.g., from log10 consumption
back to raw consumption units).

*   **Aesthetic Mapping:** Factors can be mapped to `color`, `linetype`
(for lines), and `shape` (for points).

*   **Faceting:** Supports `ggplot2` faceting via `facet_formula`.

*   **Axis Transformations:** Allows `x_trans` and `y_trans` (e.g., "log10", "pseudo_log").

### Example 1: Plotting a Single-Factor Model

Let's use `fit_one_factor_dose` (modeling demand for Alfentanil
by `dose`, with `y_ll4` as the dependent variable).

```{r plot_one_factor, cache = FALSE, fig.cap="Demand curves for Alfentanil by dose (population level). y-axis on natural scale."}
plot(
  fit_one_factor_dose,
  inv_fun = ll4_inv,
  color_by = "dose",
  shape_by = "dose",
  observed_point_alpha = 0.7,
  title = "Alfentanil Demand by Dose (Population Fit)"
)
```

This plot shows the population-level demand curves for each dose of Alfentanil.
The y-axis has been back-transformed to the natural consumption scale using inv_fun.

### Example 2: Plotting Subject-Specific Lines

We can also visualize the individual subject fits by setting `show_pred_lines = "individual"`.

```{r plot_one_factor_individual, cache = FALSE}
plot(
  fit_one_factor_dose,
  show_pred_lines = "individual",
  inv_fun = ll4_inv,
  color_by = "dose",
  observed_point_alpha = 0.4,
  y_trans = "pseudo_log",
  ind_line_alpha = .5,
  title = "Alfentanil Demand by Dose (Subject-Specific Fits)"
) +
  ggplot2::guides(color = guide_legend(override.aes = list(alpha = 1))) +
  facet_grid(~monkey)
```

Here, each thin line represents the fitted curve for an individual subject (monkey), colored by dose.

### Example 3: Customizing Axes (e.g., Log Scale)

You can use `x_trans` and `y_trans` for axis transformations.

```{r plot_one_factor_axes, cache = FALSE}
plot(
  fit_one_factor_dose,
  inv_fun = ll4_inv,
  color_by = "dose",
  x_trans = "pseudo_log",
  y_trans = "pseudo_log",
  title = "Alfentanil Demand (Log10 Price Scale)"
)
```

Users can further customize the returned ggplot
object by adding more layers or theme adjustments. For instance, to add custom
axis limits or breaks:

```r
plot_object +
    ggplot2::scale_x_continuous(
        limits = c(0, 1000),
        breaks = c(0, 100, 500, 1000)
    )
```

# Conclusion

The beezdemand package provides a suite of tools for robustly fitting
nonlinear mixed-effects demand models and interpreting their parameters. By
parameterizing $Q_{0}$ and $\alpha$ on the log10 scale, numerical stability is enhanced,
while helper functions allow for easy back-transformation and interpretation on
the natural scale.

# See Also

- `vignette("mixed-demand-advanced")` -- Multi-factor models, collapsing levels, EMMs, comparisons, covariates, and trends
- `vignette("model-selection")` -- Choosing the right model class
- `vignette("hurdle-demand-models")` -- Two-part hurdle demand models
- `vignette("beezdemand")` -- Getting started with beezdemand
