---
title: "Advanced Mixed-Effects Demand Modeling"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Mixed-Effects Demand Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE, # Keep code chunks separate for clarity
  comment = "#>",   # R console style comments
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,  # Suppress warnings for cleaner output
  message = FALSE,  # Suppress messages for cleaner output
  cache = TRUE,     # Cache for faster rebuilds
  cache.path = "mixed-demand-advanced_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)

data("apt", package = "beezdemand", envir = environment())
data("ko", package = "beezdemand", envir = environment())

# Set `BEEZDEMAND_VIGNETTE_MODE=full` to run the slower / less stable examples.
render_fast <- tolower(Sys.getenv("BEEZDEMAND_VIGNETTE_MODE", "fast")) != "full"

cache_key_object <- function(x) {
  tmp <- tempfile(fileext = ".rds")
  saveRDS(x, tmp)
  on.exit(unlink(tmp), add = TRUE)
  unname(tools::md5sum(tmp))
}

# Invalidate cache when the package data changes.
knitr::opts_chunk$set(
  cache.extra = list(
    beezdemand_version = as.character(utils::packageVersion("beezdemand")),
    apt = cache_key_object(apt),
    ko = cache_key_object(ko)
  )
)

# Optional fits used later (kept NULL in fast renders).
fit_two_factors_add <- NULL
fit_two_factors_int <- NULL
fit_simplified_example <- NULL
fit_collapsed_same <- NULL
fit_collapsed_asymmetric <- NULL
fit_drug_slope_random <- NULL
fit_age_add <- NULL
fit_age_rhs <- NULL
```

# Introduction

This vignette covers advanced topics for mixed-effects nonlinear demand
modeling with `beezdemand`. It assumes you are already familiar with the basics
covered in `vignette("mixed-demand")`, including fitting models with
`fit_demand_mixed()`, inspecting results with `tidy()` / `glance()` /
`augment()`, and basic plotting.

Topics covered here include:

- **Multi-factor models** (additive and interaction)
- **Collapsing factor levels** (asymmetric specifications for Q0 and alpha)
- **Estimated marginal means** and **pairwise comparisons**
- **Multi-factor visualization** with faceting
- **Complex random effects structures**
- **Continuous covariates** and `fixed_rhs`
- **Trends** with `emtrends`

```{r shared-setup}
quick_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100,
  niterEM = 20,
  maxIter = 100, # Low iterations for speed
  pnlsTol = 0.1,
  tolerance = 1e-4, # Looser tolerance
  opt = "nlminb",
  msVerbose = FALSE
)

# Prepare data subsets used throughout
ko_alf <- ko[ko$drug == "Alfentanil", ]

# Fit the one-factor model used as a baseline/fallback in later sections
fit_one_factor_dose <- fit_demand_mixed(
  data = ko_alf,
  y_var = "y_ll4",
  x_var = "x",
  id_var = "monkey",
  factors = "dose",
  equation_form = "zben",
  nlme_control = quick_nlme_control,
  start_value_method = "heuristic"
)
```

## Model with Two Factors (Additive)

Here, $Q_{0}$ and $\alpha$ vary by drug and dose additively.
For this, we need more data than just ko_alf, so we use `ko`.
Note: With complex models and small sample sizes, convergence can be challenging.
The start_value_method = "pooled_nls" is often more robust for complex models.

```{r two_factor_add_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r two_factor_add, eval = !render_fast, include = !render_fast}
# This model might be slow or fail to converge with the small 'ko' and quick_nlme_control
# In a real vignette, use a dataset known to converge or increase iterations.
# For demonstration, we'll proceed.
fit_two_factors_add <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug", "dose"),
    factor_interaction = FALSE, # Additive effects
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_two_factors_add, "try-error")) {
  print(fit_two_factors_add)
} else {
  cat(
    "Two-factor additive model failed to converge with current settings for the vignette.\n"
  )
  fit_two_factors_add <- NULL # Ensure it's NULL for later checks
}
```

### Model with Two Factors and Interaction

This allows the effect of dose to be different for each drug (and vice-versa).

```{r two_factor_int_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r two_factor_int, eval = !render_fast, include = !render_fast}
fit_two_factors_int <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug", "dose"),
    factor_interaction = TRUE, # Interaction effect
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_two_factors_int, "try-error")) {
  print(fit_two_factors_int)
} else {
  cat(
    "Two-factor interaction model failed to converge with current settings for the vignette.\n"
  )
  fit_two_factors_int <- NULL
}
```

### Using Different equation_form and y_var

The simplified equation form expects y_var to be raw consumption.

```{r simplified_equation_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "This example is omitted from the rendered vignette (it can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run it.\n"
  )
}
```

```{r simplified_equation, eval = !render_fast, include = !render_fast}
better_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100, # Max iterations for nlmeStruct optimization
  niterEM = 50, # Max EM iterations
  maxIter = 100, # Max iterations for the optim call for fixed effects
  pnlsTol = 1e-4, # Tolerance for PNLS convergence
  tolerance = 1e-7, # Tolerance for outer optimization loop
  opt = "nlminb",
  msVerbose = TRUE # Set to TRUE to see optimization steps
)

fit_simplified_example <- try(
  fit_demand_mixed(
    data = ko_alf, # Using simpler Alfentanil data
    y_var = "y", # Raw consumption
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    equation_form = "simplified",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_simplified_example, "try-error")) {
  print(fit_simplified_example)
} else {
  cat(
    "Simplified equation model failed to converge with current settings for the vignette.\n"
  )
  fit_simplified_example <- NULL
}
```

## Collapsing Factor Levels

Sometimes you might want to group levels of a factor, and you may want different
groupings for $Q_{0}$ versus $\alpha$. The `collapse_levels` argument allows you
to specify separate collapsing schemes for each parameter.

**Structure:**
```r
collapse_levels = list(
  Q0 = list(factor_name = list(new_level = c("old_level1", "old_level2"), ...)),
  alpha = list(factor_name = list(new_level = c("old_level1", ...), ...))
)
```

Either `Q0` or `alpha` (or both) can be omitted to keep original levels for that parameter.

### Example: Same collapsing for both parameters

```{r}
# Ensure levels to collapse are present in ko_alf$dose
# levels(ko_alf$dose) are "0.001", "0.003", "3e-04"
fit_collapsed_same <- try(
  fit_demand_mixed(
    data = ko_alf,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    collapse_levels = list(
      Q0 = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      ),
      alpha = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      )
    ),
    equation_form = "zben",
    nlme_control = quick_nlme_control
  ),
  silent = TRUE
)

if (
  !is.null(fit_collapsed_same) &&
    !inherits(fit_collapsed_same, "try-error") &&
    !is.null(fit_collapsed_same$model)
) {
  print(fit_collapsed_same)
  cat("\nQ0 params:", fit_collapsed_same$param_info$num_params_Q0, "\n")
  cat("alpha params:", fit_collapsed_same$param_info$num_params_alpha, "\n")
} else {
  cat("Collapsed levels model failed to converge.\n")
}
```

### Example: Different collapsing for Q0 and alpha

This is particularly useful when you want fine-grained distinctions for one
parameter but not the other. For instance, you might hypothesize that maximum
consumption ($Q_{0}$) varies by dose, but sensitivity ($\alpha$) does not.

```{r}
# Q0: keep 2 collapsed levels (low vs high)
# alpha: collapse all to 1 level (intercept only)
fit_collapsed_asymmetric <- try(
  fit_demand_mixed(
    data = ko_alf,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    collapse_levels = list(
      Q0 = list(
        dose = list(low_doses = c("3e-04", "0.001"), high_dose = "0.003")
      ),
      alpha = list(dose = list(all_doses = c("3e-04", "0.001", "0.003")))
    ),
    equation_form = "zben",
    nlme_control = quick_nlme_control
  ),
  silent = TRUE
)

if (
  !is.null(fit_collapsed_asymmetric) &&
    !inherits(fit_collapsed_asymmetric, "try-error") &&
    !is.null(fit_collapsed_asymmetric$model)
) {
  print(fit_collapsed_asymmetric)
  cat("\nQ0 params:", fit_collapsed_asymmetric$param_info$num_params_Q0, "\n")
  cat(
    "alpha params:",
    fit_collapsed_asymmetric$param_info$num_params_alpha,
    "\n"
  )
  cat(
    "\nQ0 formula:",
    fit_collapsed_asymmetric$formula_details$fixed_effects_formula_str_Q0,
    "\n"
  )
  cat(
    "alpha formula:",
    fit_collapsed_asymmetric$formula_details$fixed_effects_formula_str_alpha,
    "\n"
  )
} else {
  cat("Asymmetric collapsed levels model failed to converge.\n")
}
```

**Note:** When all levels of a factor are collapsed to a single level for a
parameter, that factor is automatically removed from the formula for that
parameter (it contributes only to the intercept). A message will inform you
when this occurs.

### EMMs and Comparisons with Collapsed Factors

When using `collapse_levels`, the `get_demand_param_emms()` and
`get_demand_comparisons()` functions automatically handle the asymmetric
factor structures. EMMs will be computed using the collapsed levels, and
comparisons will only be performed when there are multiple levels to compare.

**Differential Collapsing Output:** When Q0 and alpha have different numbers
of factor levels (e.g., Q0 keeps original `dose` levels while alpha collapses
to fewer groups), the EMM output will include separate factor columns:

- Original factor column (e.g., `dose`) for the uncollapsed parameter (Q0)
- Suffixed column (e.g., `dose_alpha`) for the collapsed parameter (alpha)

This results in a cross-joined table showing all combinations of Q0 levels
and alpha levels. For example, if dose has 5 original levels and alpha
collapses to 2 groups, the EMM table will have 10 rows (5 × 2).

```{r}
# Using the asymmetric collapsed model from above (if it converged)
if (
  !is.null(fit_collapsed_asymmetric) &&
    !inherits(fit_collapsed_asymmetric, "try-error") &&
    !is.null(fit_collapsed_asymmetric$model)
) {
  cat("--- EMMs with collapsed factors ---\n")
  # EMMs will show collapsed Q0 levels but single alpha value
  collapsed_emms <- get_demand_param_emms(
    fit_obj = fit_collapsed_asymmetric,
    factors_in_emm = "dose", # Use original factor name
    include_ev = TRUE
  )
  print(collapsed_emms)

  cat("\n--- Comparisons with collapsed factors ---\n")
  # Q0 will have comparisons (2 levels), alpha will have none (1 level)
  collapsed_comparisons <- get_demand_comparisons(
    fit_obj = fit_collapsed_asymmetric,
    compare_specs = ~dose, # Use original factor name
    params_to_compare = c("Q0", "alpha")
  )

  cat("\nQ0 contrasts (collapsed levels):\n")
  print(collapsed_comparisons$Q0$contrasts_log10)

  cat("\nalpha contrasts (intercept-only, no comparisons):\n")
  print(collapsed_comparisons$alpha$contrasts_log10)
} else {
  cat("Asymmetric collapsed model not available for EMM demonstration.\n")
}
```

## Visualizing Multi-Factor Models

### Plotting a Two-Factor Model with Faceting

Let's use fit_two_factors_add (if it converged: y_ll4 ~ drug + dose). We can
facet by one factor and color by another.

```{r plot_two_factor_faceting, cache = FALSE}
# Assuming fit_two_factors_add was successfully created earlier
# If it failed, this chunk won't produce a plot.
active_two_factor_fit <- if (
  !is.null(fit_two_factors_add) && !is.null(fit_two_factors_add$model)
) {
  fit_two_factors_add
} else {
  # Fallback to a simpler model if the two-factor one failed in the vignette
  if (!is.null(fit_one_factor_dose$model)) fit_one_factor_dose else NULL
}

if (
  !is.null(active_two_factor_fit$model) &&
    !is.null(active_two_factor_fit$param_info$factors) &&
    length(active_two_factor_fit$param_info$factors) >= 1
) {
  # Determine factors for aesthetics based on what's in active_two_factor_fit
  color_factor <- if ("dose" %in% active_two_factor_fit$param_info$factors) {
    "dose"
  } else {
    NULL
  }
  facet_factor_name <- if (
    "drug" %in% active_two_factor_fit$param_info$factors
  ) {
    "drug"
  } else {
    # If only 'dose' is available from fit_one_factor_dose, we can't facet by 'drug'
    # So, maybe facet by dose instead, or don't facet.
    if (
      "dose" %in%
        active_two_factor_fit$param_info$factors &&
        is.null(color_factor)
    ) {
      color_factor <- "dose" # color by dose if not faceting by it
      NULL # No faceting
    } else {
      NULL
    }
  }

  facet_formula_plot <- if (!is.null(facet_factor_name)) {
    stats::as.formula(paste("~", facet_factor_name))
  } else {
    NULL
  }

  plot(
    active_two_factor_fit,
    inv_fun = ll4_inv,
    color_by = color_factor,
    # linetype_by = if("dose" %in% active_two_factor_fit$param_info$factors) "dose" else NULL, # Example
    facet_formula = facet_formula_plot,
    title = "Demand Curves (Population Fit)",
    observed_point_alpha = 0.5,
    ind_line_alpha = .5
  )
} else {
  cat(
    "A suitable two-factor or one-factor model object not available for this plotting example.\n"
  )
}
```

This example attempts to use fit_two_factors_add. If that model didn't converge
(common with minimal iterations for vignette speed), it falls back to
fit_one_factor_dose and adjusts aesthetics. The plot will show population lines,
colored by one factor and faceted by another (if two factors are available).

### Plotting a Model Fit with equation_form = "simplified"

If you fit a model using equation_form = "simplified" (which models raw y), the
inv_fun is typically identity because predictions are already on the natural scale.

```{r plot_simplified_model, cache = FALSE}
# Assuming fit_simplified_example converged earlier
if (
  !is.null(fit_simplified_example) && !is.null(fit_simplified_example$model)
) {
  plot(
    fit_simplified_example,
    inv_fun = identity, # Predictions are already on raw y scale
    color_by = "dose",
    shape_by = "dose",
    title = "Demand Model ('simplified' equation, Raw Y)"
  )
} else {
  cat("fit_simplified_example model object not available for plotting.\n")
}
```

Users can further customize the returned ggplot
object by adding more layers or theme adjustments. For instance, to add custom
axis limits or breaks:

```r
plot_object +
    ggplot2::scale_x_continuous(
        limits = c(0, 1000),
        breaks = c(0, 100, 500, 1000)
    )
```

## Analyzing Estimated Marginal Means (`get_demand_param_emms`)

This function helps interpret how factors affect $Q_{0}$ and $\alpha$, providing estimates on both log10 and natural scales, and optionally Essential Value (EV).

**Note on `collapse_levels`:** When a model is fit using `collapse_levels` with
asymmetric specifications (different collapsing for Q0 and alpha), the EMM functions
automatically use the appropriate collapsed factor names for each parameter. You
still specify the original factor name in `factors_in_emm`, and the output will
show the collapsed levels. If a parameter has only one level (intercept-only),
that parameter's values will be the same across all rows.

```{r}
# We'll use a model with factors.
# If fit_two_factors_add converged, use it. Otherwise, use fit_one_factor_dose.
# For the vignette, let's ensure we use one that is likely available.
# If fit_two_factors_add is NULL (failed to converge in example), this will use fit_one_factor_dose
emm_model_to_use <- if (
  !is.null(fit_two_factors_add) && !is.null(fit_two_factors_add$model)
) {
  fit_two_factors_add
} else if (!is.null(fit_one_factor_dose$model)) {
  fit_one_factor_dose
} else {
  NULL
}

if (!is.null(emm_model_to_use)) {
  cat(
    "--- EMMs for model with factors:",
    paste(emm_model_to_use$param_info$factors, collapse = ", "),
    "---\n"
  )

  factors_for_emms <- emm_model_to_use$param_info$factors

  demand_emms_output <- get_demand_param_emms(
    fit_obj = emm_model_to_use,
    factors_in_emm = factors_for_emms, # Use factors from the model
    include_ev = TRUE
  )
  print(demand_emms_output)

  cat("\n--- EMMs for observed factor combinations only: ---\n")
  # This is useful if the EMM grid includes combinations not in data
  observed_demand_emms <- get_observed_demand_param_emms(
    fit_obj = emm_model_to_use,
    factors_in_emm = factors_for_emms,
    include_ev = TRUE
  )
  print(observed_demand_emms)
} else {
  cat(
    "No suitable model with factors converged for EMM analysis in the vignette.\n"
  )
}
```

## Performing Pairwise Comparisons (`get_demand_comparisons`)

Compare levels of factors for $Q_{0}$ and $\alpha$.

**Note on `collapse_levels`:** When using models fit with asymmetric
`collapse_levels`, comparisons are only performed for parameters that have
multiple levels. If a parameter was collapsed to a single level (intercept-only),
the comparisons for that parameter will be empty.

```{r}
# Using the same emm_model_to_use
if (
  !is.null(emm_model_to_use) && length(emm_model_to_use$param_info$factors) > 0
) {
  factors_present <- emm_model_to_use$param_info$factors

  if ("dose" %in% factors_present) {
    cat(
      "--- Pairwise comparisons for 'dose' (averaging over other factors if any): ---\n"
    )
    comparisons_dose <- get_demand_comparisons(
      fit_obj = emm_model_to_use,
      compare_specs = ~dose,
      contrast_type = "pairwise",
      adjust = "fdr",
      report_ratios = TRUE
    )
    print(comparisons_dose)
  }

  if (all(c("drug", "dose") %in% factors_present)) {
    cat(
      "\n--- Pairwise comparisons for 'drug' within each level of 'dose': ---\n"
    )
    # EMMs calculated over drug*dose, then contrast drug within each dose
    comparisons_drug_by_dose <- get_demand_comparisons(
      fit_obj = emm_model_to_use,
      compare_specs = ~ drug * dose,
      contrast_type = "pairwise",
      contrast_by = "dose", # Compare 'drug' levels, holding 'dose' constant
      adjust = "fdr",
      report_ratios = TRUE
    )
    print(comparisons_drug_by_dose)
  }
} else {
  cat(
    "No suitable model with factors converged for comparisons in the vignette.\n"
  )
}
```

## Advanced Topics

**Note**: This section covers advanced modeling techniques for experienced users.
These advanced topics require deeper understanding of mixed-effects models and
may be computationally intensive.

```{r advanced_topics_note, echo=FALSE, results='asis'}
if (render_fast) {
  cat(
    "**Performance Note**: The sections below are omitted from the rendered vignette by default ",
    "(they can be slow and may fail to converge). ",
    "Set `render_fast <- FALSE` in the setup chunk to run them.\n"
  )
}
```

### More Complex Random Effects Structures

This example shows how to specify a more complex random-effects structure (e.g.,
random slopes) for $Q_{0}$ and $\alpha$.

```{r complex_random_effects_fit, eval = !render_fast, include = !render_fast}
robust_nlme_control <- nlme::nlmeControl(
  msMaxIter = 100,
  niterEM = 50,
  maxIter = 100,
  pnlsTol = 1e-4,
  tolerance = 1e-7,
  opt = "nlminb",
  msVerbose = TRUE # Set to FALSE for less output once working
)

fit_drug_slope_random <- try(
  fit_demand_mixed(
    data = ko,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey", # This is your 'monkey' identifier
    factors = c("drug", "dose"),
    factor_interaction = FALSE, # Additive fixed effects as requested
    equation_form = "zben",
    random_effects = Q0 + alpha ~ drug, # Random slopes by drug
    covariance_structure = "pdDiag",
    nlme_control = robust_nlme_control,
    start_value_method = "pooled_nls" # Good for more complex models
  ),
  silent = TRUE
)

if (
  !inherits(fit_drug_slope_random, "try-error") &&
    !is.null(fit_drug_slope_random$model)
) {
  print(fit_drug_slope_random)

  cat("\nFixed Effects:\n")
  print(nlme::fixef(fit_drug_slope_random))

  cat("\nRandom Effects Structure (Standard Deviations and Correlations):\n")
  print(nlme::VarCorr(fit_drug_slope_random$model))

  cat(
    "\nRandom Effect Estimates for each Monkey (deviations from fixed effects):\n"
  )
  print(utils::head(nlme::ranef(fit_drug_slope_random)))

  cat("\nSubject-Specific Coefficients (Fixed + Random, log10 scale):\n")
  print(head(coef(fit_drug_slope_random))) # type="combined" is default

  all_params_emms <- get_demand_param_emms(
    fit_obj = fit_drug_slope_random,
    factors_in_emm = c("drug", "dose"), # Get all cells
    at = list(monkey = "A"), # Condition on monkey A
    include_ev = TRUE # Or FALSE if not needed
  )

  observed_params_emms <- get_observed_demand_param_emms(
    fit_obj = fit_drug_slope_random,
    factors_in_emm = c("drug", "dose"), # Get all cells
    at = list(monkey = "A"), # Condition on monkey A
    include_ev = TRUE # Or FALSE if not needed
  )
} else {
  cat("Random-slope model failed to converge with the current settings.\n")
  fit_drug_slope_random <- NULL
}
```

Now let's demonstrate how to extract individual-level predicted coefficients using the
`get_individual_coefficients()` function. This combines fixed and random effects to
calculate individual-level parameter estimates for each subject.

```{r complex_random_effects_coefs, eval = !render_fast, include = !render_fast}
if (!is.null(fit_drug_slope_random) && !is.null(fit_drug_slope_random$model)) {
  # Get individual coefficients in wide format (default)
  individual_coefs_wide <- get_individual_coefficients(
    fit_drug_slope_random,
    params = c("Q0", "alpha"),
    format = "wide"
  )

  cat("Individual coefficients (wide format, log10 scale):\n")
  print(individual_coefs_wide)

  # Get individual coefficients in long format
  individual_coefs_long <- get_individual_coefficients(
    fit_drug_slope_random,
    params = c("Q0", "alpha"),
    format = "long"
  )

  cat("\nIndividual coefficients (long format, log10 scale):\n")
  print(head(individual_coefs_long, 10))

  # Convert to natural scale for interpretation
  individual_coefs_natural <- individual_coefs_wide

  # Convert Q0 coefficients to natural scale
  q0_cols <- grep(
    "^estimated_Q0_",
    names(individual_coefs_natural),
    value = TRUE
  )
  for (col in q0_cols) {
    natural_col <- gsub("estimated_Q0_", "Q0_natural_", col)
    individual_coefs_natural[[natural_col]] <- 10^individual_coefs_natural[[
      col
    ]]
  }

  # Convert alpha coefficients to natural scale
  alpha_cols <- grep(
    "^estimated_alpha_",
    names(individual_coefs_natural),
    value = TRUE
  )
  for (col in alpha_cols) {
    natural_col <- gsub("estimated_alpha_", "alpha_natural_", col)
    individual_coefs_natural[[natural_col]] <- 10^individual_coefs_natural[[
      col
    ]]
  }

  cat("\nExample individual coefficients on natural scale:\n")
  print(individual_coefs_natural[
    1:3,
    c(
      "id",
      grep("Q0_natural_", names(individual_coefs_natural), value = TRUE)[1:2],
      grep("alpha_natural_", names(individual_coefs_natural), value = TRUE)[1:2]
    )
  ])

  # You can also get coefficients for specific parameters only
  q0_only <- get_individual_coefficients(
    fit_drug_slope_random,
    params = "Q0",
    format = "wide"
  )

  cat("\nQ0 coefficients only:\n")
  print(q0_only)
} else {
  cat("Random-slope model not available for coefficient extraction.\n")
}
```

### Continuous Covariates and `fixed_rhs`

In addition to factor predictors, you can include continuous covariates in the fixed-effects
linear models for `Q0` and `alpha`.

There are two convenient ways to do this:

1. Add continuous covariates additively using the `continuous_covariates` argument (no formula writing required).
2. Specify a full right-hand-side using `fixed_rhs` (formula), which gives you complete control,
   including interactions (e.g., factor-by-continuous).

Below we demonstrate both, using a made-up continuous covariate `age` assigned per subject (`monkey`).

```{r continuous_setup, eval = !render_fast, include = !render_fast}
set.seed(123)

# Create a per-monkey continuous covariate: age
monkey_age <- unique(ko[, c("monkey")]) |>
  dplyr::mutate(age = round(runif(dplyr::n(), min = 18, max = 25), 1))

ko_age <- dplyr::left_join(ko, monkey_age, by = "monkey")
# Ensure a numeric version of dose is available for continuous modeling
ko_age <- ko_age |>
  dplyr::mutate(dose_num = suppressWarnings(as.numeric(as.character(dose))))

# Robust median for dose_num (fallbacks if all values are NA)
median_dose_num <- suppressWarnings(stats::median(
  ko_age$dose_num,
  na.rm = TRUE
))
if (is.na(median_dose_num) || !is.finite(median_dose_num)) {
  finite_vals <- ko_age$dose_num[is.finite(ko_age$dose_num)]
  if (length(finite_vals) > 0) {
    median_dose_num <- finite_vals[1]
  } else {
    median_dose_num <- 0
  }
}

# Sanity check
head(ko_age[, c("monkey", "age")])
```

#### A) Additive continuous covariate via `continuous_covariates`

Here we keep a single factor (`dose`) and add `age` as an additive continuous covariate. The RHS becomes
`~ 1 + dose + age` for both parameters.

```{r continuous_age_add, eval = !render_fast, include = !render_fast}
fit_age_add <- try(
  fit_demand_mixed(
    data = ko_age,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = "dose",
    continuous_covariates = c("age"),
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_age_add, "try-error") && !is.null(fit_age_add$model)) {
  print(fit_age_add)

  # EMMs over dose at age = 21
  emms_age_add <- get_demand_param_emms(
    fit_obj = fit_age_add,
    factors_in_emm = c("dose"),
    at = list(age = 21),
    include_ev = TRUE
  )
  cat("\nEMMs (age = 21):\n")
  print(emms_age_add)

  # Pairwise comparisons for dose at age = 21 (log10 scale; ratios reported too)
  comps_age_add <- get_demand_comparisons(
    fit_obj = fit_age_add,
    compare_specs = ~dose,
    contrast_type = "pairwise",
    at = list(age = 21),
    adjust = "fdr",
    report_ratios = TRUE
  )
  cat("\nPairwise comparisons for dose (age = 21):\n")
  print(comps_age_add)

  # Plot population lines at age = 21 (natural scale on y)
  plot(
    fit_age_add,
    at = list(age = 21),
    inv_fun = ll4_inv,
    color_by = "dose",
    x_trans = "pseudo_log",
    y_trans = "pseudo_log",
    title = "Demand by Dose with Age as Covariate (Population Fit at age = 21)"
  )
} else {
  cat(
    "\nAdditive continuous covariate model failed to converge for the vignette settings.\n"
  )
}
```

#### B) fixed_rhs with a factor (drug) and a continuous covariate (dose)

Here we specify `drug` as a factor but treat `dose` as a continuous covariate
(`dose_num`, a numeric version of `dose`),
and include `age` as an additional continuous covariate. This uses a shared RHS
for `Q0` and `alpha`:
`~ 1 + drug + dose_num + age`.

```{r continuous_age_rhs, eval = !render_fast, include = !render_fast}
fit_age_rhs <- try(
  fit_demand_mixed(
    data = ko_age,
    y_var = "y_ll4",
    x_var = "x",
    id_var = "monkey",
    factors = c("drug"), # only 'drug' is treated as a factor
    fixed_rhs = "~ 1 + drug + dose_num + age",
    equation_form = "zben",
    nlme_control = quick_nlme_control,
    start_value_method = "pooled_nls"
  ),
  silent = TRUE
)

if (!inherits(fit_age_rhs, "try-error") && !is.null(fit_age_rhs$model)) {
  print(fit_age_rhs)

  # EMMs over drug (factor) at age = 21 and dose set to its median
  emms_age_rhs <- get_demand_param_emms(
    fit_obj = fit_age_rhs,
    factors_in_emm = c("drug"),
    at = list(age = 21, dose_num = median_dose_num),
    include_ev = TRUE
  )
  cat("\nEMMs (drug; age = 21, dose_num = median):\n")
  print(emms_age_rhs)

  # Pairwise comparisons: drug at age = 21 and dose set to its median
  comps_drug_by_dose_age <- get_demand_comparisons(
    fit_obj = fit_age_rhs,
    compare_specs = ~drug,
    contrast_type = "pairwise",
    at = list(age = 21, dose_num = median_dose_num),
    adjust = "fdr",
    report_ratios = TRUE
  )
  cat("\nPairwise comparisons of drug (age = 21, dose_num = median):\n")
  print(comps_drug_by_dose_age)

  # Plot population lines at age = 21 and dose_num = median, color by drug and facet by drug
  plot(
    fit_age_rhs,
    at = list(age = 21, dose_num = median_dose_num),
    inv_fun = ll4_inv,
    color_by = "drug",
    facet_formula = ~drug,
    x_trans = "pseudo_log",
    y_trans = "pseudo_log",
    title = "Demand by Drug with Dose Continuous (Population Fit at age = 21; dose_num = median)"
  )
} else {
  cat(
    "\nfixed_rhs model (drug factor, dose continuous) failed to converge for the vignette settings.\n"
  )
}
```

### Trends with `emtrends`

We can examine how the parameters change with continuous covariates using
`get_demand_param_trends()`, which wraps `emmeans::emtrends()` and returns
tidy results for Q0 and alpha trends on the log10 scale.

Below we compute trends with respect to `age` and `dose_num`, first overall and
then by `drug` (if the model includes `drug` as a factor).

```{r trends, eval = !render_fast, include = !render_fast}
# Choose a fitted model for trends (prefer fixed_rhs fit)
fit_for_trends <- NULL
if (
  exists("fit_age_rhs") && !is.null(fit_age_rhs) && !is.null(fit_age_rhs$model)
) {
  fit_for_trends <- fit_age_rhs
} else if (
  exists("fit_age_add") && !is.null(fit_age_add) && !is.null(fit_age_add$model)
) {
  fit_for_trends <- fit_age_add
}

if (!is.null(fit_for_trends)) {
  # Determine which of the candidate covariates are actually present in the model RHS
  # Use Q0 formula (or alpha if Q0 is NULL) - they may differ if collapse_levels was used
  rhs_formula_str <- fit_for_trends$formula_details$fixed_effects_formula_str_Q0
  if (
    is.null(rhs_formula_str) ||
      is.na(rhs_formula_str) ||
      !nzchar(rhs_formula_str)
  ) {
    rhs_formula_str <- fit_for_trends$formula_details$fixed_effects_formula_str_alpha
  }
  rhs_vars_tr <- tryCatch(
    all.vars(as.formula(rhs_formula_str)),
    error = function(e) character(0)
  )
  covariates_candidate <- c("age", "dose_num")
  covariates_in_model <- intersect(rhs_vars_tr, covariates_candidate)

  if (length(covariates_in_model) > 0) {
    cat("\n--- Trends (overall) at age = 21 and dose_num = median ---\n")
    trends_overall <- get_demand_param_trends(
      fit_obj = fit_for_trends,
      params = c("Q0", "alpha"),
      covariates = covariates_in_model,
      specs = ~1,
      at = list(age = 21, dose_num = median_dose_num)
    )
    print(trends_overall)

    # If 'drug' is a model factor, show trends by drug as well
    model_factors_tr <- fit_for_trends$param_info$factors
    if (!is.null(model_factors_tr) && ("drug" %in% model_factors_tr)) {
      cat("\n--- Trends by drug at age = 21 and dose_num = median ---\n")
      trends_by_drug <- get_demand_param_trends(
        fit_obj = fit_for_trends,
        params = c("Q0", "alpha"),
        covariates = covariates_in_model,
        specs = ~drug,
        at = list(age = 21, dose_num = median_dose_num)
      )
      print(trends_by_drug)
    } else {
      cat(
        "\nNote: The selected model does not include 'drug' as a factor; skipping by-drug trends.\n"
      )
    }
  } else {
    cat(
      "\nNo candidate continuous covariates ('age', 'dose_num') are present in the model RHS; skipping trends.\n"
    )
  }
} else {
  cat(
    "\nNo suitable fitted model available for computing trends in the vignette.\n"
  )
}
```

# See Also

- `vignette("mixed-demand")` -- Basic mixed-effects demand modeling
- `vignette("model-selection")` -- Choosing the right model class
- `vignette("hurdle-demand-models")` -- Two-part hurdle demand models
- `vignette("beezdemand")` -- Getting started with beezdemand
