---
title: "Power Analysis for Within-Subject Demand Designs"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power Analysis for Within-Subject Demand Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(beezdemand)
```

# Why simulation-based power?

Before running a study, a researcher planning a demand experiment needs an
answer to: *"How many subjects do I need to detect an effect of this size,
with this design, at 80% power?"* Closed-form power formulas do not exist for
nonlinear mixed-effects demand models, and Wald asymptotics are unreliable at
realistic study sizes, especially near boundary parameters. `power_demand()`
therefore estimates power by Monte Carlo simulation:

1. Simulate a dataset under assumed true population parameters, a
   hypothesized effect size, and the study design.
2. Refit it with the package's TMB mixed-effects fitter,
   [fit_demand_tmb()].
3. Extract the Wald standard error of the target condition contrast from the
   TMB `sdreport`.
4. Record both a p-value verdict (Wald test against `alpha`) and a
   confidence-interval verdict (does the Wald CI exclude 0). Both are
   referred to a t distribution with `n_subjects - 1` degrees of freedom
   rather than the asymptotic normal: the calibration battery showed the
   z-test is anticonservative at study-relevant N (empirical Type I error
   0.089 at nominal .05 with 15 subjects), while the null Wald statistics
   closely follow the t with the design's paired-contrast df. The two
   verdicts use the same standard error and reference distribution, so they
   always agree -- both report formats are returned. `df = Inf` recovers the
   asymptotic z-test if you need it.
5. Repeat `n_sim` times. Power is the proportion of *usable* fits
   (converged, positive-definite Hessian, finite SE) that reject, reported
   with a Wilson confidence interval because it is itself an estimate from
   finitely many replicates.

The data-generating model is the package's existing within-subject demand
simulator: every subject is observed at every price under both levels of a
within-subject condition, with per-(subject, condition) random effects on
log Q0 and log alpha and lognormal residual error. The effect of interest is
a shift on natural-log Q0 (`delta_q0`) *or* natural-log alpha
(`delta_alpha`) in condition 2 relative to condition 1. For example,
`delta_q0 = log(1.5)` means condition 2's intensity is 1.5 times
condition 1's.

# Estimating power at a fixed sample size

Suppose we expect a 50% increase in Q0 (`delta_q0 = log(1.5) ≈ 0.41`) and
plan 20 subjects. The `design` list sets the assumed population values and
price grid (defaults shown in `?power_demand`):

```{r power-fixed-n}
res <- power_demand(
  n_subjects = 20,
  effect = list(delta_q0 = log(1.5)),
  design = list(
    prices = c(0.1, 0.5, 1, 2, 5, 10),
    sigma_b = 0.3,  # subject SD on log Q0, per condition
    sigma_d = 0.3,  # subject SD on log alpha, per condition
    sigma_e = 0.1   # residual SD on log consumption
  ),
  n_sim = 40,       # small for a fast vignette; see guidance below
  seed = 1234,
  verbose = FALSE
)
res
```

The Monte Carlo confidence interval (`power_mc_ci`) is wide at `n_sim = 40`
-- that is the point of reporting it. Interactive exploration is fine at a
few hundred replicates; numbers destined for a grant application should use
`n_sim = 2000` or more (see the guidance section below).

Convergence diagnostics are part of the output, not a footnote: replicates
whose refit fails are excluded from the power denominator and counted in
`n_converged` / `n_used`, never treated as "no effect detected".

```{r diagnostics}
res$n_converged
res$n_used
table(res$replicates$status)
```

# Searching for a sample size

`find_n_demand()` wraps the engine in a bisection search over `n_subjects`
for the smallest N whose estimated power reaches the target. The search is
aware of Monte Carlo noise: at each candidate N it adds replicates until the
Wilson interval falls clearly above or below the target (up to `n_sim_max`),
and it re-evaluates the selected N and its lower neighbor before reporting.

```{r find-n}
search <- find_n_demand(
  target_power = 0.8,
  effect = list(delta_q0 = log(2)),
  design = list(prices = c(0.1, 0.5, 1, 2, 5, 10)),
  n_range = c(4, 30),
  n_sim = 30,       # small for a fast vignette
  seed = 5678,
  verbose = FALSE
)
search
```

The result is an *estimated minimum under Monte Carlo uncertainty* (the
`status` field says whether the confirmation pass was conclusive). For a
defensible grant number, rerun `power_demand()` at the returned N with a
large `n_sim` and report that estimate with its Monte Carlo interval.

# Sensitivity to the assumed effect size

Power claims are conditional on the assumed effect and variance components.
A small sweep makes that dependence visible:

```{r sensitivity}
deltas <- c(log(1.25), log(1.5), log(2))
sweep <- vapply(deltas, function(d) {
  power_demand(
    n_subjects = 20, effect = list(delta_q0 = d),
    design = list(prices = c(0.1, 0.5, 1, 2, 5, 10)),
    n_sim = 30, seed = 42, verbose = FALSE
  )$power
}, numeric(1))
data.frame(q0_ratio = exp(deltas), power = sweep)
```

# How many replicates do you need?

The Monte Carlo standard error of a power estimate near `p` is
`sqrt(p (1 - p) / n_sim)`:

| `n_sim` | MC SE near power = 0.8 | Use |
|--------:|-----------------------:|-----|
| 100     | 0.040 | quick interactive exploration |
| 500     | 0.018 | default; serious exploration |
| 2000    | 0.009 | grant applications, preregistrations |

Always report the Wilson interval (`power_mc_ci`) alongside the point
estimate; a power of "0.80" from 100 replicates is `[0.71, 0.87]`.

# Validity and Limitations

**Validity checks.** The package test suite
(`tests/testthat/test-power-demand.R`) verifies, with preregistered seeds
and tolerance bands fixed before the tests were first run:

- **Type I error calibration** (the load-bearing check): with the effect set
  to zero, the empirical false-positive rate at nominal `alpha = .05` over
  1,200 replicates must fall in [0.03, 0.07] -- a band of 3.18 binomial
  standard errors that excludes both half and 1.5 times the nominal rate.
  `n_sim = 1200` was computed from that tolerance (`9 * .05 * .95 / .02^2 ≈
  1069`), not guessed. A second null check runs at a realistic N = 40 on the
  alpha contrast. This check is what motivated the t reference distribution:
  the first run, using the asymptotic z-test, failed it (empirical rate
  0.089 at n = 15), and the t(n - 1) reference was adopted and re-validated
  rather than widening the band.
- **Convergence handling**: a configuration that reliably produces
  non-convergence confirms failed replicates are excluded from the power
  denominator and surfaced via `n_converged` / `n_used`, not silently
  counted as misses.
- **Closed-form benchmark**: in a degenerate configuration (tiny residual
  and alpha-side variability) the design reduces to a paired comparison of
  per-condition log-Q0 values, and the Monte Carlo estimate must match the
  analytic one-sample power from `pwr::pwr.t.test()` within a preregistered
  tolerance of 0.10.
- **Monotonicity**: power increases with `n_subjects` and effect size and
  decreases with random-effect SD, within Monte Carlo slack.
- **Reproducibility**: identical `seed` gives identical results, asserted
  exactly.

These checks validate the default configuration: `equation = "simplified"`
with independent per-condition random effects (`rho_bd = 0`). Note that the
refit is a *working model*: the simulator draws multiplicative lognormal
errors while the simplified equation fits additive Gaussian errors on raw
consumption (they share the mean function; the error models differ). The
approximation is closest at small residual SD, and the null calibration is
verified at both the default `sigma_e = 0.1` and a stress value of 0.3.
Other equations remain available as sensitivity analyses, but their
condition contrasts are defined on different parameterizations and are not
validated against the simulated delta; `rho_bd` is locked at 0 in this
version.

The `find_n_*` search statuses (`"confirmed"`, `"uncertain"`,
`"unresolved"`) describe a heuristic Monte Carlo decision rule -- repeated
looks at ordinary Wilson intervals across candidate N -- not a formal
sequential testing procedure with a guaranteed error rate. When the
confirmation pass contradicts the search, the function returns `n = NA`
rather than an unsupported number.

**Explicitly out of scope in v1** (flagged as future work, not silently
approximated):

- Simultaneous multi-parameter effects (e.g., a joint Q0 + alpha shift).
- Power for derived measures (Pmax, Omax, elasticity at a point,
  breakpoint).
- Arbitrary user-supplied designs beyond the package's within-subject
  simulator (two conditions, all subjects at all prices in both conditions;
  between-subject designs are the province of
  `beezdiscounting::power_discounting()`'s design or future work here).
- Any graphical or interactive interface.

**What the numbers can bear.** A reported power estimate is (a) conditional
on the assumed population parameters and variance components -- vary them and
look at the sensitivity of the answer; (b) conditional on usable fits -- take
the `n_used` warning seriously if it fires; and (c) a Monte Carlo estimate --
cite it with its interval at an `n_sim` sized for the decision it supports.
