---
title: "TMB Mixed-Effects Demand Models"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TMB Mixed-Effects Demand Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "tmb-mixed-effects_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)

data("apt_full", package = "beezdemand", envir = environment())

# Set `BEEZDEMAND_VIGNETTE_MODE=full` to run the slower examples.
render_fast <- tolower(Sys.getenv("BEEZDEMAND_VIGNETTE_MODE", "fast")) != "full"

cache_key_object <- function(x) {
  tmp <- tempfile(fileext = ".rds")
  saveRDS(x, tmp)
  on.exit(unlink(tmp), add = TRUE)
  unname(tools::md5sum(tmp))
}

knitr::opts_chunk$set(
  cache.extra = list(
    beezdemand_version = as.character(utils::packageVersion("beezdemand")),
    apt_full = cache_key_object(apt_full)
  )
)
```

```{r data-setup, include = FALSE}
# Consistent subset used throughout the vignette
set.seed(42)
ids_sub <- sample(unique(apt_full$id), 100)
dat <- apt_full[apt_full$id %in% ids_sub, ]
```

## Introduction

`fit_demand_tmb()` fits continuous mixed-effects demand models using
[Template Model Builder](https://github.com/kaskr/adcomp) (TMB). It is the
modern alternative to `fit_demand_mixed()` (which uses `nlme`) and provides
several advantages:

- **Automatic differentiation** -- exact gradients via compiled C++, replacing
  the numerical finite-difference approximations used by `nlme`
- **Laplace approximation** -- integrates over random effects analytically
  rather than relying on iterative linearization
- **Multi-start optimization** -- automatically tries multiple starting-value
  sets and keeps the best fit
- **Four equation forms** -- exponential, exponentiated, simplified, and
  zero-bounded exponential (zben)
- **Factor and covariate support** -- design matrices for group comparisons
  with estimated marginal means (EMMs) and pairwise contrasts

This vignette covers `fit_demand_tmb()` for continuous consumption data. For
**two-part hurdle models** that explicitly model zero consumption, see
`vignette("hurdle-demand-models")`. For **individual NLS curves**, see
`vignette("fixed-demand")`.

## Quick Start

```{r quick-fit}
fit <- fit_demand_tmb(
  dat,
  y_var = "y", x_var = "x", id_var = "id",
  equation = "exponential",
  random_effects = c("q0", "alpha"),
  verbose = 0
)
fit
```

```{r quick-summary}
summary(fit)
```

```{r quick-plot, fig.cap = "Population demand curve from the exponential equation."}
plot(fit, type = "demand")
```

The `exponential` equation (Hursh & Silberberg, 2008) models log-transformed
consumption and is the most reliable choice for 2-random-effect models. It
automatically drops zero-consumption observations.

## Choosing an Equation

`fit_demand_tmb()` supports four demand equations:

| Equation | Response | Zeros | k | Best for |
|----------|----------|-------|---|----------|
| `"exponential"` | log(Q) | Dropped | Estimated or fixed | Most datasets; robust 2-RE convergence |
| `"exponentiated"` | Raw Q | Allowed | Estimated or fixed | Data with few zeros; 1-RE models |
| `"simplified"` | Raw Q | Allowed | None | Simpler model without k |
| `"zben"` | LL4(Q) | Allowed (via transform) | None | Wide dynamic range with LL4 compression |

### Mathematical Specifications

**Exponential** (Hursh & Silberberg, 2008):
$$\log(Q_{ij}) = \log(Q_{0i}) + k \left(e^{-\alpha_i \cdot Q_{0i} \cdot C_j} - 1\right) + \varepsilon_{ij}$$

**Exponentiated** (Koffarnus et al., 2015):
$$Q_{ij} = Q_{0i} \cdot 10^{k \left(e^{-\alpha_i \cdot Q_{0i} \cdot C_j} - 1\right)} + \varepsilon_{ij}$$

**Simplified**:
$$Q_{ij} = Q_{0i} \cdot e^{-\alpha_i \cdot Q_{0i} \cdot C_j} + \varepsilon_{ij}$$

**Zero-bounded exponential (zben)** (see `src/MixedDemand.h` for full
specification):
$$\log_{10}(Q_{0i}) \cdot e^{-\frac{\alpha_i}{\log_{10}(Q_{0i})} \cdot Q_{0i} \cdot C_j} + \varepsilon_{ij}$$

where $Q_{0i} = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}_{Q_0} + b_i)$,
$\alpha_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}_\alpha + c_i)$, and
$(b_i, c_i)$ are correlated random effects.

```{r eq-exponential}
fit_exp <- fit_demand_tmb(dat, equation = "exponential",
                          random_effects = c("q0", "alpha"), verbose = 0)
```

```{r eq-exponentiated}
# exponentiated converges reliably with 1-RE
fit_expon <- fit_demand_tmb(dat, equation = "exponentiated",
                            random_effects = "q0", verbose = 0)
```

```{r eq-simplified}
fit_simp <- fit_demand_tmb(dat, equation = "simplified",
                           random_effects = "q0", verbose = 0)
```

```{r eq-zben}
dat$y_ll4 <- ll4(dat$y)
fit_zben <- fit_demand_tmb(dat, y_var = "y_ll4", equation = "zben",
                           random_effects = "q0", verbose = 0)
```

```{r eq-compare}
data.frame(
  equation = c("exponential", "exponentiated", "simplified", "zben"),
  random_effects = c("2-RE", "1-RE", "1-RE", "1-RE"),
  converged = c(fit_exp$converged, fit_expon$converged,
                fit_simp$converged, fit_zben$converged),
  AIC = round(c(AIC(fit_exp), AIC(fit_expon),
                AIC(fit_simp), AIC(fit_zben)), 1)
)
```

**Note:** AIC values are not directly comparable across equations because they
model different response scales (log Q vs raw Q vs LL4(Q)).

**Convergence tip:** The `exponential` equation is the most reliable for
2-random-effect models. The `exponentiated` equation works well with 1-RE but
can struggle to converge with 2-RE, especially with smaller samples.

## Random Effects Structure

`fit_demand_tmb()` supports two configurations:

- `random_effects = "q0"` -- **1-RE**: random intercept on $Q_0$ only;
  $\alpha$ is constant across subjects
- `random_effects = c("q0", "alpha")` -- **2-RE**: random effects on both
  $Q_0$ and $\alpha$ with an estimated correlation

```{r re-compare}
fit_1re <- fit_demand_tmb(dat, equation = "exponential",
                          random_effects = "q0", verbose = 0)
fit_2re <- fit  # reuse from Quick Start (exponential, 2-RE)
```

```{r re-compare-output}
compare_models(fit_1re, fit_2re)
```

A significant LRT p-value and substantially lower AIC/BIC for the 2-RE model
indicate that subjects differ meaningfully in both intensity ($Q_0$) and
elasticity ($\alpha$).

```{r re-ranef}
head(nlme::ranef(fit_2re))
```

The `b_i` column is the random deviation on log($Q_0$) and `c_i` is the
random deviation on log($\alpha$) for each subject.

## Estimating vs. Fixing k

For the `exponential` and `exponentiated` equations, $k$ scales the range of
the demand curve. By default it is estimated (`estimate_k = TRUE`). You can
fix it at the conventional value of 2 (Hursh & Silberberg, 2008):

```{r k-compare}
fit_k_free <- fit_demand_tmb(dat, equation = "exponential",
                              random_effects = c("q0", "alpha"),
                              estimate_k = TRUE, verbose = 0)
fit_k_fixed <- fit_demand_tmb(dat, equation = "exponential",
                               random_effects = c("q0", "alpha"),
                               estimate_k = FALSE, k = 2, verbose = 0)

data.frame(
  k = c("estimated", "fixed at 2"),
  converged = c(fit_k_free$converged, fit_k_fixed$converged),
  AIC = round(c(AIC(fit_k_free), AIC(fit_k_fixed)), 1),
  k_value = round(c(exp(coef(fit_k_free)[["log_k"]]), 2), 3)
)
```

The `simplified` and `zben` equations do not use a $k$ parameter.

## Examining Results

### Summary and Coefficients

```{r results-summary}
summary(fit_2re)
```

```{r results-tidy}
tidy(fit_2re)
```

```{r results-glance}
glance(fit_2re)
```

### Subject-Specific Parameters

Each subject gets empirical Bayes estimates of $Q_0$, $\alpha$, $P_{max}$, and
$O_{max}$:

```{r results-subject-pars}
spars <- get_subject_pars(fit_2re)
head(spars)
```

```{r results-subject-summary}
spars |>
  summarise(
    across(c(Q0, alpha, Pmax, Omax),
           list(median = median, mean = mean, sd = sd),
           .names = "{.col}_{.fn}")
  ) |>
  tidyr::pivot_longer(everything(), names_to = c("parameter", "stat"),
                      names_sep = "_") |>
  tidyr::pivot_wider(names_from = stat, values_from = value) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

### Confidence Intervals

```{r results-confint}
confint(fit_2re)
```

By default, confidence intervals are on the internal (log) scale. Use
`report_space = "natural"` for back-transformed intervals:

```{r results-confint-natural}
confint(fit_2re, report_space = "natural")
```

### Predictions

Three prediction types are available:

```{r predict-response}
# Fitted values for observed data
pred_resp <- predict(fit_2re, type = "response")
head(pred_resp)
```

```{r predict-demand}
# Population demand curve at specific prices
predict(fit_2re, type = "demand", prices = c(0, 0.5, 1, 2, 5, 10, 20))
```

```{r predict-parameters}
# Subject-level parameter estimates (same as get_subject_pars)
pred_pars <- predict(fit_2re, type = "parameters")
head(pred_pars)
```

### Population Metrics

```{r results-group-metrics}
calc_group_metrics(fit_2re)
```

### Visualization

```{r plot-demand, fig.cap = "Population-level demand curve with confidence band."}
plot(fit_2re, type = "demand")
```

```{r plot-individual, fig.cap = "Individual demand curves for a random sample of subjects."}
plot(fit_2re, type = "individual")
```

```{r plot-parameters, fig.cap = "Distribution of subject-specific demand parameters."}
plot(fit_2re, type = "parameters")
```

## Diagnostics

After fitting, assess model health with the built-in diagnostic tools.

```{r diagnostics-check}
# Model health check
check_demand_model(fit_2re)
```

```{r diagnostics-augment}
# Augment with fitted values and residuals
aug <- augment(fit_2re)
head(aug[, c("id", "x", "y", ".fitted", ".resid", ".std_resid")])
```

```{r diagnostics-qq, fig.cap = "Q-Q plots of subject-level random effects."}
# Q-Q plot of random effects
plot_qq(fit_2re)
```

```{r diagnostics-re, fig.cap = "Random effects diagnostic panels."}
# Random effects diagnostic panels
plot_re_diagnostics(fit_2re)
```

These diagnostics help identify:

- **Q-Q plots**: Non-normality of random effects (heavy tails, outliers)
- **`check_demand_model()`**: Convergence issues, boundary estimates, residual patterns
- **Augmented data**: Observation-level residuals for identifying poorly fitting subjects

## Group Comparisons

To test whether demand parameters differ by group, pass factor variables via
the `factors` argument. This adds fixed effects to the design matrices for both
$Q_0$ and $\alpha$.

```{r factor-setup}
# Filter to Male/Female for a clean two-level comparison
dat_mf <- dat |> filter(gender %in% c("Male", "Female"))

fit_gender <- fit_demand_tmb(
  dat_mf,
  y_var = "y", x_var = "x", id_var = "id",
  equation = "exponential",
  factors = "gender",
  random_effects = c("q0", "alpha"),
  verbose = 0
)
fit_gender
```

### Estimated Marginal Means

```{r emms-q0}
get_demand_param_emms(fit_gender, param = "Q0")
```

```{r emms-alpha}
get_demand_param_emms(fit_gender, param = "alpha")
```

EMMs are reported on both the natural scale (`estimate`) and log scale
(`estimate_log`). The natural-scale estimates represent the population-average
$Q_0$ or $\alpha$ for each group.

### Pairwise Comparisons

```{r comparisons-q0}
get_demand_comparisons(fit_gender, param = "Q0")
```

```{r comparisons-alpha}
get_demand_comparisons(fit_gender, param = "alpha")
```

The `estimate_ratio` column gives the multiplicative ratio between groups on
the natural scale (e.g., a ratio of 1.34 means Group A's $Q_0$ is 34% higher
than Group B's).

```{r factor-interaction, eval = !render_fast, include = !render_fast}
# Factor interaction: gender x age group
dat_mf$age_group <- ifelse(
  dat_mf$age < median(dat_mf$age, na.rm = TRUE), "younger", "older"
)

fit_interact <- fit_demand_tmb(
  dat_mf, y_var = "y", x_var = "x", id_var = "id",
  equation = "exponential",
  factors = c("gender", "age_group"),
  factor_interaction = TRUE,
  random_effects = c("q0", "alpha"),
  verbose = 0
)
get_demand_param_emms(fit_interact, param = "Q0")
```

```{r asis-interaction-note, eval = render_fast, results = 'asis', echo = FALSE}
cat("*Factor interaction and continuous covariate sections are skipped in fast",
    "render mode. Set `BEEZDEMAND_VIGNETTE_MODE=full` to include them.*\n")
```

```{r continuous-covariates, eval = !render_fast, include = !render_fast}
# Continuous covariates enter the design matrices for Q0 and alpha
fit_cov <- fit_demand_tmb(
  dat_mf, y_var = "y", x_var = "x", id_var = "id",
  equation = "exponential",
  continuous_covariates = "age",
  random_effects = c("q0", "alpha"),
  verbose = 0
)
tidy(fit_cov)
```

```{r collapse-levels, eval = !render_fast, include = !render_fast}
# Asymmetric factor collapsing: different groupings for Q0 vs alpha
dat_mf$age_group <- cut(
  dat_mf$age,
  breaks = c(0, 25, 40, Inf),
  labels = c("young", "mid", "older")
)

fit_collapsed <- fit_demand_tmb(
  dat_mf, y_var = "y", x_var = "x", id_var = "id",
  equation = "exponential",
  factors = "age_group",
  # Collapse "young" and "mid" into "young_mid" for Q0,

  # but collapse "mid" with "older" for alpha
  collapse_levels = list(
    Q0 = list(age_group = c("young" = "young_mid", "mid" = "young_mid",
                            "older" = "older")),
    alpha = list(age_group = c("young" = "young", "mid" = "older",
                               "older" = "older"))
  ),
  random_effects = "q0",
  verbose = 0
)
tidy(fit_collapsed)
```

## Model Comparison

Use `compare_models()` to compare nested TMB models via likelihood ratio test,
AIC, and BIC:

```{r model-comparison}
compare_models(fit_1re, fit_2re)
```

**Valid comparisons** require models fit on the same data with the same
equation and response scale. Models with different equations (e.g., exponential
vs exponentiated) model different responses and cannot be compared via AIC.

## Convergence Tips

If a model fails to converge, try these strategies in order:

| Strategy | How | When |
|----------|-----|------|
| Reduce random effects | `random_effects = "q0"` | 2-RE models struggling |
| Use exponential equation | `equation = "exponential"` | Exponentiated/simplified not converging |
| Fix k | `estimate_k = FALSE, k = 2` | k estimate drifting to extreme values |
| Increase iterations | `tmb_control = list(iter_max = 2000)` | "false convergence" messages |
| Try L-BFGS-B | `tmb_control = list(optimizer = "L-BFGS-B")` | nlminb not making progress |
| Set parameter bounds | `tmb_control = list(lower = ..., upper = ...)` | Estimates at boundaries |
| Warm start | `tmb_control = list(warm_start = prev_fit$opt$par)` | Refining a near-converged fit |

```{r convergence-examples, eval = FALSE}
# Switch optimizer
fit <- fit_demand_tmb(
  dat, equation = "exponential",
  tmb_control = list(optimizer = "L-BFGS-B"),
  verbose = 0
)

# Warm-start from a previous fit
fit2 <- fit_demand_tmb(
  dat, equation = "exponential",
  tmb_control = list(warm_start = fit$opt$par),
  verbose = 0
)

# Apply parameter bounds
fit3 <- fit_demand_tmb(
  dat, equation = "exponential",
  tmb_control = list(
    lower = c(log_k = -2),
    upper = c(log_k = 4)
  ),
  verbose = 0
)

# Disable multi-start for faster iteration during exploration
fit4 <- fit_demand_tmb(
  dat, equation = "exponential",
  multi_start = FALSE,
  verbose = 2
)
```

### tmb_control Options

| Field | Default | Description |
|-------|---------|-------------|
| `optimizer` | `"nlminb"` | `"nlminb"` or `"L-BFGS-B"` |
| `iter_max` | 1000 | Maximum iterations |
| `eval_max` | 2000 | Maximum evaluations (nlminb only) |
| `rel_tol` | 1e-10 | Relative tolerance (nlminb only) |
| `lower` / `upper` | NULL | Named numeric bounds on log-scale parameters |
| `warm_start` | NULL | Starting values from a previous `fit$opt$par` |
| `trace` | 0 | Optimizer trace level |

## Choosing Between fit_demand_tmb() and fit_demand_mixed()

| Feature | `fit_demand_tmb()` | `fit_demand_mixed()` |
|---------|-------------------|---------------------|
| Backend | TMB (C++, automatic differentiation) | nlme (R, numerical gradients) |
| Equations | exponential, exponentiated, simplified, zben | zben, simplified |
| k parameter | Estimated or fixed | Not available |
| Random effects | 1 or 2 (Q0, alpha) | Configurable via nlme |
| Convergence | Robust (AD + Laplace + multi-start) | Can struggle with nonlinear equations |
| Speed | Fast (compiled C++) | Variable |
| Post-hoc EMMs | `get_demand_param_emms()` | `get_demand_param_emms()` (via emmeans) |
| Factors/covariates | Design matrices | Formula-based |

Prefer `fit_demand_tmb()` for new work. `fit_demand_mixed()` remains useful
when you need specific `nlme` features like custom correlation structures or
when working with existing pipelines.

## References

Hursh, S. R., & Silberberg, A. (2008). Economic demand and essential value.
*Psychological Review*, 115(1), 186--198.

Koffarnus, M. N., Franck, C. T., Stein, J. S., & Bickel, W. K. (2015). A
modified exponential behavioral economic demand model to better describe
consumption data. *Experimental and Clinical Psychopharmacology*, 23(6),
504--512.

Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H., & Bell, B. M. (2016).
TMB: Automatic differentiation and Laplace approximation. *Journal of
Statistical Software*, 70(5), 1--21.

## See Also

- `vignette("fixed-demand")` -- Individual NLS demand curves
- `vignette("mixed-demand")` -- NLME-based mixed-effects models
- `vignette("mixed-demand-advanced")` -- Advanced topics: multi-factor designs, collapse_levels
- `vignette("hurdle-demand-models")` -- Two-part hurdle models for zero-heavy data
- `vignette("model-selection")` -- Choosing the right demand model
