---
title: "TMB Advanced Random-Effects Structures"
author: "Brent Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TMB Advanced Random-Effects Structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dpi = 144,
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "tmb-advanced-random-effects_cache/",
  autodep = TRUE
)

library(beezdemand)
library(dplyr)
library(ggplot2)
library(nlme)

# Set `BEEZDEMAND_VIGNETTE_MODE=full` to run the slower examples.
render_fast <- tolower(Sys.getenv("BEEZDEMAND_VIGNETTE_MODE", "fast")) != "full"
```

# Overview

This vignette covers random-effects (RE) structures beyond intercepts-only
mixed-effects in `fit_demand_tmb()`. If you are new to TMB-based demand
modeling, start with `vignette("tmb-mixed-effects")` for the
intercept-only and basic 2-RE cases. This vignette assumes that
foundation and walks through:

1. **Factor-expanded REs** — random slopes on a within-subject factor
   via `pdDiag(Q0 + alpha ~ condition)` or
   `pdSymm(Q0 + alpha ~ condition)`.
2. **Multi-block `pdBlocked` REs** — independent RE blocks at the same
   grouping level. Combines a baseline block with a per-condition
   slopes block to recover within-subject heterogeneity that
   intercepts-only fits miss.
3. **Reading subject-level results** — the long-form
   `get_subject_pars(fit, expanded = TRUE)` interface, when to use it,
   and the relationship to `predict()`.
4. **Group metrics with conditioning** — the `at` argument on
   `calc_group_metrics()` for explicit conditioning on covariates and
   factor levels.
5. **Diagnostics** — variance components per block, multi-start advice,
   convergence troubleshooting.

The motivating use case for the multi-block work is a within-subject
demand study where Q₀ varies by experimental condition AND subjects
differ in their condition response. An intercepts-only fit on such data
can invert the population-level Q₀ direction; a multi-block
`pdBlocked(list(pdSymm(Q0+α~1), pdDiag(Q0+α~condition-1)))` recovers it.

# 1. Decision tree: which RE structure?

```
                        +-------------------------------+
                        | Within-subject factor in data?|
                        +-------------------------------+
                          |                       |
                         No                      Yes
                          |                       |
                          v                       v
                +--------------------+   +-----------------------------+
                | Intercepts only:   |   | Are subjects' responses    |
                | random_effects =   |   | to the factor heterogeneous?|
                | Q0 + alpha ~ 1     |   | (i.e., do you need random   |
                +--------------------+   |  slopes?)                   |
                                         +-----------------------------+
                                              |                |
                                             No               Yes
                                              |                |
                                              v                v
                                +--------------+   +--------------------------+
                                | Single block,|   | Need independence between|
                                | factor in    |   | baseline and slopes?     |
                                | factors arg  |   | (cigarette M1 case)      |
                                +--------------+   +--------------------------+
                                                       |             |
                                                      No            Yes
                                                       |             |
                                                       v             v
                                          +--------------------+  +-------------+
                                          | Single block:      |  | Multi-block |
                                          | pdSymm(Q0+α~       |  | pdBlocked   |
                                          | factor)            |  | (see below) |
                                          +--------------------+  +-------------+
```

In short:

- **Intercepts only** when the factor (if any) is between-subject and a
  single per-subject Q₀/α captures the heterogeneity.
- **Single-block factor expansion** when subjects respond differently to
  a within-subject factor, and you want the baseline-vs-slope REs to
  be correlated.
- **Multi-block `pdBlocked`** when you specifically want baseline REs
  and condition-slope REs to be uncorrelated. The classic case is a
  study where the population-level Q₀ ordering across conditions is
  what you want to recover, and an intercepts-only fit gets the
  direction wrong on within-subject heterogeneous data.

# 2. Factor-expanded REs (Phase 2)

`fit_demand_tmb()` accepts a single `pdDiag` or `pdSymm` block with a
factor on the right-hand side of the random-effects formula. The factor
is expanded to indicator columns (one per non-reference level under
treatment contrasts, or one per level with `~ factor - 1`).

```{r factor-sim, eval = !render_fast}
sim <- beezdemand:::.simulate_within_subject_demand(
  n_subjects = 50,
  n_conditions = 3,
  prices = c(0.25, 0.5, 1, 2, 4, 8, 16, 32),
  log_q0_pop = log(15),
  log_alpha_pop = log(0.0015),
  delta_q0 = c(0, -0.4, -0.9),
  delta_alpha = c(0, 0.2, 0.5),
  sigma_b = 0.4,
  sigma_d = 0.4,
  seed = 42
)
sim$id <- factor(sim$id)
sim$condition <- factor(sim$condition)

fit_factor <- fit_demand_tmb(
  sim,
  equation = "simplified",
  factors = "condition",
  random_effects = pdSymm(Q0 + alpha ~ condition),
  multi_start = FALSE,
  verbose = 0
)
summary(fit_factor)
```

Random slopes on `condition` give each subject their own deviation from
the population-level condition shift. `pdDiag` makes those deviations
independent across conditions; `pdSymm` allows them to correlate.

# 3. Multi-block `pdBlocked` (Phase 3)

The single most consequential RE structure is the two-block `pdBlocked`
spec used in our cigarette demand work:

```r
pdBlocked(list(
  pdSymm(Q0 + alpha ~ 1),                # baseline (correlated)
  pdDiag(Q0 + alpha ~ condition - 1)     # condition-specific slopes
))
```

The first block is a 2-dimensional baseline: each subject has correlated
intercepts on log-Q₀ and log-α. The second block is a 6-dimensional
slopes block (3 conditions × 2 parameters): each subject has independent
deviations on log-Q₀ and log-α at each condition. The two blocks are
independent of each other by virtue of `pdBlocked`.

Why does this matter? In data where Q₀ heterogeneity correlates with
condition response, an intercepts-only fit can invert the
population-level Q₀ direction across conditions. The multi-block spec
absorbs the within-subject heterogeneity into the slopes block, freeing
the fixed effects to recover the correct direction.

```{r m1-fit, eval = !render_fast}
fit_m1 <- fit_demand_tmb(
  sim,
  equation = "simplified",
  factors = "condition",
  random_effects = pdBlocked(list(
    pdSymm(Q0 + alpha ~ 1),
    pdDiag(Q0 + alpha ~ condition - 1)
  )),
  multi_start = FALSE,
  verbose = 0
)
emms_q0 <- get_demand_param_emms(fit_m1, param = "Q0")
emms_q0
```

The simulator was configured with `delta_q0 = c(0, -0.4, -0.9)`, so the
population Q₀ ordering should be C1 > C2 > C3. The M1 fit recovers it.

# 4. Reading subject-level results (Phase 5A)

For fits with within-subject random-effects design (factor-expanded
single block, multi-block `pdBlocked`, or numeric within-id RE-RHS
terms), the wide one-row-per-subject `subject_pars` returns `NA` in
`Q0`, `alpha`, `Pmax`, and `Omax` because no single subject-level value
is well-defined.

```{r expanded-default, eval = !render_fast}
spars_wide <- get_subject_pars(fit_m1)
head(spars_wide)
```

To get per-(subject, factor-level) values, use `expanded = TRUE`:

```{r expanded-true, eval = !render_fast}
spars_long <- get_subject_pars(fit_m1, expanded = TRUE)
head(spars_long)
```

The long-form table has one row per (subject, condition) combination.
`b_i` and `c_i` repeat across conditions for the same subject (they are
subject-level RE intercepts); `Q0`, `alpha`, `Pmax`, and `Omax` differ
per row because they incorporate the per-condition slope REs.

Numeric within-id-varying covariates (e.g., trial number) are
conditioned at the subject's mean rather than expanded over their
values — only factor levels drive row expansion.

For power users who want the raw per-block RE matrices:

```{r re-mats, eval = !render_fast}
re_q0_mat <- attr(spars_wide, "re_q0_mat")
re_alpha_mat <- attr(spars_wide, "re_alpha_mat")
str(re_q0_mat)
```

These are `n_subjects × re_dim` matrices ordered by block. The first
column is the baseline intercept RE; subsequent columns are the
condition slopes.

## Plot and amplitude/persistence guards

Two consumers — `plot(fit, type = "individual")` and
`calculate_amplitude_persistence()` — read `subject_pars$Q0` and
`$alpha` directly. When those are NA, both abort with a targeted message
pointing at `expanded = TRUE`. Native expanded-shape support in those
consumers is planned for a follow-up release.

```{r plot-guard, eval = FALSE}
plot(fit_m1, type = "individual")
#> Error: Subject-level Q0/alpha are "NA" for this fit.
#> i Default `subject_pars` has no well-defined per-subject value when
#>   a within-subject factor varies within id (factor-expanded or
#>   multi-block RE specs).
#> i Call `get_subject_pars(fit, expanded = TRUE)` for per-(subject,
#>   factor-level) parameters and reduce/aggregate before plotting,
#>   or fit a different RE structure.
#> x Per-(subject, factor-level) plotting is planned for a follow-up
#>   release.
```

The `predict()` and `ranef()` methods work correctly on M1-style fits;
they do not read `subject_pars$Q0` and instead consume the
`re_q0_mat` / `re_alpha_mat` attributes.

# 5. Group metrics with conditioning (Phase 5C)

`calc_group_metrics()` returns population-level Pmax, Omax, Qmax. By
default, continuous covariates are evaluated at their training mean and
factors are marginalized across observed levels with equal weights.

```{r group-default, eval = !render_fast}
metrics_default <- calc_group_metrics(fit_m1)
metrics_default[c("Pmax", "Omax", "Qmax")]
metrics_default$conditioned_on
```

The `conditioned_on` field reports the actual conditioning point. For a
fit with `condition` as a factor, the default is `"marginal"` — Pmax/Omax
are derived from the average log-Q₀ and log-α across condition cells.

To condition on a specific level or covariate value, use `at`:

```{r group-at, eval = !render_fast}
metrics_C1 <- calc_group_metrics(fit_m1, at = list(condition = "C1"))
metrics_C3 <- calc_group_metrics(fit_m1, at = list(condition = "C3"))
data.frame(
  condition = c("C1", "C3"),
  Pmax      = c(metrics_C1$Pmax, metrics_C3$Pmax),
  Omax      = c(metrics_C1$Omax, metrics_C3$Omax)
)
```

## Marginalization order

For derived metrics that depend nonlinearly on (Q₀, α) jointly,
`calc_group_metrics()` uses **parameter-first marginalization**:

1. Compute log-Q₀ and log-α EMMs at each cell of the reference grid.
2. Marginalize each parameter independently across cells (equal weights).
3. Derive Pmax/Omax/Qmax from the marginalized log-parameters.

This is "metrics evaluated at the average parameter values," NOT
"average metrics across cells." For nonlinear transforms (Lambert-W
solutions for Pmax) the two approaches give different answers; we use
parameter-first because it matches the convention `get_demand_param_emms()`
already uses for individual parameters and produces consistent answers
across the EMM family of helpers.

# 6. Diagnostics

## Variance components per block

`summary(fit)` prints variance components grouped by block. For a
multi-block fit, the variance estimates label which block they belong to
(e.g., `block1 sigma_b[(Intercept)]` vs `block2 sigma_b[conditionA]`).

```{r variance-cmp, eval = !render_fast}
summary(fit_m1)
```

A condition-slope variance that lands at the boundary (≈ 0) is a sign
the data does not support per-condition heterogeneity for that
parameter; consider dropping that slope from the spec or pooling
conditions.

## Multi-start

For multi-block fits with rich variance structures, optimizer
sensitivity to starting values increases. The default `multi_start = TRUE`
runs the optimizer from multiple random starts and keeps the best
result; this is recommended for serious fits. The examples in this
vignette use `multi_start = FALSE` purely for speed.

## Convergence troubleshooting

If `fit$converged` is `FALSE` or the Hessian is non-positive-definite:

1. Check `summary(fit)$notes` for optimizer messages.
2. Try `tmb_control = list(optimizer = "L-BFGS-B")` as a fallback.
3. Reduce model complexity: drop the slopes block, or fit pdDiag
   instead of pdSymm.
4. For ill-conditioned Z matrices (e.g., near-collinear factor levels),
   inspect `attr(fit$subject_pars, "re_q0_mat")` and
   `attr(fit$subject_pars, "re_alpha_mat")` for degenerate columns.

# 7. End-to-end workflow

```{r e2e, eval = FALSE}
# Simulate
sim <- beezdemand:::.simulate_within_subject_demand(
  n_subjects = 80, n_conditions = 3,
  log_q0_pop = log(15), log_alpha_pop = log(0.0015),
  delta_q0 = c(0, -0.4, -0.9), delta_alpha = c(0, 0.2, 0.5),
  sigma_b = 0.4, sigma_d = 0.4, seed = 42
)
sim$id <- factor(sim$id)
sim$condition <- factor(sim$condition)

# Fit M1 spec
fit <- fit_demand_tmb(
  sim, equation = "simplified", factors = "condition",
  random_effects = pdBlocked(list(
    pdSymm(Q0 + alpha ~ 1),
    pdDiag(Q0 + alpha ~ condition - 1)
  )),
  multi_start = TRUE, verbose = 0
)

# Inspect
summary(fit)
emms <- get_demand_param_emms(fit, param = "Q0")
cmps <- get_demand_comparisons(fit, param = "Q0")

# Subject-level long form
spars <- get_subject_pars(fit, expanded = TRUE)
head(spars)

# Group metrics, default + per-condition
calc_group_metrics(fit)
calc_group_metrics(fit, at = list(condition = "C1"))
calc_group_metrics(fit, at = list(condition = "C3"))
```

# Cross-references

- `vignette("tmb-mixed-effects")` — intercept-only and basic 2-RE TMB
  fits.
- `vignette("migration-guide")` — migrating from `FitCurves()` to the
  modern S3 API.
- `vignette("convergence-guide")` — optimizer settings and start-value
  strategies.
- `vignette("group-comparisons")` — extra sum-of-squares F-test for
  group-level comparisons.
- `?fit_demand_tmb`, `?get_subject_pars`, `?calc_group_metrics`.
