---
title: "Introduction to R and beezdemand"
author: "Brent Kaplan"
date: "Last Compiled: `r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    highlight: kate
fontsize: 12pt
clean: false
header-includes:
  - \usepackage{fontspec}
  - \usepackage{hyperref}
  - \setmainfont{Georgia}
  - \hypersetup{colorlinks = true, urlcolor = blue, linkcolor = black}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA)
```

# Rationale Behind `beezdemand`

Behavioral economic demand is gaining in popularity. The motivation behind `beezdemand` was to 
create an alternative tool to conduct these
analyses. This package is not necessarily meant to be a replacement
for other softwares; rather, it is meant to serve as an
additional tool in the behavioral economist's toolbox. It is meant for
researchers to conduct behavioral economic (be) demand the easy (ez) way.

## Note About Use

Currently, this version (0.1.0) is the first minor release and is stable. I encourage you to use it
but be aware that, as with any software release, there might be (unknown) bugs present. I've tried hard to make this version usable while including the core functionality (described more below). However, if you find issues or would like to contribute, please open an issue on my [GitHub page](http://github.com/brentkaplan/beezdemand) or [email me](mailto:bkaplan.ku@gmail.com).

## Note About this Document

This specific document is a companion for the paper, "The R Package `beezdemand`: Behavioral Economic Easy Demand", published in [Perspectives on Behavior Science](https://link.springer.com/journal/40614) coauthored by myself, Shawn P. Gilroy, Derek D. Reed, Mikhail N. Koffarnus, and Steven R. Hursh, and is largely adapted from the vignette created for and accessible from the `beezdemand` package. This document is written in R Markdown and is compiled to a PDF with the `rmarkdown` package. R Markdown is a file format for making dynamic documents using `R`. The bulk of this document is written in markdown with `R` code chunks littered throughout. A dynamic document is one that can be updated "on the fly." Over time, a document like this is likely to grow and expand. One such document is the vignette for `beezdemand`, which will be updated as new functions are introduced. You can access the `beezdemand` vignette by either visiting the package website on [CRAN](https://CRAN.R-project.org/package=beezdemand) or [GitHub](http://github.com/brentkaplan/beezdemand), or in `R` using the command, `browseVignettes("beezdemand")`.

The R Markdown version of this document can be found [here](https://github.com/brentkaplan/beezdemand/blob/master/pobs/intro-R-beezdemand.Rmd).

The PDF version of this document can be found [here](https://github.com/brentkaplan/beezdemand/blob/master/pobs/intro-R-beezdemand.pdf).

## Installing `beezdemand`

### CRAN Release (recommended method)

The latest stable version of `beezdemand` (currently v.0.1.0) can be found on [CRAN](https://CRAN.R-project.org/package=beezdemand) and installed using the following command. The first time you install the package, you may be asked to select a CRAN mirror. Simply select the mirror geographically closest to you.

```{r cran-install, eval = FALSE}
install.packages("beezdemand")

library(beezdemand)
```

### GitHub Release

To install a stable release directly from [GitHub](http://github.com/brentkaplan/beezdemand), first install and load the `devtools` package. Then, use `install_github` to install the package and associated vignette. You *don't* need to download anything directly from [GitHub](http://github.com/brentkaplan/beezdemand), as you should use the following instructions:

```{r git-install, eval = FALSE}
install.packages("devtools")

devtools::install_github("brentkaplan/beezdemand", build_vignettes = TRUE)

library(beezdemand)
```

### GitHub Development Version

To install the development version of the package, specify the development branch in `install_github`:

```{r gitdev-install, eval = FALSE}
devtools::install_github("brentkaplan/beezdemand@develop")
```

# Brief Introduction to `R`

The following is a brief introduction to `R`. I highly recommend consulting the recommended resources for a more comprehensive overview of `R`. If you are already familiar with `R`, skip ahead to [Using the Package](#usingpackage)

## Installing `R`

You first need to install `R` from https://cran.r-project.org. The [Center for Research Methods and Data Analysis](www.crmda.ku.edu) at the University of Kansas has helpful resources for installing `R` on [Macintosh](https://crmda.ku.edu/guide-47-mac_r_setup) and [Windows](https://crmda.ku.edu/guide-46-windows_r_setup).

## Integrated Development Environments (IDE)

Rstudio (generally recommended for new and existing users as Rstudio has desirable features)

* https://www.rstudio.com

Emacs + Emacs Speaks Statistics (generally recommended for advanced users)

* https://www.gnu.org/software/emacs/ + http://ess.r-project.org

## Recommended resources for `R`

Learn `R` using `R`

* https://swirlstats.com  

`R` bootcamp

* https://www.jaredknowles.com/r-bootcamp/

Datacamp introduction to `R`

* https://www.datacamp.com/courses/free-introduction-to-r

Official `R` documentation

* https://cran.r-project.org/doc/contrib/usingR.pdf

Another good resource

* https://www.statmethods.net

Specific topic books

* https://www.springer.com/series/6991?detailsPage=titles

Several links and more resources

* https://stats.stackexchange.com/questions/138/free-resources-for-learning-r

## Interacting in `R`

After installing `R`, take some time to play around with some different commands. You will almost certainly need to "assign" variables and this is done using the `<-` symbol. As you interact directly with `R`'s console, you will see `>`, which expects an input. If you run a line and see `+` in the console, then `R` is expecting additional inputs to "complete the command".

```{r interacting}
## assign `x` a series of numbers
x <- c(3, 5, 7, 9, 11)

## see what `x` contains
x

## simple things we can do with a series of numbers
mean(x)

var(x)
```

```{r interacting2}
## assign `y` a series of numbers
y <- c(2, 4, 6, 8, 10)

## plot `x` and `y`
plot(y ~ x, main = "A Nice Plot")
```

Note also the specificity of some functions (described later) that require words to be quoted. For example, in the last line of code above, the title of the plot, `"A Nice Plot"` is in quotes. Throughout this document you will see `##`, which indicates a comment. A comment will not be evaluated or ran by `R`. A comment is for human readability to explain or describe what the code does.

```{r interacting3}
# this is a comment
## this is a comment
### this is a comment
```

## Packages

Packages are collections of functions and sometimes data. Also referred to as libraries, you will most certainly need to install and use a package at some point (e.g., `beezdemand`). You only need to install the package once. From then on, for every unique `R` script, you will want to use the function `library(packagename)` to load the package into the current session.

```{r example-package, eval = FALSE}
install.packages("packagename")
```

Some recommended packages and ones that this document uses include:

```{r example-package1, eval = FALSE}
install.packages("knitr")

install.packages("tidyverse")

install.packages("beezdemand")
```

After the package installs, use the `library(packagename)` command to load the package. Notice that when installing the package, the package name needs to be in quotations. However, the package name does not need to be in quotations when loading the package. 

```{r packages, include = FALSE, echo = FALSE}
if (!require(knitr)) {
  install.packages("knitr")
  library(knitr)
}

if (!require(tidyverse)) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require(beezdemand)) {
  install.packages("beezdemand")
  library(beezdemand)
}
```

## Where Are You?

It's always helpful to know where `R` thinks you are in your computer. The reason why is because one needs to specify exactly where a file is to be saved or where a file is to be read from. If you click on a file in your system and start `R`, the "working directory" will usually be the location of that file. If you start `R` by clicking on the program icon, the working directory will often times be somewhere else. To figure out where you are, use the following command:

```{r where-are-you}
## get working directory
getwd()
```

To change where you are or want to be, use the following command:

```{r where-are-you2, eval = FALSE}
## set working directory
setwd("the/path/goes/here")
```

One important aspect to understand is the idea of relative and absolute paths. For more information on relative and absolute paths, see this [Wiki](https://en.wikipedia.org/wiki/Path_(computing)) entry. Say you have a folder structure like this:
```
projectfolder
│   README.md
│   file001.txt    
│
└───analysisfolder
│   │   file011.R
│   │   file012.sas
│   │
│   └───subfolder1
│       │   file111.txt
│       │   file112.txt
│       │   ...
│   
└───datafolder
    │   file021.csv
    │   file022.txt
```

If you are working on the file `file011.R`, then the working directory (but not the full, absolute path) would be `/projectfolder/analysisfolder/`. A relative path to the project folder (which is one level higher than the analysis folder) would be `../` or a relative path to the `datafolder` would be `../datafolder`. The `../` indicates go one level up from where I am now. In `R`, we can create folders (or directories) using the following command: 

```{r make-dir, eval = FALSE}
dir.create("newfolder")
```

For example, some functions in `beezdemand` will allow you to specify a directory in which to save plots. If we were working in our `analysisfolder` and we wanted to create a folder to save our plots (`plotsfolder`), we would use the following command:

```{r make-dir2, eval = FALSE}
## go up on level (projectfolder) and create a new directory/folder called 
## plotsfolder
dir.create("../plotsfolder")
```

Our resulting folder hierarchy would look like:
```
projectfolder
│   README.md
│   file001.txt    
│
└───analysisfolder
│   │   file011.R
│   │   file012.sas
│   │
│   └───subfolder1
│       │   file111.txt
│       │   file112.txt
│       │   ...
│   
└───datafolder
│   │   file021.csv
│   │   file022.txt
│
└───plotsfolder
```

## Reading Files into `R`

Different file types can be read into `R`. I typically work with plain text files (e.g., .csv) or Excel files (e.g., .xls, .xlsx) with no major formatting. To read in files a variety of options are available, but you should assign the contents of the file into a variable (`dat` in the following example).

```{r example-import, eval = FALSE}
## csv files
dat <- read.csv("path/filename.csv")

## excel files (note, this requires the "readxl" package)
dat <- readxl::read_excel("path/filename.xslx")

## reading a file using a relative directory using the previous example
dat <- read.csv("../datafolder/filename.csv")
```

For the purposes of the current document, we will use the dataset I've included in the `beezdemand` package (see [Example Dataset](#exdata) for more information):

```{r beez-data-set}
dat <- beezdemand::apt
```

## Object Types

There are a number of different object types in `R` and the object types will dictate the procedures or functions you can apply to them. You can see that the object, `dat` is a `data.frame`, which is similar to a table or sheet in Excel, and it has `r ncol(dat)` columns and `r nrow(dat)` rows. Note also that a `data.frame` is just the joining of several different vectors, which are $n$ $X$ $1$ objects containing one type of data (e.g., factor, character, numeric, integer).

```{r data-types}
class(dat)

str(dat)

dplyr::as_tibble(dat)
```

Contrasted with a `matrix`, which can only support one data type (e.g., numeric, character), a `data.frame` can consist of several vectors that each have a different data type. We will primary deal with `data.frame`s because they can support different data types.

```{r data-types2}
exampledf <- data.frame("character" = c("A", "B", "C"), "number" = c(1, 2, 3))
dplyr::as_tibble(exampledf)

exampledf$number <- as.character(exampledf$number)
dplyr::as_tibble(exampledf)
```

The last object type is a `list`. A `list` can have many different data types and `list`s can be nested within one another. 

```{r data-types3}
examplelst <- vector(mode = "list", length = 3)
examplelst

## assign the first slot in examplelst with a subset (rows 1-5) of dat
examplelst[[1]] <- dat[1:5, ]  
## assign the second slot of examplelst with our exampledf
examplelst[[2]] <- exampledf
examplelst
```

## Functions

Nearly everything in `R` is a function. A function can and usually takes some input and returns an output. Functions are especially useful if you need to do some task many times, rather than copying and pasting lines of identical code. Although you don't need to *make* your own functions for using `beezdemand`, you will need to *use* many of the functions in the package.

Making or declaring the function:
```{r functions}
## addTwoNumbers is the function name
## num1 and num2 are arguments that the function takes
## num1 and num2 are added together and assigned to res
## a message is displayed
## the sum of the numbers (res) is returned from the function
addTwoNumbers <- function(num1, num2) {
  res <- num1 + num2
  message(paste("The result of adding", num1, "and", num2, "is:"))
  res
}
```

Calling the function and the result of the function:
```{r functions2}
addTwoNumbers(num1 = 2, num2 = 3)
```

## Getting Help

Almost every function that is available in an `R` package comes with documentation. This documentation is helpful for understanding how to use the function. For example, the documentation for the `sum` function can be accessed using `?` in front of the function name (minus the `()`).

```{r help, eval = F}
?sum
help(sum)
```

To access the help page for a package, use the following command. Note, this will only work if you have the package installed.

```{r help2, eval = F}
help(package = "beezdemand")
```

## What To Do If Something Doesn't Work

If you get an error or warning, first read and try and understand the problem. Some error and warning messages are more informative than others. If you can't figure out why something is not working, make sure the code and data are correct and you are not missing anything. Some functions won't work if the data are not in the correct format. For example, if the function is expecting a `data.frame` but you provide it a single vector, the function is not likely to work and may or may not tell you why. Other times, you will refer to functions located in a package but forget to load the package (e.g., `library(packagename)`). 

If you've read the error/warning message and made sure the code and data are correctly specified, search for the problem on [Google](www.google.com) or a site such as [StackOverflow](https://stackoverflow.com). Most likely, someone has encountered and solved the same or similar issue you're encountering. There is likely to be less information and solutions to problems resulting from using `beezdemand`. When this is the case, please open an issue on the package's [GitHub page](http://github.com/brentkaplan/beezdemand) or [email me](mailto:bkaplan.ku@gmail.com).

# Using the Package {#usingpackage}

## Example Dataset {#exdata}

An example dataset of responses on an Alcohol Purchase Task is provided. This object is called `apt` and is located within the `beezdemand` package. These data are a subset of from the paper by [Kaplan & Reed (2018)](http://psycnet.apa.org/record/2018-02774-001). Participants (id) reported the number of alcoholic drinks (y) they would be willing to purchase and consume at various prices (x; USD). Note the format of the data, which is called "long format". Long format data are data structured such that repeated observations are stacked in multiple rows, rather than across columns. First, take a look at an extract of the dataset `apt`, where I've subsetted rows 1 through 10 and 17 through 26:

```{r example-data-set, echo=FALSE, results='asis'}
knitr::kable(apt[c(1:10, 17:26), ])
```

The first column contains the row number. The second column contains the id number of the series within the dataset. The third column contains the x values (in this specific dataset, price per drink) and the fourth column contains the associated responses (number of alcoholic drinks purchased at each respective price). There are replicates of id because for each series (or participant), several x values were presented.

## Converting from Wide to Long and Vice Versa

Take for example the format of most datasets that would be exported from a data collection software such as Qualtrics or SurveyMonkey or Google Forms:

```{r example-wide}
## the following code takes the apt data, which are in long format, and converts
## to a wide format that might be seen from data collection software
wide <- spread(apt, x, y)
colnames(wide) <- c("id", paste0("price_", seq(1, 16, by = 1)))
knitr::kable(wide[1:5, 1:10])
```

A dataset such as this is referred to as "wide format" because each participant series contains a single row and multiple measurements within the participant are indicated by the columns. This data format is fine for some purposes; however, for `beezdemand`, data are required to be in "long format" (in the same format as the example data described [earlier](#exdata)). In order to convert to the long format, some steps will be required.

First, it is helpful to rename the columns to what the prices actually were. For example, for the purposes of our example dataset, price_1 was \$0.00 (free), price_2 was \$0.50, price_3 was \$1.00, and so on. 

```{r example-wide1}
## make an object to hold what will be the new column names
newcolnames <- c("id", "0", "0.5", "1", "1.50", "2", "2.50", "3", 
                 "4", "5", "6", "7", "8", "9", "10", "15", "20")
## current column names
colnames(wide)

## replace current column names with new column names
colnames(wide) <- newcolnames

## how new data look (first 5 rows only)
knitr::kable(wide[1:5, ])
```

Now we can convert into a long format using some of the helpful functions in the `tidyverse` package (make sure the package is loaded before trying the commands below).

```{r example-w2l}
## using the dataframe 'wide', we specify the key will be 'price', the values 
## will be 'consumption', and we will select all columns besides the first ('id')
long <- tidyr::gather(wide, price, consumption, -id)

## we'll sort the rows by id
long <- arrange(long, id)

## view the first 20 rows
knitr::kable(long[1:20, ])
```

Two final modifications we will make will be to (1) rename our columns to what the functions in `beezdemand` will expect to see: `id`, `x`, and `y`, and (2) ensure both x and y are in numeric format.
```{r example-w2l2}
colnames(long) <- c("id", "x", "y")

long$x <- as.numeric(long$x)
long$y <- as.numeric(long$y)
knitr::kable(head(long))
```

The dataset is now "tidy" because: (1) each variable forms a column, (2) each observation forms a row, and (3) each type of observational unit forms a table (in this case, our observational unit is the Alcohol Purchase Task data). To learn more about the benefits of tidy data, readers are encouraged to consult Hadley Wikham's essay on [Tidy Data](http://vita.had.co.nz/papers/tidy-data.html).

## Obtain Descriptive Data

Descriptive values of responses at each price can be obtained easily. The resulting table includes mean, standard
deviation, proportion of zeros, number of NAs, and minimum and maximum values. If `bwplot = TRUE`, a box-and-whisker plot is also created and saved. By default, this location is a folder named "plots" one level up from the current working directory.  The user may additionally specify the directory that the plot should save into, the type of file (either `"png"` or `"pdf"`), and the filename. Notice the red crosses indicate the mean. Defaults are shown here:

```{r descriptive, eval = FALSE}
GetDescriptives(dat = apt, bwplot = FALSE, outdir = "../plots/", device = "png", 
                filename = "bwplot")
```

To actually run the code and generate the file, we will turn `bwplot = TRUE`. The function will create a folder one level higher than the current folder (i.e., the `../` portion) and save the file, "bwplot.png" in the folder.

```{r descriptive2, results = 'hide'}
GetDescriptives(dat = apt, bwplot = TRUE, outdir = "../plots/", device = "png", 
                filename = "bwplot")
```

\begin{center}
\includegraphics[width=0.75\textwidth]{../plots/bwplot.png}
\end{center}

And here is the table that is returned from the function:

```{r descriptive-output, echo=FALSE, results='asis'}
descr <- GetDescriptives(dat = apt)
knitr::kable(descr)
```

## Change Data

There are certain instances in which data are to be modified before fitting, for example when using an equation that logarithmically transforms y values. The following function can help with modifying data:

* `nrepl` indicates number of replacement 0 values, either as an integer or `"all"`. If this value is an integer, `n`, then the first `n` 0s will be replaced. 

* `replnum` indicates the number that should replace 0 values

* `rem0` removes all zeros

* `remq0e` removes y value where x (or price) equals 0

* `replfree` replaces where x (or price) equals 0 with a specified number

```{r change-data, eval = FALSE}
ChangeData(dat = apt, nrepl = 1, replnum = 0.01, rem0 = FALSE, remq0e = FALSE, 
           replfree = NULL)
```

## Identify Unsystematic Responses

Using the following function, we can examine the consistency of demand data using [Stein et al.'s (2015)](http://psycnet.apa.org/record/2015-30199-001)
alogrithm for identifying unsystematic responses. Default values
shown, but they can be customized.

```{r unsystematic, eval=FALSE}
CheckUnsystematic(dat = apt, deltaq = 0.025, bounce = 0.1, reversals = 0, ncons0 = 2)
```

```{r unsystematic-output, echo=FALSE, results='asis'}
knitr::kable(head(CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.1, 
                                    reversals = 0, ncons0 = 2), 5))
```

## Analyze Demand Data

Results of the analysis return both empirical and derived measures for
use in additional analyses and model specification. Equations include
the linear model, exponential model, and exponentiated model. Soon, I
will be including the nonlinear mixed effects model, mixed effects
versions of the exponential and exponentiated model, and others. However, currently these models are not yet supported.

### Obtaining Empirical Measures

Empirical measures can be obtained separately on their own:

```{r empirical, eval=FALSE}
GetEmpirical(dat = apt)
```

```{r empirical-output, echo=FALSE, results='asis'}
knitr::kable(head(GetEmpirical(apt), 5))
```

### Obtaining Derived Measures

`FitCurves()` has several important arguments that can be passed. For the purposes of this document, focus will be on the two contemporary demand equations.

* `equation = "hs"` is the default but can accept the character strings `"linear"`, `"hs"`, or `"koff"`, the latter two of which are the contemporary equations proposed by [Hursh & Silberberg (2008)](http://psycnet.apa.org/record/2008-00265-008) and [Koffarnus et al. (2015)](http://psycnet.apa.org/record/2015-37520-001), respectively. 

* `k` can take accept a specific number but by default will be calculated based on the maximum and minimum y values of the entire sample and adding .5. Adding this amount was originally proposed by Steven R. Hursh in an early iteration of a Microsoft Excel spreadsheet used to calculate demand metrics. This adjustment was adopted for two reasons. First, when fitting $Q_0$ as a derived parameter, the value may exceed the empirically observed intensity value. Thus, a k value calculated based only on the observed range of data may underestimate the full fitted range of the curve. Second, we have found that values of $\alpha$ (as well as values that rely on $\alpha$, i.e. approximate $P_{max}$) display greater discrepancies when smaller values of k are used compared to larger values of k. Other options include `"ind"`, which will calculate k based on individual basis, `"fit"`, which will fit k as a free parameter on an individual basis, `"share"`, which will fit k as a single shared parameter across all data sets (while fitting individual $Q_0$ and $\alpha$).

* `agg = NULL` is the default, which means no aggregation. When `agg = "Mean"`, models are fit to the averaged data disregarding any error. When `agg = "Pooled"`, all data are used and clustering within individual is ignored.

* `detailed = FALSE` is the default. This will output a single dataframe of results, as shown below. When `detailed = TRUE`, the output is a 3 element list that includes (1) dataframe of results, (2) list of nonlinear regression model objects, (3) list of dataframes containing predicted x and y values (to be used in subsequent plotting), and (4) list of individual dataframes used in fitting.

* `lobound` and `hibound` can accept named vectors that will be used as lower and upper bounds, respectively during fitting. If `k = "fit"`, then it should look as follows (values are nonspecific): `lobound = c("q0" = 0, "k" = 0, "alpha" = 0)` and `hibound = c("q0" = 25, "k" = 10, "alpha" = 1)`. If `k` is not being fit as a parameter, then only `"q0"` and `"alpha"` should be used in bounding.

Note: Fitting with an equation (e.g., `"linear"`, `"hs"`) that doesn't work happily with zero consumption values results in the following. One, a message will appear saying that zeros are incompatible with the equation. Two, because zeros are removed prior to finding empirical (i.e., observed) measures, resulting BP0 values will be all NAs (reflective of the data transformations). The warning message will look as follows:

```{r zero-warning, eval=FALSE}
Warning message:
Zeros found in data not compatible with equation! Dropping zeros!
```

The simplest use of `FitCurves()` is shown here, only needing to specify `dat` and `equation`. All other arguments shown are set to their default values.

```{r hs, eval=FALSE}
FitCurves(dat = apt, equation = "hs", agg = NULL, detailed = FALSE, 
          xcol = "x", ycol = "y", idcol = "id", groupcol = NULL)
```

Which is equivalent to:
```{r hs2, eval=FALSE}
FitCurves(dat = apt, equation = "hs")
```

Note that this ouput returns a message (`No k value specified. Defaulting to empirical mean range +.5`) and the aforementioned warning (`Warning message: Zeros found in data not compatible with equation! Dropping zeros!`). With `detailed = FALSE`, the only output is the dataframe of results (broken up to show the different types of results). This example fits the exponential equation proposed by [Hursh & Silberberg (2008)](http://psycnet.apa.org/record/2008-00265-008):

```{r hs-setup, include=FALSE}
fc <- FitCurves(dat = apt, equation = "hs")
hs1 <- head(fc, 5)[ , 1:6]
hs2 <- head(fc, 5)[ , 7:11]
hs3 <- head(fc, 5)[ , 12:20]
hs4 <- head(fc, 5)[ , 21:24]
```

```{r hs-output, echo=FALSE, results='asis'}
knitr::kable(hs1, caption = "Empirical Measures")
knitr::kable(hs2, caption = "Fitted Measures")
knitr::kable(hs3, caption = "Uncertainty and Model Information")
knitr::kable(hs4, caption = "Derived Measures")
```

Here, the simplest form is shown specifying another equation, `"koff"`. This fits the modified exponential equation proposed by [Koffarnus et al. (2015)](http://psycnet.apa.org/record/2015-37520-001):

```{r koff, eval=FALSE}
FitCurves(dat = apt, equation = "koff")
```

```{r koff-setup, include = FALSE}
fc <- FitCurves(dat = apt, equation = "koff")
koff1 <- head(fc, 5)[ , 1:6]
koff2 <- head(fc, 5)[ , 7:11]
koff3 <- head(fc, 5)[ , 12:20]
koff4 <- head(fc, 5)[ , 21:24]
```

```{r koff-output, echo=FALSE, results='asis'}
knitr::kable(koff1, caption = "Empirical Measures")
knitr::kable(koff2, caption = "Fitted Measures")
knitr::kable(koff3, caption = "Uncertainty and Model Information")
knitr::kable(koff4, caption = "Derived Measures")
```

By specifying `agg = "Mean"`, y values at each x value are aggregated and a single curve is fit to the data (disregarding error around each averaged point):

```{r agg-mean, eval = FALSE}
FitCurves(dat = apt, equation = "hs", agg = "Mean")
```

```{r agg-mean-setup, include = FALSE}
mn <- FitCurves(dat = apt, equation = "hs", agg = "Mean")
mn1 <- head(mn)[ , 1:6]
mn2 <- head(mn)[ , 7:11]
mn3 <- head(mn)[ , 12:20]
mn4 <- head(mn)[ , 21:24]
```

```{r agg-mean-output, echo = FALSE, results = 'asis'}
knitr::kable(mn1, caption = "Empirical Measures")
knitr::kable(mn2, caption = "Fitted Measures")
knitr::kable(mn3, caption = "Uncertainty and Model Information")
knitr::kable(mn4, caption = "Derived Measures")
```

By specifying `agg = "Pooled"`, y values at each x value are aggregated and a single curve is fit to the data and error around each averaged point (but disregarding within-subject clustering):

```{r agg-pooled, eval = FALSE}
FitCurves(dat = apt, equation = "hs", agg = "Pooled")
```

```{r agg-pooled-setup, include = FALSE}
pl <- FitCurves(dat = apt, equation = "hs", agg = "Pooled")
pl1 <- head(pl)[ , 1:6]
pl2 <- head(pl)[ , 7:11]
pl3 <- head(pl)[ , 12:20]
pl4 <- head(pl)[ , 21:24]
```

```{r agg-pooled-output, echo = FALSE, results = 'asis'}
knitr::kable(pl1, caption = "Empirical Measures")
knitr::kable(pl2, caption = "Fitted Measures")
knitr::kable(pl3, caption = "Uncertainty and Model Information")
knitr::kable(pl4, caption = "Derived Measures")
```

## Share k Globally; Fit Other Parameters Locally

As mentioned earlier, in the function `FitCurves`, when `k = "share"` this parameter will be a shared parameter across all datasets (globally) while estimating $Q_0$ and $\alpha$ locally. While this works, it may take some time with larger sample sizes.

```{r share, eval=FALSE}
FitCurves(dat = apt, equation = "hs", k = "share")
```

```{r, include=FALSE}
df <- FitCurves(dat = apt, equation = "hs", k = "share")
```

```{r share-output, echo=FALSE, results='asis'}
knitr::kable(head(df, 5)[ , 1:6], caption = "Empirical Measures")
knitr::kable(head(df, 5)[ , 7:11], caption = "Fitted Measures")
knitr::kable(head(df, 5)[ , 12:20], caption = "Uncertainty and Model Information")
knitr::kable(head(df, 5)[ , 21:24], caption = "Derived Measures")
```

## Compare Values of $\alpha$ and $Q_0$ via Extra Sum-of-Squares _F_-Test

When one has multiple groups, it may be beneficial to compare whether separate curves are preferred over a single curve. This is accomplished by the Extra Sum-of-Squares _F_-test. This function (using the argument `compare`) will determine whether a single $\alpha$ or a single $Q_0$ is better than multiple $\alpha$s or $Q_0$s. A single curve will be fit, the residual deviations calculated and those residuals are compared to residuals obtained from multiple curves. A resulting _F_ statistic will be reporting along with a _p_ value. 

```{r ftest}
## setting the seed initializes the random number generator so results will be 
## reproducible
set.seed(1234)

## manufacture random grouping
apt$group <- NA
apt[apt$id %in% sample(unique(apt$id), length(unique(apt$id))/2), "group"] <- "a"
apt$group[is.na(apt$group)] <- "b"

## take a look at what the new groupings look like in long form
knitr::kable(apt[1:20, ])
```

```{r ftest2}
## in order for this to run, you will have had to run the code immediately
## preceeding (i.e., the code to generate the groups)
ef <- ExtraF(dat = apt, equation = "koff", k = 2, groupcol = "group", verbose = TRUE)
```

A summary table (broken up here for ease of display) will be created when the option `verbose = TRUE`. This table can be accessed as the `dfres` object resulting from `ExtraF`. In the example above, we can access this summary table using `ef$dfres`:

```{r ftest-ouput, results = 'asis', echo=FALSE}
knitr::kable(ef$dfres[, 1:5], caption  = "Fitted Measures")
knitr::kable(ef$dfres[, c(1, 6:8)], caption = "Uncertainty and Model Information")
knitr::kable(ef$dfres[, c(1, 9:11)], caption = "Derived Measures")
knitr::kable(ef$dfres[, c(1, 12, 14)], caption = "Convergence and Summary Information")
```

When `verbose = TRUE`, objects from the result can be used in subsequent graphing. The following code generates a plot of our two groups. We can use the predicted values already generated from the `ExtraF` function by accessing the `newdat` object. In the example above, we can access these predicted values using `ef$newdat`. Note that we keep the linear scaling of y given we used [Koffarnus et al. (2015)'s](http://psycnet.apa.org/record/2015-37520-001) equation fitted to the data.

```{r plot-ftest, warning = FALSE}
## be sure that you've loaded the tidyverse package (e.g., library(tidyverse))
ggplot(apt, aes(x = x, y = y, group = group)) +
  ## the predicted lines from the sum of squares f-test can be used in subsequent
  ## plots by calling data = ef$newdat
  geom_line(aes(x = x, y = y, group = group, color = group), 
            data = ef$newdat[ef$newdat$x >= .1, ]) +
  stat_summary(fun.data = mean_se, aes(width = .05, color = group), 
               geom = "errorbar") +
  stat_summary(fun.y = mean, aes(fill = group), geom = "point", shape = 21, 
               color = "black", stroke = .75, size = 4) +
  scale_x_log10(limits = c(.4, 50), breaks = c(.1, 1, 10, 100)) +
  scale_color_discrete(name = "Group") +
  scale_fill_discrete(name = "Group") +
  labs(x = "Price per Drink", y = "Drinks Purchased") +
  theme(legend.position = c(.85, .75)) +
  ## theme_apa is a beezdemand function used to change the theme in accordance
  ## with American Psychological Association style
  theme_apa()
```

## Plots

Plots can be created using the `PlotCurves` function. This function takes the output from `FitCurves` when the argument from `FitCurves`, `detailed = TRUE`. The default will be to save figures into a plots folder created one directory above the current working directory. Figures can be saved as either PNG or PDF. If the argument `ask = TRUE`, then plots will be shown interactively and not saved (`ask = FALSE` is the default). Graphs can automatically be created at both an aggregate and individual level.

As a demonstration, let's first use `FitCurves` on the `apt` dataset, specifying `k = "share"` and `detailed = T`. This will return a list of objects to use in `PlotCurves`. In `PlotCurves`, we will feed in our new object, `out`, and tell the function to save the plots in the directory `"../plots/"` and `ask = FALSE` because we don't want `R` to interactively show us each plot. Because we have 10 datasets in our `apt` example, 10 plots will be created and saved in the `"../plots/"` directory.
```{r plots1}
out <- FitCurves(dat = apt, equation = "hs", k = "share", detailed = T)

PlotCurves(dat = out, outdir = "../plots/", device = "png", ask = F)
```

\begin{center}
\includegraphics[width=0.75\textwidth]{../plots/Participant-19.png}
\end{center}

We can also make a plot of the mean data. Here, we again use `FitCurves`, this time calculating a k from the observed range of the data (thus not specifying any k) and specifying `agg = "Mean"`.
```{r plots2}
mn <- FitCurves(dat = apt, equation = "hs", agg = "Mean", detailed = T)

PlotCurves(dat = mn, outdir = "../plots/", device = "png", ask = F)
```

\begin{center}
\includegraphics[width=0.75\textwidth]{../plots/Participant-mean.png}
\end{center}

## Learn More About Functions

To learn more about a function and what arguments it takes, type "?" in front of the function name.

```{r learn, eval=FALSE}
?CheckUnsystematic
```

```{r learn-output, eval=FALSE}
CheckUnsystematic          package:beezdemand          R Documentation

Systematic Purchase Task Data Checker

Description:

     Applies Stein, Koffarnus, Snider, Quisenberry, & Bickels (2015)
     criteria for identification of nonsystematic purchase task data.

Usage:

     CheckUnsystematic(dat, deltaq = 0.025, bounce = 0.1, reversals = 0,
       ncons0 = 2)

Arguments:

     dat: Dataframe in long form. Colums are id, x, y.

  deltaq: Numeric vector of length equal to one. The criterion by which
          the relative change in quantity purchased will be compared.
          Relative changes in quantity purchased below this criterion
          will be flagged. Default value is 0.025.

  bounce: Numeric vector of length equal to one. The criterion by which
          the number of price-to-price increases in consumption that
          exceed 25% of initial consumption at the lowest price,
          expressed relative to the total number of price increments,
          will be compared. The relative number of price-to-price
          increases above this criterion will be flagged. Default value
          is 0.10.

reversals:Numeric vector of length equal to one. The criterion by
          which the number of reversals from number of consecutive (see
          ncons0) 0s will be compared. Number of reversals above this
          criterion will be flagged. Default value is 0.

  ncons0: Number of consecutive 0s prior to a positive value is used to
          flag for a reversal. Value can be either 1 (relatively more
          conservative) or 2 (default; as recommended by Stein et al.,
          (2015).

Details:

     This function applies the 3 criteria proposed by Stein et al.,
     (2015) for identification of nonsystematic purchase task data. The
     three criteria include trend (deltaq), bounce, and reversals from
     0. Also reports number of positive consumption values.

Value:

     Dataframe

Author(s):

     Brent Kaplan <bkaplan.ku@gmail.com>

Examples:

     ## Using all default values
     CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.10, reversals = 0, ncons0 = 2)
     ## Specifying just 1 zero to flag as reversal
     CheckUnsystematic(apt, deltaq = 0.025, bounce = 0.10, reversals = 0, ncons0 = 1)
```

# Free `beezdemand` Alternative with Graphical User Interface

If you are interested in using open source software for analyzing demand curve data but don't know `R`, please check out the [Demand Curve Analyzer](http://smallnstats.com/DemandAnalysis.html). The Demand Curve Analyzer contains many of the functions `beezdemand` provides but with the benefits of easy Excel-like functionality. See the companion article [here]().

# Acknowledgments

- Shawn P. Gilroy, Contributor [GitHub](https://github.com/miyamot0)

- Derek D. Reed, Applied Behavioral Economics Laboratory
[www.behavioraleconlab.com](www.behavioraleconlab.com)

- Mikhail N. Koffarnus, Addiction Recovery Research Center 
[https://labs.vtc.vt.edu/arrc/](https://labs.vtc.vt.edu/arrc/)

- Steven R. Hursh, Institutes for Behavior Resources, Inc.
[www.ibrinc.org](www.ibrinc.org)

- Paul E. Johnson, Center for Research Methods and Data Analysis, University of Kansas
[www.crmda.ku.edu](www.crmda.ku.edu)

- Peter G. Roma, Institutes for Behavior Resources, Inc.
[www.ibrinc.org](www.ibrinc.org)

- W. Brady DeHart, Addiction Recovery Research Center
[https://labs.vtc.vt.edu/arrc/](https://labs.vtc.vt.edu/arrc/)

- Michael Amlung, Cognitive Neuroscience of Addictions Laboratory
[www.cnalab.weebly.com](www.cnalab.weebly.com)


Special thanks to the following people who helped provide feedback on this document:

- Alexandra M. Mellis

- Mr. Jeremiah "Downtown Jimbo Brown" Brown

- Gideon Naudé


# Recommended Readings

- Reed, D. D., Niileksela, C. R., & Kaplan, B. A. (2013). Behavioral economics: A tutorial for behavior analysts in practice. *Behavior Analysis in Practice, 6* (1), 34–54. https://doi.org/10.1007/BF03391790

- Reed, D. D., Kaplan, B. A., & Becirevic, A. (2015). Basic research on the behavioral economics of reinforcer value. In *Autism Service Delivery* (pp. 279-306). Springer New York. https://doi.org/10.1007/978-1-4939-2656-5_10

- Hursh, S. R., & Silberberg, A. (2008). Economic demand and essential value. *Psychological Review, 115* (1), 186-198. http://dx.doi.org/10.1037/0033-295X.115.1.186

- Koffarnus, M. N., Franck, C. T., Stein, J. S., & Bickel, W. K. (2015). A modified exponential behavioral economic demand model to better describe consumption data. *Experimental and Clinical Psychopharmacology, 23* (6), 504-512. http://dx.doi.org/10.1037/pha0000045

- Stein, J. S., Koffarnus, M. N., Snider, S. E., Quisenberry, A. J., & Bickel, W. K. (2015). Identification and management of nonsystematic purchase task data: Toward best practice. *Experimental and Clinical Psychopharmacology 23* (5), 377-386. http://dx.doi.org/10.1037/pha0000020

- Hursh, S. R., Raslear, T. G., Shurtleff, D., Bauman, R., & Simmons, L. (1988). A cost‐benefit analysis of demand for food. *Journal of the Experimental Analysis of Behavior, 50* (3), 419-440. https://doi.org/10.1901/jeab.1988.50-419

