---
title: "Deferred Cross-Price Hurdle and Joint Hurdle Vignette Sections"
note: "Deferred from v0.2.0. Restore for next release."
---

## Two-Part Hurdle Cross-Price Model

The hurdle cross-price model extends the standard cross-price demand framework to handle
datasets with many zero consumption values. It uses a two-part structure:

**Part I (Zero Model):** Logistic regression for probability of zero consumption
$$\text{logit}(P(\text{zero})) = \beta_0 + \beta_1 \ln(x + \epsilon) + a_i$$

**Part II (Intensity Model):** Exponential cross-price model for positive consumption
$$\ln(Q) = \ln(Q_{alone}) + I \cdot \exp(-\beta \cdot x) + b_i + e$$

Where:
- $Q_{alone}$ is baseline consumption when alternative is unavailable
- $I$ is the interaction parameter (negative = substitutes, positive = complements)
- $\beta$ is the rate of decay
- $a_i, b_i$ are subject-specific random effects

### Fitting a 2-RE Hurdle Model

```{r hurdle-2re, warning=FALSE}
# Filter to E-Cigarettes alternative condition
ecig_data <- etm |>
    dplyr::filter(group == "E-Cigarettes", target == "alt") |>
    as.data.frame()

# Check zero proportion
cat("Zero consumption:", round(100 * mean(ecig_data$y == 0), 1), "%\n")

# Fit 2-RE hurdle model (random effects on zeros and Qalone)
fit_hurdle_2re <- fit_cp_hurdle(
    data = ecig_data,
    y_var = "y",
    x_var = "x",
    id_var = "id",
    random_effects = c("zeros", "qalone"),
    verbose = 0
)

summary(fit_hurdle_2re)
```

### Fitting a 3-RE Hurdle Model

The 3-RE model adds a random effect on the interaction parameter $I$:

```{r hurdle-3re, warning=FALSE}
# Fit 3-RE hurdle model
fit_hurdle_3re <- fit_cp_hurdle(
    data = ecig_data,
    y_var = "y",
    x_var = "x",
    id_var = "id",
    random_effects = c("zeros", "qalone", "I"),
    verbose = 0
)

summary(fit_hurdle_3re)
```

### Comparing Models

Use a likelihood ratio test to compare nested models:

```{r hurdle-compare}
compare_cp_hurdle_models(fit_hurdle_3re, fit_hurdle_2re)
```

### Subject-Level Parameters

```{r hurdle-params}
# Get parameter summary
get_cp_hurdle_param_summary(fit_hurdle_2re)

# View individual subject parameters
head(predict(fit_hurdle_2re, type = "parameters"))
```

### Predictions

```{r hurdle-predict}
# Population-level predictions
new_prices <- data.frame(x = c(2, 8, 16, 32, 64))
cbind(
    new_prices,
    demand = predict(fit_hurdle_2re, new_prices, type = "response", level = "population")$.fitted,
    prob_zero = predict(fit_hurdle_2re, new_prices, type = "probability", level = "population")$.fitted
)
```

### Plotting

```{r hurdle-plot, fig.width=7, fig.height=5}
# Population demand curve
plot(fit_hurdle_2re, type = "demand")

# Probability of zero consumption
plot(fit_hurdle_2re, type = "probability")

# Individual curves
plot(fit_hurdle_2re, type = "individual")
```

## Joint Hurdle Cross-Price Model

The joint hurdle model extends the separate demand and cross-price models by estimating
all parameters within a **single likelihood**. This approach:

- Shares information across conditions via correlated random effects
- Provides a unified framework for analyzing complete cross-price experiments
- Enables direct comparison of target demand with and without an alternative

### Model Structure

The joint model estimates three "streams" simultaneously:

1. **alone.target**: Target commodity demand when alternative is absent
2. **own.target**: Target commodity demand when alternative is present
3. **own.alt**: Alternative commodity consumption as a function of target price

**Key design features:**

- **Shared hurdle**: A single subject-level random effect ($a_i$) captures propensity
  to purchase anything, with stream-specific intercepts
- **Global k**: The scaling parameter $k$ is shared across target demand streams
- **price_T is the predictor for all streams**: Alternative price is assumed fixed
  and is not modeled as a predictor

### Preparing Data for Joint Model

The `prepare_joint_data()` function transforms data into the required format:

```{r joint-prep}
# Load the cross-price example dataset
data("cp", package = "beezdemand")

# Examine the data structure
cat("Data dimensions:", nrow(cp), "rows x", ncol(cp), "cols\n")
cat("Subjects:", length(unique(cp$id)), "\n")
cat("Conditions:", paste(unique(cp$target), collapse = ", "), "\n")

# View sample data
head(cp)

# Prepare for joint model
prep <- prepare_joint_data(
    data = cp,
    y_var = "y",
    price_var = "x",
    id_var = "id",
    target_var = "target",
    alone_value = "alone",
    own_value = "own",
    alt_value = "alt"
)

# Check stream counts
prep$stream_counts
```

### Comparing Separate Fits vs Joint Fit

#### Step 1: Separate Fits (Traditional Workflow)

First, let's fit each condition separately using existing functions:

```{r joint-separate-fits}
# Filter data by condition
alone_data <- cp |> dplyr::filter(target == "alone")
own_data <- cp |> dplyr::filter(target == "own")
alt_data <- cp |> dplyr::filter(target == "alt")

# Fit target demand (alone condition)
fit_alone <- fit_demand_fixed(
    data = alone_data,
    equation = "koff",
    k = 2
)

coef_alone <- coef(fit_alone)
Q0_alone <- coef_alone$estimate[coef_alone$term == "q0"]
Alpha_alone <- coef_alone$estimate[coef_alone$term == "alpha"]

cat("=== Alone Condition (Target Demand) ===\n")
cat("Q0:", round(Q0_alone, 2), "\n")
cat("Alpha:", round(Alpha_alone, 6), "\n")

# Fit target demand (own condition)
fit_own <- fit_demand_fixed(
    data = own_data,
    equation = "koff",
    k = 2
)

coef_own <- coef(fit_own)
Q0_own <- coef_own$estimate[coef_own$term == "q0"]
Alpha_own <- coef_own$estimate[coef_own$term == "alpha"]

cat("\n=== Own Condition (Target Demand) ===\n")
cat("Q0:", round(Q0_own, 2), "\n")
cat("Alpha:", round(Alpha_own, 6), "\n")

# Fit cross-price model (alt condition)
fit_alt <- fit_cp_nls(
    dat = alt_data,
    equation = "exponentiated",
    return_all = TRUE
)

cat("\n=== Alt Condition (Cross-Price) ===\n")
cat("Qalone:", round(coef(fit_alt)["qalone"], 2), "\n")
cat("I:", round(coef(fit_alt)["I"], 4), "\n")
cat("beta:", round(coef(fit_alt)["beta"], 4), "\n")
```

#### Step 2: Joint Model Fit

Now fit all streams simultaneously with the joint hurdle model:

```{r joint-fit, warning=FALSE}
# Fit the joint hurdle model
fit_joint <- fit_joint_hurdle(
    data = cp,
    y_var = "y",
    price_var = "x",
    id_var = "id",
    target_var = "target",
    alone_value = "alone",
    own_value = "own",
    alt_value = "alt",
    k = 2, # Fixed k (matches existing workflow)
    epsilon = 0.001,
    verbose = 1
)
```

### Examining Joint Model Results

```{r joint-summary}
# Full summary
summary(fit_joint)
```

### Extracting Coefficients

```{r joint-coef}
# All coefficients
coef(fit_joint, type = "all")

# Hurdle parameters only
coef(fit_joint, type = "hurdle")

# Demand parameters (target streams)
coef(fit_joint, type = "demand")

# Cross-price parameters
coef(fit_joint, type = "crossprice")
```

### Comparing Parameter Estimates

```{r joint-compare-params}
# Create comparison table
comparison <- data.frame(
    Stream = c("alone.target", "own.target", "own.alt"),
    Separate_Q0 = c(
        fit_alone$dfres$Q0d,
        fit_own$dfres$Q0d,
        coef(fit_alt)["qalone"]
    ),
    Joint_Q0 = c(
        exp(coef(fit_joint)["logQ0_AT"]),
        exp(coef(fit_joint)["logQ0_OT"]),
        exp(coef(fit_joint)["logQalone_OA"])
    ),
    Separate_Sens = c(
        fit_alone$dfres$Alpha,
        fit_own$dfres$Alpha,
        coef(fit_alt)["I"]
    ),
    Joint_Sens = c(
        coef(fit_joint)["alpha_AT"],
        coef(fit_joint)["alpha_OT"],
        coef(fit_joint)["I"]
    )
)

comparison
```

### Predictions by Stream

```{r joint-predict}
# Population-level predictions for all streams
prices <- data.frame(price_T = c(0.5, 1, 2, 4, 8, 16, 32))

# Response predictions (expected consumption)
pred_response <- predict(fit_joint, newdata = prices, type = "response", stream = "all")
head(pred_response, 12)

# Probability of zero consumption
pred_prob <- predict(fit_joint, newdata = prices, type = "probability", stream = "all")
head(pred_prob, 12)
```

### Subject-Level Parameters

```{r joint-subject-pars}
# Extract subject-specific parameters
subject_pars <- predict(fit_joint, type = "parameters")
subject_pars
```

### Visualizing Joint Model Results

```{r joint-plot-demand, fig.width=8, fig.height=5}
# Demand curves for all streams
plot(fit_joint, type = "demand")
```

```{r joint-plot-prob, fig.width=8, fig.height=4}
# Zero probability curves
plot(fit_joint, type = "probability")
```

### Using Tidy Methods

```{r joint-tidy}
# Tidy coefficient table
tidy(fit_joint)

# Model summary statistics
glance(fit_joint)
```

### Key Takeaways

**When to use the joint model:**

- You have data across multiple conditions (alone, own, alt)
- You want correlated random effects across conditions
- You want a single unified analysis rather than separate fits

**When separate fits may be preferred:**

- You only have one condition
- You want maximum flexibility in model specification per condition
- Computational resources are limited

**Important notes:**

- `price_T` (target price) is the predictor for **all** streams
- Alternative price is assumed **fixed** and does not enter the model
- The joint model produces consistent parameter orderings with the separate fits

---

## Variant 2: Latent-Trait Joint Model

### Why a Latent-Trait Model?

The saturated joint model (Variant 1) uses **stream-specific random effects** — each
subject has separate deviations for alone-target, own-target, and own-alt streams.
This is flexible but can be unstable with sparse data.

The **latent-trait model** (Variant 2) takes a different approach:

- Instead of stream-specific random effects, subjects have **4 latent behavioral traits**
- These traits **load into** the model parameters via fixed mappings
- Result: fewer parameters, more interpretable structure, better stability

| Latent Trait | Behavioral Interpretation |
|--------------|---------------------------|
| `u_buy`      | Propensity to purchase (drives hurdle) |
| `u_val`      | Baseline consumption intensity |
| `u_sens`     | Price sensitivity (elasticity) |
| `u_sub`      | Substitution propensity |

**Key insight:** Variant 2 trades flexibility for structure, not correctness. Both
models use the exact same mean functions — only the random effect structure differs.

### Fitting the Latent-Trait Model

```{r fit-latent-model}
# Fit the latent-trait joint model
fit_latent <- fit_joint_hurdle(
    data = cp,
    joint_type = "latent",
    k = 2,
    verbose = 0
)

print(fit_latent)
```

### Comparing Variant 1 and Variant 2

Let's compare the two variants side-by-side:
```{r compare-variants}
# Hurdle parameters (should be similar)
cat("=== Hurdle Parameters ===\n")
cat("\nSaturated (Variant 1):\n")
print(coef(fit_joint, type = "hurdle"))
cat("\nLatent (Variant 2):\n")
print(coef(fit_latent, type = "hurdle"))

# Cross-price parameters
cat("\n=== Cross-Price Parameters ===\n")
cat("\nSaturated (Variant 1):\n")
print(coef(fit_joint, type = "crossprice"))
cat("\nLatent (Variant 2):\n")
print(coef(fit_latent, type = "crossprice"))
```

```{r compare-fit-stats}
# Model fit comparison
comparison <- data.frame(
    Variant = c("Saturated", "Latent"),
    LogLik = c(fit_joint$logLik, fit_latent$logLik),
    AIC = c(fit_joint$AIC, fit_latent$AIC),
    BIC = c(fit_joint$BIC, fit_latent$BIC),
    Converged = c(fit_joint$convergence == 0, fit_latent$convergence == 0)
)
comparison
```

### Latent Trait Loadings

The latent model includes **loading parameters** that control how the substitution
trait (`u_sub`) differentially affects each stream:

```{r latent-loadings}
# Extract latent loadings
coef(fit_latent, type = "latent")
```

- `lambda_sub_q0`: How substitution affects own-target baseline demand
- `lambda_sub_alpha`: How substitution affects own-target price sensitivity
- `lambda_sub_alt`: How substitution affects alternative consumption

### Predictions with the Latent Model

Predictions work identically to the saturated model:

```{r latent-predictions}
# Predicted demand across streams
prices <- data.frame(price_T = c(0.5, 1, 2, 4, 8, 16))

pred_response <- predict(fit_latent, newdata = prices, type = "response", stream = "all")
pred_response

# Zero probability predictions
pred_prob <- predict(fit_latent, newdata = prices, type = "probability", stream = "all")
pred_prob
```

### Interpreting Latent Traits

The latent traits provide a **behavioral decomposition** of individual differences:

| Trait | High Value Means | Low Value Means |
|-------|------------------|-----------------|
| `u_buy` | More likely to purchase at any price | More likely to abstain |
| `u_val` | Higher baseline consumption | Lower baseline consumption |
| `u_sens` | More price-sensitive (steeper curves) | Less price-sensitive |
| `u_sub` | Stronger substitution behavior | Weaker substitution |

**Important:** These are latent constructs, not directly observed. Their values
are estimated from the data pattern and should be interpreted with appropriate
caution.

### When to Use Each Variant

| Criterion | Saturated (Variant 1) | Latent (Variant 2) |
|-----------|----------------------|-------------------|
| **Data richness** | Many obs per subject | Sparse data |
| **Primary interest** | Stream-specific effects | Behavioral traits |
| **Stability** | May require more data | More stable |
| **Interpretability** | Random effects per stream | Explicit trait meanings |
| **Flexibility** | Maximum flexibility | Constrained structure |

**Recommendation:**

- Start with **Variant 1 (saturated)** if you have rich data and want maximum flexibility
- Use **Variant 2 (latent)** if:
 - Data is sparse
 - Model convergence is problematic
 - You want interpretable behavioral dimensions
 - Stability is more important than stream-specific flexibility

Both variants produce valid inference — the choice depends on your research goals
and data characteristics.
